<!-- To be compatible with workflowr, we need to specify the absolute directory under which the plots are generated, otherwise they will be saved into figure/<basename>.Rmd/<chunk-name>.<extension. and not visible in the html document: https://github.com/workflowr/workflowr/issues/111 -->
```{r template-options, include = FALSE}
library(ggbeeswarm)
knitr::opts_knit$set(
  base.dir = "docs"
  )
```


## Download results

<!-- Write DGE results to xlsx file -->
```{r write-dge-xlsx}
# File name summary
rmd_file <- current_input()
if(is.null(rmd_file))
  rmd_file <- 'tmp'
file_xlsx <- file.path('./docs/file',rmd_file, 'dge.xlsx')

dir.create(dirname(file_xlsx), recursive = TRUE, showWarnings = FALSE)

# Generate workbook
wb <- createWorkbook()
for(i in names(dge_list)) {
  addWorksheet(wb, i)
  res <- dge_list[[i]]$results %>% 
    dplyr::select(feature, gene_name, mean, detected, logFC, PValue, FDR)
  writeData(wb, i, res)
}
saveWorkbook(wb, file_xlsx, TRUE)

dge_file_xlsx <- file_xlsx
```

<!-- Write GSE results to xlsx file -->
```{r write-gse_list-xlsx}
gse_collections <- list(
  GSEA = names(gmt_files_symbols),
  gseGO = c("BP", "MF")
)

# File name summary
rmd_file <- current_input()
if(is.null(rmd_file))
  rmd_file <- 'tmp'

file_xlsx_list <- list()
for(sr in names(gse_collections)) {
  file_xlsx_list[[sr]] <- list()
  for(gc in gse_collections[[sr]]) {
    # Generate file
    file_xlsx <- file.path('./docs/file', 
                           rmd_file, 
                           paste0('gse_', tolower(sr), '_', tolower(gc), '.xlsx')
                           )
    file_xlsx_list[[sr]][[gc]] <- file_xlsx
    dir.create(dirname(file_xlsx), recursive = TRUE, showWarnings = FALSE)
    
    # Generate workbook
    wb <- createWorkbook()
    for(i in names(gse_list)) {
      addWorksheet(wb, i)
      if(!is.null(gse_list[[i]][[sr]][[gc]])) {
        res <- gse_list[[i]][[sr]][[gc]]@result
        writeData(wb, i, res)
      }
    }
    saveWorkbook(wb, file_xlsx, TRUE)
  }
}
```

<!-- Change dge_list names after writing to xlsx -->
```{r change-dge-names}
comp_names <- comp_list_description$comparison_long %>% 
  set_names(comp_list_description$comparison)
names(comp_list) <- comp_names[names(comp_list)]
names(dge_list) <- comp_names[names(dge_list)]
names(gse_list) <- comp_names[names(gse_list)]
```


The tables of results can be downloaded using the following links:

- [**Differential gene expression**](`r gsub("docs/", "" , dge_file_xlsx)`)
<!-- Write links to download GSE results -->
```{r write-links, echo = FALSE}
for(sr in names(file_xlsx_list)) {
  for(gc in names(file_xlsx_list[[sr]])) {
    file_xlsx <- gsub("docs/", "", file_xlsx_list[[sr]][[gc]])
    if(sr == 'GSEA') {
      cat(paste0("- [**Gene set enrichment analysis: ", gc, "**](", file_xlsx, ")"), '\n\n')
    }
    if(sr == 'gseGO') {
      cat(paste0("- [**Gene set enrichment analysis: GO:", gc, "**](", file_xlsx, ")"), '\n\n')
    }
  }
}
```

## Differential gene expression analysis

### Download results

- [**Differential gene expression**](`r gsub("docs/", "" , dge_file_xlsx)`)

### Summary of results

#### FDR distribution
```{r dge-summary-fdr-dist}
i <- names(dge_list[1])
res <- foreach(i = names(dge_list), .combine = rbind) %do% {
  x <- dge_list[[i]]$results
  c(
    `FDR < 0.1` = x %>% filter(FDR <= 0.1) %>% nrow,
    `FDR < 0.05` = x %>% filter(FDR <= 0.05) %>% nrow,
    `FDR < 0.01` = x %>% filter(FDR <= 0.01) %>% nrow,
    `FDR < 0.001` = x %>% filter(FDR <= 0.001) %>% nrow,
    `FDR < 0.0001` = x %>% filter(FDR <= 0.0001) %>% nrow
  )
}
if(class(res)[1] == 'integer'){
  res <- data.frame(res) %>% t() 
}

rownames(res) <- names(dge_list)
res %>% 
  data.frame(check.names = F) %>% 
  rownames_to_column('comparison') %>% 
  kbl(caption = 'Number of genes at different FDR thresholds') %>%
  kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  column_spec(1, bold = TRUE)
```

#### Number of genes at FDR 0.05
```{r dge-summary-fdr-05}
i <- names(dge_list[1])
res <- foreach(i = names(dge_list), .combine = rbind) %do% {
  x <- dge_list[[i]]$results %>% 
    filter(FDR <= 0.05)
  c(
    `Up` = x %>% filter(logFC > 0) %>% nrow,
    `Down` = x %>% filter(logFC < 0) %>% nrow
  )
}
if(class(res)[1] == 'integer'){
  res <- data.frame(res) %>% t() 
}

rownames(res) <- names(dge_list)
res %>% 
  data.frame(check.names = F) %>% 
  rownames_to_column('comparison') %>% 
  kbl(caption = 'Number of genes at at FDR < 0') %>%
  kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  column_spec(1, bold = TRUE)
```



### Volcano-plots
Top-20 genes based on adjusted P-values are labeled.
```{r dge-volcano-plot}
i=names(dge_list)[1]
p_thrs <- 0.05
for(i in names(dge_list)) {
  cat("#### Comparison:", i, "\n\n")
  # cat(dge_list[[i]]$comparison, "\n\n")
  x <- dge_list[[i]]$results
  res <- plot.volcano(
    x$logFC, 
    x$FDR, 
    label = x$gene_name, 
    fill.palette = c("gray80", "red"), 
    top.labels = 20, 
    pval.thr = p_thrs, 
    hline.color =  'grey60', hline.type = 2, hline.size = 0.5) +
    geom_vline(xintercept = 0, linetype = 2, linewidth = 0.5, color = 'grey60')
  print(res)
  cat("\n\n")
}
```


### Heatmaps
```{r dge-heatmap-plot, fig.width=7.2, fig.asp = 1.8}
i <- names(dge_list)[1]
p_thrs <- 0.05
lfc_thrs <- 1
topn <- 50
max_color_range <- 3 # put the color range to -3 to 3

for(i in names(dge_list)) {
  cat("#### Comparison", i, "\n\n")
  use_comp <- comp_list[[i]]
  
  # Extract samples from comparisons
  use_samples <- unlist(use_comp[1:2])
  use_se <- se[,use_samples]
  
  # Arrange samples by condition (first control and then sample)
  use_annot <- colData(use_se) %>% data.frame %>% 
    mutate(
      condition_new = ifelse(sample_alias %in% use_comp[[1]], 
                             names(use_comp)[1], 
                             names(use_comp)[2]),
      condition_new = factor(condition_new, levels = names(use_comp)[2:1]),
      condition_num = as.numeric(condition_new)
    ) %>% 
    arrange(condition_num, donor)
  use_samples <- use_annot$sample_alias
  
  # Extract DGE results
  use_dge <- dge_list[[i]]$results %>% 
    filter(FDR < p_thrs & abs(logFC) > lfc_thrs) %>% 
    mutate(direction = ifelse(logFC > 0, 'Up', 'Down')) %>% 
    arrange(FDR) %>% 
    head(topn) %>% 
    arrange(logFC)
  if(nrow(use_dge) > 50){
    cat(paste0("Showing ", nrow(use_dge)," features with FDR < ", p_thrs, " and absolute log2 fold-change > ", lfc_thrs, "\n"))
  } else {
    cat(paste0("Only ", nrow(use_dge)," features with FDR < ", p_thrs, " and absolute log2 fold-change > ", lfc_thrs, ". Showing top-", topn," features based on P value and absolute log2 fold-change >", lfc_thrs,"\n"))
    use_dge <- dge_list[[i]]$results %>% 
      filter(abs(logFC) > lfc_thrs) %>% 
      arrange(PValue) %>% 
      mutate(direction = ifelse(logFC > 0, 'Up', 'Down')) %>% 
      head(topn) %>% 
      arrange(logFC)
  }
  
  # Extract matrix of expression
  mat <- assay(use_se, 'logcounts')[rownames(use_dge), use_samples]
  rownames(mat) <- make.names(use_dge$gene_name, unique = TRUE)
  zmat <- t(apply(mat, 1, scale, center = TRUE, scale = TRUE))
  colnames(zmat) <- colnames(mat)
  zmat <- zmat[,rownames(use_annot)]
  
  # Col Annotation
  use_annot_levels <-levels(use_annot$condition_new)
  column_ha <-  HeatmapAnnotation(
    condition = anno_block(
      # gp = gpar(fill = c('#fdbf6f', '#6a3d9a') %>% set_names(use_annot_levels)),
      labels = use_annot_levels, 
      labels_gp = gpar(col = "black", fontsize = 8)
    ),
    `Sample type` = use_annot$sample_type_grouped_3l,
    `Mouse` = use_annot$mouse_id,
    col = list(`Sample type` = sample_type_barcode_palette)
  )
    
  # Row annotation (right)
  row_annot_right <-  rowAnnotation(
    log2FC = anno_numeric(
      round(use_dge$logFC,1), 
      labels_gp = gpar(fontsize = 6), 
      bg_gp = gpar(
        fill = c("#74add1", "#f46d43"), 
        col = 'white'
      ),
      bar_width = unit(1, "npc") - unit(1.5, "pt")
    ), 
    annotation_name_rot = 0)
  
  # Row annotation (left)
  direction_palette <- c('Up' = "#f46d43", 'Down' = "#74add1")
  direction_palette <- direction_palette[names(direction_palette) %in% use_dge$direction]
  
  row_annot_left <-  rowAnnotation(
    direction = anno_block(
      gp = gpar(fill = direction_palette),
      labels = names(direction_palette), 
      labels_gp = gpar(col = "white", fontsize = 8)
    )
  )
  
  # Heatmap color
  zmax <- zmat %>% abs %>% max
  if(!is.null(max_color_range))
    zmax <- max_color_range
  col_fun <-  colorRamp2(
    seq(-zmax, zmax, length.out = 11),
    rev(brewer.pal(11, "RdBu"))
  )
  
  # Generate Heatmap
  htz <- Heatmap(
    zmat,
    name= 'z-score',
    col = col_fun,
    row_names_gp = gpar(fontsize = 8),
    column_names_gp = gpar(fontsize = 8),
    
    top_annotation = column_ha,
    column_split = use_annot$condition_num,
    #cluster_column_slices = turn off the clustering on slices, and now you can precisely control the order of slices
    cluster_column_slices = FALSE, 
    column_title = NULL,
    
    right_annotation = row_annot_right,
    row_split = use_dge$direction,
    # left_annotation = row_annot_left,
    row_title = NULL
  )
  
  draw(
    htz, 
    # heatmap_legend_side = "bottom", 
    annotation_legend_side = "bottom"
  )
    
  cat("\n\n")
  
}
```



## Functional enrichment analysis
```{r set-gsea-collections}
gse_collections <- list(
  GSEA = names(gmt_files_symbols),
  gseGO = c("BP", "MF")
)

```

### Download results
<!-- Write links to download GSE results -->
```{r write-links-gse-2, echo = FALSE}
for(sr in names(file_xlsx_list)) {
  for(gc in names(file_xlsx_list[[sr]])) {
    file_xlsx <- gsub("docs/", "", file_xlsx_list[[sr]][[gc]])
    if(sr == 'GSEA') {
      cat(paste0("- [**Gene set enrichment analysis: ", gc, "**](", file_xlsx, ")"), '\n\n')
    }
    if(sr == 'gseGO') {
      cat(paste0("- [**Gene set enrichment analysis: GO:", gc, "**](", file_xlsx, ")"), '\n\n')
    }
  }
}
```

### Summary of results
```{r gse-summary-fdr}
res <- foreach(i = names(gse_list), .combine = rbind) %do% {
  foreach(sr = names(gse_collections), .combine = rbind) %do% {
    foreach(gc = gse_collections[[sr]], .combine = rbind) %do% {
      if(sr == 'GSEA') {
        x <- gse_list[[i]][[sr]][[gc]]@result %>% 
          filter(p.adjust <= 0.05)
        if(gc == 'msigdb.c2.cp') {
          x %<>% 
            filter(grepl('REACTOME|KEGG|WP', ID))
        }
      }
      if(sr == 'gseGO') {
        x <- gse_list[[i]][[sr]][[gc]]@result %>% 
          filter(p.adjust <= 0.05)
      }
      c(
        Comparison = i,
        `Collection` = gc,
        `Total` = nrow(x),
        `Supressed` = x %>% filter(NES < 0) %>% nrow,
        `Activated` = x %>% filter(NES > 0) %>% nrow
      )
    }
  }
}
res <- res %>% data.frame(check.names = F)
```

#### Table
```{r gse-summary-fdr-table}
res %>% 
  data.frame(check.names = F) %>% 
  kbl(caption = 'Number of gene sets at FDR < 0.05',
              row.names = F) %>%
  kable_paper(bootstrap_options = c("striped", "hover", "condensed"), 
              full_width = F) %>%
  column_spec(1, bold = TRUE)
```

#### Barplot
```{r gse-summary-fdr-barplot, fig.asp = 1.3}
res %>% 
  pivot_longer(cols = c('Total', 'Supressed', 'Activated'), names_to = 'Direction', values_to = 'Count') %>% 
  mutate(
    comb_names = paste(Comparison, Collection),
    Count = as.numeric(Count),
    Comparison = gsub("-over-", " over\n", Comparison),
    Comparison = gsub(" corrected", "\ncorrected", Comparison)
    ) %>% 
  ggplot(aes(Collection, Count, fill = Direction)) +
  geom_col(position = "dodge") +
  coord_flip() +
  facet_grid(rows = vars(Comparison)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_fill_manual(values = c(Supressed = '#74add1', 
                               Activated = '#f46d43',
                               Total = 'grey80')
                    ) +
  labs(x = '', fill = '') +
  theme_facet +
  theme(
    axis.line = element_blank(),
    axis.ticks = element_line(linewidth = one_pt/4, color = 'black'),
    strip.text = element_text(size = 5)
    
  )
```

### Dotplots
<!-- # Subchunkify dotplots -->
<!-- ## chunk options need to be set to echo=FALSE, results='asis' -->
<!-- ## Cause results='asis',we must use raw HTML to print text and headings -->
```{r gse-dotplot, echo=FALSE, results='asis'}
knitr::opts_knit$set(
  base.dir = "docs"
  )

use_linewidth <- 1/2.141959
root_section_number <- 8.2
section_n <- 1
subsection_n <- 1

for(i in names(gse_list)) {
  h_section_number <- paste(root_section_number, section_n, sep=".")
  cat("<h4> <span class=\"header-section-number\">",h_section_number,"</span>Comparison", i, "</h4>\n\n") #Caue results='asis',we must use raw HTML
  section_n <- section_n + 1
  for(sr in names(gse_collections)) {
    for(gc in gse_collections[[sr]]) {
      if(sr == 'GSEA') {
        x <- gse_list[[i]][[sr]][[gc]]@result %>% 
          filter(p.adjust <= 0.05) %>% 
          mutate(Description = clean_msigdb_names(Description))
        if(gc == 'msigdb.c2.cp') {
          x %<>% 
            filter(grepl('REACTOME|KEGG|WP', ID))
        }
      }
      if(sr == 'gseGO') {
        x <- gse_list[[i]][[sr]][[gc]]@result %>% 
          filter(p.adjust <= 0.05 & simplify)
      }
      if(nrow(x) > 0) {
        h_subsection_number <- paste(h_section_number, subsection_n, sep=".")
        cat("<h5> <span class=\"header-section-number\">",h_subsection_number,"</span>Collection", gc, "</h5>\n\n") #Caue results='asis',we must use raw HTML
        subsection_n <- subsection_n + 1
        
        x %<>% 
          mutate(log10P = -log10(p.adjust)) %>% 
          arrange(p.adjust) %>% 
          head(30)
        
        res <- gse.dotplot(
          x,
          x = 'NES',
          color = 'NES',
          size = "log10P",
          # color2log10 = TRUE,
          showCategory = 20,
          label_format = 60
        ) +
          geom_vline(xintercept = 0, lty = 3) +
          scale_size_continuous(range=c(0.3, 2.5)) +
          scale_color_gradient2(trans = "reverse") +
          scale_x_continuous(expand = expansion(mult = c(0.05, 0.05))) +
          labs(
            size = expression(paste("lo", g[10],"(Adjusted P value)")),
            caption = 'Showing only top-30 gene sets\nNES = Normalized Enrichment Score'
            # size = 'Gene set\nsize'
          ) +
          theme(
            panel.border = element_rect(color = "black", fill = NA, size = use_linewidth/2),
            
            axis.line = element_blank(),
            axis.ticks = element_line(color = "black", size = use_linewidth/2),
            axis.text.y = element_text(size = 4),
            axis.text.x = element_text(size = 4),
            
            legend.text = element_text(size = 4), 
            legend.title = element_text(size = 4),
            legend.key.width = unit(0.2, "cm")
          )
        
        subchunkify(res, 
                    fig.width = 3.5, 
                    fig.height = (3.5*0.3) + nrow(x)*0.05,
                    dpi = 600,
                    prefix = 'gse-dotplot',
                    suffix = paste0(tolower(i) %>% gsub(" ", "-", .), '_', sr, '_', gc) )
        cat("\n\n")
      }
    }
  }
}

```



### Gene-set similarity matrix

```{r gse-jcmatrix-by-gset, echo=FALSE}
#
# Generate a list of JC similarity matrices by gene-set
#
library(enrichplot)
p_thrs <- 0.05
topn <- Inf
use_comp <- names(gse_list)[1]
jc_matrix_list <- foreach(use_comp = names(gse_list)) %do% {
  data_gse <- foreach(sr = names(gse_collections)) %do% {
    data_by_collection <- foreach(gc = gse_collections[[sr]])  %do% {
      
      if(sr == 'GSEA') {
        use_enrich_results <- gse_list[[use_comp]][[sr]][[gc]]
        top_descriptions <- use_enrich_results@result %>% 
          filter(p.adjust <= p_thrs) %>% 
          head(topn) %>% 
          pull(Description)
        use_n_categories <-  use_enrich_results@result %>% 
          filter(p.adjust <= p_thrs) %>% 
          nrow
      }
      if(sr == 'gseGO') {
        use_enrich_results <- gse_list[[use_comp]][[sr]][[gc]]
        top_descriptions <- use_enrich_results@result %>% 
          filter(p.adjust <= p_thrs & simplify) %>% 
          head(topn) %>% 
          pull(Description)
        use_n_categories <-  use_enrich_results@result %>% 
          filter(p.adjust <= p_thrs) %>% 
          nrow
      }
      if(sr %in% c('gseKEGG', 'gseReactome')) {
        use_enrich_results <- gse_list[[use_comp]][[sr]]
        top_descriptions <- use_enrich_results@result %>% 
          filter(p.adjust <= p_thrs) %>% 
          head(topn) %>% 
          pull(Description)
        use_n_categories <-  use_enrich_results@result %>% 
          filter(p.adjust <= p_thrs) %>% 
          nrow
      }
      
      if(length(top_descriptions) > 1) {
        use_enrich_results <- pairwise_termsim(
          use_enrich_results, 
          showCategory = use_n_categories, 
          method = "JC")
        use_enrich_results@termsim[top_descriptions, top_descriptions]
      } else {
        NULL
      }
    }
    names(data_by_collection) <- gse_collections[[sr]]
    return(compact(data_by_collection))
  }
  names(data_gse) <- names(gse_collections)
  return(data_gse)
}
names(jc_matrix_list) <- names(gse_list)
```


Gene-set similarity matrix by comparison and gene set collections. Heatmap showing the pair-wise similarity matrix of top `r topn` gene sets with an adjusted P value < `r p_thrs` in each gene group. Heatmap colors represent the Jaccard similarity coefficient.

```{r gse-jcmatrix-by-gset-heatmap, echo=FALSE, results='asis'}
knitr::opts_knit$set(
  base.dir = "docs"
  )

library(NbClust)

root_section_number <- 6.5
section_n <- 1
subsection_n <- 1

col_fun <- colorRamp2(seq(0, 1, length.out = 4), brewer.pal(4, "GnBu"))

i <- names(jc_matrix_list)[1]
for(i in names(jc_matrix_list)) {
  h_section_number <- paste(root_section_number, section_n, sep=".")
  
  cat("<h4> <span class=\"header-section-number\">",h_section_number,"</span>Comparison", i, "</h4>\n\n") #Caue results='asis',we must use raw HTML
  
  section_n <- section_n + 1
  
  for(sr in names(jc_matrix_list[[i]])) {
    
    for(gc in names(jc_matrix_list[[i]][[sr]])) {
      use_mat <- jc_matrix_list[[i]][[sr]][[gc]]
      if(nrow(use_mat) >= 10) {
        h_subsection_number <- paste(h_section_number, subsection_n, sep=".")
        cat("<h5> <span class=\"header-section-number\">",h_subsection_number,"</span>Collection", gc, "</h5>\n\n") #Caue results='asis',we must use raw HTML
        subsection_n <- subsection_n + 1
        colnames(use_mat) <- colnames(use_mat) %>% clean_msigdb_names %>% tolower() %>% str_to_title %>% ep_str_wrap(70)
        rownames(use_mat) <- rownames(use_mat) %>% clean_msigdb_names %>% tolower() %>% str_to_title %>% ep_str_wrap(70)
        
        # Define number of clusters
        use_max.nc <- ifelse(nrow(use_mat) <= 15, 5, round(nrow(use_mat) * 0.3))
        nbclust_res <- NbClust(data = NULL, diss = as.dist(1-use_mat), distance = NULL, 
                               min.nc = 2, max.nc = use_max.nc, method = "ward.D",
                               index = 'silhouette')
        
        
        n_split <- nbclust_res$Best.nc['Number_clusters']
        
        # ha_top <- HeatmapAnnotation(
        #   foo = anno_block(
        #     # labels = c("Translation", "Cell division"),
        #     labels = 1:n_split,
        #     labels_gp = gpar(col = "black", fontsize = 8),
        #     gp = gpar(lwd = 0, lty = 0))
        # )
        # use_border <- ifelse(nrow(use_mat) > 30, FALSE, TRUE)
        use_gap_size <- ifelse(nrow(use_mat) > 30, 0.2, 1)
        use_border_lwd <- ifelse(nrow(use_mat) > 30, one_pt,one_pt/5)

        ht <- Heatmap(
          use_mat,
          name = 'Jaccard index',
          column_split = n_split,
          row_split = n_split,
          column_title = NULL,
          row_title = NULL,
          row_names_gp = gpar(fontsize = 4),
          col = col_fun,
          show_column_dend = FALSE,
          show_column_names = FALSE,
          border = TRUE,
          border_gp = gpar(col = "black", lwd = use_border_lwd),
          row_gap = unit(use_gap_size, "mm"),
          column_gap = unit(use_gap_size, "mm"),
          # top_annotation = ha_top,
          heatmap_legend_param = list(title_gp = gpar(fontsize = 8),
                                      labels_gp = gpar(fontsize = 8)
          ),
          width = unit(7, "cm"),
          
          clustering_method_rows = "ward.D",
          clustering_method_columns = "ward.D"
          # clustering_distance_rows = function(x) as.dist(1 - x),
          # clustering_distance_columns = function(x) as.dist(1 - x)
        )
        use_width <-  7.2 + nrow(use_mat) * 0.06
        use_asp <-  0.618 + nrow(use_mat) * 0.05
        use_height <- use_width * use_asp
        subchunkify(draw(ht),
                    fig.width = use_width,
                    fig.height = use_height,
                    dpi = 600,
                    prefix = 'gse-jcmatrix-by-gset-heatmap',
                    suffix = paste0(tolower(i) %>% gsub(" ", "-", .), '_', sr, '_', gc) )
        cat("\n\n")
      }
    }
  }
}
```






