---
title: "Cell type analysis using deconvolution"
subtitle: "Integration of xenografts, syngeneic and patient results"
author: "Francesc Castro-Giner"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
params:
  date: '`r format(Sys.Date(), "%B %d, %Y")`'
  deconv_dir_regex: ./output/rnaseq/MODEL/deconvolution
  se_path_mm: ./data/rnaseq/syngeneic/se_qc.rds,./data/rnaseq/xenografts/se_qc_mm.rds
  se_path_hs: ./data/rnaseq/patients_zh_h/se_qc.rds,./data/rnaseq/patients_tx/se.rds
  model: xenografts,syngeneic,patients_zh_h,patients_tx
---
<!-- se_path: ./output/rnaseq/syngeneic/se_qc.rds,./output/rnaseq/xenografts/se_qc_mm.rds,./output/rnaseq/patients/se_qc.rds,./output/rnaseq/p28634_o31042_o32050_Tx/summarizedExperiment/se.rds -->

## Load libraries, additional functions and data

Setup environment
```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(results='asis', echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.align = 'center', fig.width = 3.5, fig.asp = 0.618, dpi = 600, dev = c("png", "pdf"), fig.showtext = FALSE, engine.opts = list(bash = "-l"))

options(stringsAsFactors = FALSE)

use_seed <- 1100101
set.seed(use_seed)

# dir.create(params$output_dir, recursive = TRUE, showWarnings = FALSE)
```

Load packages
```{r load-libraries}
library(tidyverse)
library(foreach)
library(knitr)
library(ggbeeswarm)
library(magrittr)
library(SummarizedExperiment)
library(edgeR)
library(scater)
library(xCell2)
library(immunedeconv)
```


Load ggplot theme
```{r ggplot-theme}
source("./configuration/rmarkdown/ggplot_theme.R")
```

Set color palettes
```{r, echo = FALSE}
source("./configuration/rmarkdown/color_palettes.R")
```

Clean files generated in previous runs
```{r clean-files}
rmd_file <- current_input()
if(!is.null(rmd_file)) {
  figures_dir <- file.path('./docs/figure',rmd_file)
  if(dir.exists(figures_dir)) {
    unlink(file.path(figures_dir, "*"))
  }
}
```



## Data wrangling

Process params
```{r process-params}
list_models <- str_split(params$model, ',', simplify = FALSE) %>% unlist
se_mm_list <-  str_split(params$se_path_mm, ',', simplify = FALSE) %>% unlist
se_hs_list <-  str_split(params$se_path_hs, ',', simplify = FALSE) %>% unlist
se_list <- c(se_mm_list, se_hs_list)
```

Run cell type deconvolution using xcell2 TMECompendium (Xenografts and Syngeneic)
```{r xcell2-deconvolution-mm, eval = FALSE}
se_file <- se_mm_list[1]
for(se_file in se_mm_list) {
  model <- dirname(se_file) %>% basename()
  out_dir <- gsub('MODEL', model, params$deconv_dir_regex)
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  
  se <- readRDS(se_file)
  
  # Normalization
  norm_method <- "TMM"
  exprs_mat_for_norm <- assay(se, 'counts')
  norm_factors <- calcNormFactors(exprs_mat_for_norm, method = norm_method)
  lib_size <- colSums(exprs_mat_for_norm)/10^6
  se$size_factor <- (norm_factors * lib_size)
  se <- logNormCounts(se, size.factors =se$size_factor)
  se <- logNormCounts(se, size.factors =se$size_factor, log = FALSE)

  
  mat <- assay(se, 'normcounts')
  rownames(mat) <- make.names(rowData(se)$gene_name, unique = TRUE)
  human_mat <- immunedeconv::convert_human_mouse_genes(mat, convert_to = 'human')
  
  # TME Compendium
  data(TMECompendium.xCell2Ref, package = "xCell2")
  xCell2ref <- TMECompendium.xCell2Ref
  genes_ref <- getGenesUsed(xCell2ref)
  genes_mix <- rownames(human_mat)
  overlap_percentage <- round(length(intersect(genes_mix, genes_ref)) / length(genes_ref) * 100, 2)
  print(paste0("Overlap between mixture and reference genes is: ", overlap_percentage, "%"))
  xcell2_results <- xCell2Analysis(
    mix = human_mat,
    xcell2object = xCell2ref,
    minSharedGenes = 0.75 # Allow for a lower overlap threshold
  )
  saveRDS(xcell2_results, file.path(out_dir, 'xcell2_TMECompendium_a75.rds'))
  rm(TMECompendium.xCell2Ref)

}
```

Run cell type deconvolution using xcell2 TMECompendium (Patients)
```{r xcell2-deconvolution-hs, eval = FALSE}
se_file <- se_hs_list[1]
for(se_file in se_hs_list) {
  model <- dirname(se_file) %>% basename()
  out_dir <- gsub('MODEL', model, params$deconv_dir_regex)
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

  se <- readRDS(se_file)
  
  # Normalization
  norm_method <- "TMM"
  exprs_mat_for_norm <- assay(se, 'counts')
  norm_factors <- calcNormFactors(exprs_mat_for_norm, method = norm_method)
  lib_size <- colSums(exprs_mat_for_norm)/10^6
  se$size_factor <- (norm_factors * lib_size)
  se <- logNormCounts(se, size.factors =se$size_factor)
  se <- logNormCounts(se, size.factors =se$size_factor, log = FALSE)

  
  mat <- assay(se, 'normcounts')
  rownames(mat) <- make.names(rowData(se)$gene_name, unique = TRUE)
  mat <- mat[,colnames(mat) != 'Luz_PDAC_007_10']
  
  # TME Compendium
  data(TMECompendium.xCell2Ref, package = "xCell2")
  xCell2ref <- TMECompendium.xCell2Ref
  genes_ref <- getGenesUsed(xCell2ref)
  genes_mix <- rownames(mat)
  overlap_percentage <- round(length(intersect(genes_mix, genes_ref)) / length(genes_ref) * 100, 2)
  print(paste0("Overlap between mixture and reference genes is: ", overlap_percentage, "%"))
  xcell2_results <- xCell2Analysis(
    mix = mat,
    xcell2object = xCell2ref,
    minSharedGenes = 0.75 # Allow for a lower overlap threshold
  )
  saveRDS(xcell2_results, file.path(out_dir, 'xcell2_TMECompendium_a75.rds'))
  rm(TMECompendium.xCell2Ref)

}
```

Load sample annotation from SummarizedExperiment objects for each model
```{r load-se-objects}
model <- list_models[1]
sample_annot <- foreach(model = list_models, .combine = rbind) %do% {
  se_file <- grep(model, se_list, value = TRUE)
  use_se <- readRDS(se_file)
  res <- use_se %>% colData %>% data.frame 
  if(! 'anatomical_origin' %in% colnames(res)) {
    res$anatomical_origin <- 'blood'
  }
  
  res %>% 
    select(donor, sample_alias, anatomical_origin, origin:ctc_cluster_id)
}


sample_annot %<>% 
  mutate(
    # Re origin names
    origin = ifelse(origin == 'patient_autopsy', 'patient', origin),
    # Change sample type
    sample_type = ifelse(sample_type == 'ctc_cluster_wbc', 'ctc_cluster_heterotypic', sample_type),
    sample_type = ifelse(sample_type == 'ctc_cluster_heterotypic_dn', 'ctc_cluster_heterotypic', sample_type),
    sample_type = ifelse(sample_type == 'ctc_cluster_dn', 'ctc_cluster', sample_type),
    sample_type = ifelse(sample_type == 'cto_wbc', 'cto_heterotypic', sample_type)
  )
# Add tumor purity
sample_annot %<>% 
  mutate(
    tumor_purity = ctc_n / (ctc_n + ctc_n_wbc_attached),
    tumor_purity = ifelse(origin == 'xenograft', 0, tumor_purity)
  )
```

Load deconvolution files and process score, normalizing by tumor purity in syngeneic models and patients.
```{r load-xcell2-files}
references <- c('TMECompendium')
list_models <- str_split(params$model, ',', simplify = FALSE) %>% unlist

ref <- references[1]
model <- list_models[1]
xcell2_res <- foreach(ref = references) %do% {
  deconv_res <- foreach(model = list_models, .combine = rbind) %do% {
    use_file <- file.path(params$deconv_dir_regex, 
                          paste0('xcell2_',ref,'_a75.rds')) %>% 
      gsub('MODEL', model, .)
    if(file.exists(use_file)) {
      res <- readRDS(use_file)
      res %>% t() %>% 
        data.frame(check.names = FALSE) %>% 
        rownames_to_column('sample_alias') %>% 
        pivot_longer(-sample_alias, names_to = 'deconv_cell_type', values_to = 'score') %>% 
        mutate(score = as.numeric(score)) %>% 
        left_join(sample_annot, by = 'sample_alias')
    } else {
      NULL
    }
  }
  
  
  deconv_res %<>% 
    # Remove cell tyes not of interest
    filter(!sample_type %in% c('dn', 'dn_cell', 'dn_cluster_heterotypic')) %>% 
    mutate(
      # Normalize scores
      score_norm = score / (1-tumor_purity),
      score_norm = ifelse(score == 0, 0, score_norm),
      # In cases with Purity == 1, we assign the raw score
      score_norm = ifelse(tumor_purity == 1, score, score_norm),
      # Factor sample types
      sample_type = factor(
        sample_type,
        levels = c(
          'ctc_single',
          'ctc_single_pool',
          'ctc_cluster',
          'ctc_cluster_wbc',
          'ctc_cluster_heterotypic',
          'ctc_cluster_heterotypic_dn',
          'cto',
          'cto_wbc',
          'cto_heterotypic',
          'ctc_cluster_dn',
          'ctc_gfp'
        )
      )
    )
  
  # Set up colors for major cell types
  major_cell_types <- deconv_res %>% 
    # filter(score > 0.01) %>% 
    pull(deconv_cell_type) %>% unique %>% sort
  
  res_palette <- get_scater_palette('tableau20')[1:length(major_cell_types)] %>% 
    set_names(major_cell_types)
  
  
  return(
    list(      
      res = deconv_res,
      palette = res_palette
    )
  )
}

names(xcell2_res) <- references
```


## TMECompendium
```{r xcell2-TMECompendium-conf}
reference <- 'TMECompendium'
use_res <- xcell2_res[[reference]]$res

use_samples <- use_res %>% 
  pull(sample_alias) %>% unique

use_deconv_cell_type <- use_res %>% 
  filter(score > 0.01) %>%
  pull(deconv_cell_type) %>% unique

# Sort levels
use_palette <- c(
  `macrophage` = '#7cd0f5',
  `alternatively activated macrophage` = "#17BECF",
  `inflammatory macrophage` = "#9EDAE5",
  `monocyte` = '#1F77B4',
  `myeloid cell` = "#AEC7E8",
  `neutrophil` = "#fddc0d",
  `T cell` = "#2CA02C",
  `CD8-positive, alpha-beta T cell` = "#98DF8A",
  `gamma-delta T cell` = "#1ec397",
  `Non plasma B cell` = "#9467BD",
  # `B cell` = "#C5B0D5",
  `B cell` = "#FF7F0E",
  `natural killer cell` = "#D62728",
  # `fibroblast` = "#FF7F0E",
  `fibroblast` = "#C5B0D5",
  `malignant cell` = 'grey70' ,
  `other` = 'lightgrey' 
)


# Calculate the score for mieloid cell type
myeloid_score <- use_res %>% 
  filter(deconv_cell_type %in% c('macrophage', 'alternatively activated macrophage', 'inflammatory macrophage', 'monocyte', 'myeloid cell')) %>%
  group_by(sample_alias) %>% 
  summarise(
    myeloid_score = sum(score)
  ) %>% ungroup

use_res <- use_res %>% 
  filter(sample_alias %in% use_samples) %>% 
  left_join(myeloid_score) %>% 
  mutate(
    deconv_cell_type = ifelse(deconv_cell_type %in% use_deconv_cell_type, deconv_cell_type, 'other'),
    deconv_cell_type = factor(deconv_cell_type, levels = rev(names(use_palette)))
    )


# Palette

# library(scales)
# use_palette <- c('#fddc0d', '#7cd0f5', get_scater_palette('tableau20'))
# show_col(use_palette)

# use_palette <- get_scater_palette('tableau20')[1:length(use_deconv_cell_type)] %>% 
#   set_names(use_deconv_cell_type)
# use_palette <- c(use_palette, other = 'lightgrey')

# We remove cell type malignant cells and NA
use_res %<>% 
  filter(deconv_cell_type != 'malignant cell') %>% 
  filter(!is.na(deconv_cell_type))
```

### Syngeneic
```{r xcell2-TMECompendium-allograft-conf, fig.width=5.2, fig.asp = 0.9}
model <- 'allograft'
```

#### Barplot 
```{r xcell2-TMECompendium-allograft-barplot, fig.width=5.2, fig.asp = 0.5}
res_plot <- use_res %>% 
  filter(origin == model) %>% 
  filter(sample_type %in% c('ctc_cluster_heterotypic', 'cto_heterotypic')) %>% 
  mutate(sample_alias = fct_reorder(sample_alias, myeloid_score)) %>% 
  ggplot(aes(x = sample_alias, y = as.numeric(score), fill = deconv_cell_type)) +
  geom_col() +
  # geom_col(position = 'fill') +
  # geom_bar(stat = 'identity', position = 'fill', color = 'black', size = 0.1) +
  scale_fill_manual(values = use_palette) +
  # facet_grid(sample_type~donor, scales = 'free_x', space = 'free') +
  facet_grid(cols = vars(sample_type, donor), scales = 'free_x', space = 'free') +
  # facet_wrap(vars(donor), scales = 'free_x') +
  labs(
    x = '',
    y = 'Enrichment score',
    fill = 'Cell type',
    # title = 'Cell type composition',
    caption = 'The score generated by xCell2 is not a percentage, it can be used to compare between samples'
  )  +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    # axis.text.x = element_text(angle = 90, hjust = 1),
    strip.text.x = element_text(angle = 90, hjust = 1)
  ) +
  theme_facet

print(res_plot)
cat("\n\n")
```


### Xenograft
```{r xcell2-TMECompendium-xenograft-conf, fig.width=5.2, fig.asp = 0.9}
model <- 'xenograft'
```

#### Barplot 
```{r xcell2-TMECompendium-xenograft-barplot, fig.width=5.2, fig.asp = 0.5}
res_plot <- use_res %>% 
  filter(origin == model) %>% 
  filter(sample_type %in% c('ctc_cluster_heterotypic', 'cto_heterotypic')) %>% 
  mutate(sample_alias = fct_reorder(sample_alias, myeloid_score)) %>% 
  ggplot(aes(x = sample_alias, y = as.numeric(score), fill = deconv_cell_type)) +
  geom_col() +
  # geom_col(position = 'fill') +
  # geom_bar(stat = 'identity', position = 'fill', color = 'black', size = 0.1) +
  scale_fill_manual(values = use_palette) +
  # facet_grid(sample_type~donor, scales = 'free_x', space = 'free') +
  facet_grid(cols = vars(sample_type, donor), scales = 'free_x', space = 'free') +
  # facet_wrap(vars(donor), scales = 'free_x') +
  labs(
    x = '',
    y = 'Enrichment score',
    fill = 'Cell type',
    # title = 'Cell type composition',
    caption = 'The score generated by xCell2 is not a percentage, it can be used to compare between samples'
  )  +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    # axis.text.x = element_text(angle = 90, hjust = 1),
    strip.text.x = element_text(angle = 90, hjust = 1)
  ) +
  theme_facet

print(res_plot)
cat("\n\n")
```

### Patient
```{r xcell2-TMECompendium-patient-conf, fig.width=5.2, fig.asp = 0.9}
model <- 'patient'
```

#### Barplot 
```{r xcell2-TMECompendium-patient-barplot, fig.width=5.2, fig.asp = 0.5}
res_plot <- use_res %>% 
  filter(origin == model) %>% 
  mutate(sample_alias = fct_reorder(sample_alias, myeloid_score)) %>% 
  ggplot(aes(x = sample_alias, y = as.numeric(score), fill = deconv_cell_type)) +
  geom_col() +
  # geom_col(position = 'fill') +
  # geom_bar(stat = 'identity', position = 'fill', color = 'black', size = 0.1) +
  scale_fill_manual(values = use_palette) +
  # facet_grid(sample_type~donor, scales = 'free_x', space = 'free') +
  facet_grid(cols = vars(sample_type, donor), scales = 'free_x', space = 'free') +
  # facet_wrap(vars(donor), scales = 'free_x') +
  labs(
    x = '',
    y = 'Enrichment score',
    fill = 'Cell type',
    # title = 'Cell type composition',
    caption = 'The score generated by xCell2 is not a percentage, it can be used to compare between samples'
  )  +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    # axis.text.x = element_text(angle = 90, hjust = 1),
    strip.text.x = element_text(angle = 90, hjust = 1)
  ) +
  theme_facet

print(res_plot)
cat("\n\n")
```



