---
title: "Analysis of PDAC CTC patient samples"
subtitle: "Differential gene expression between CTC types accounting for patient differences"
author: "Francesc Castro-Giner"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
params:
  date: '`r format(Sys.Date(), "%B %d, %Y")`'
  se_path: ./data/rnaseq/patients-ctc/se_qc.rds
  output_dir: ./output/rnaseq/patients-ctc/differential_expression_edger/donor_batch/sample_type
  min_counts: 5
  min_present_prop: 0.40
  msigdb_dir: ./data/resources/MSigDB
  msigdb_vs: v2023.1.Hs
---

## Load libraries, additional functions and data

Setup environment
```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(results='asis', echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.align = 'center', fig.width = 3.5, fig.asp = 0.618, dpi = 600, dev = c("png", "pdf"), engine.opts = list(bash = "-l"))

options(stringsAsFactors = FALSE)

use_seed <- 1100101
set.seed(use_seed)

dir.create(params$output_dir, recursive = TRUE, showWarnings = FALSE)
```

Load packages
```{r load-libraries}
library(tidyverse)
library(showtext)
library(foreach)
library(DT)
library(knitr)
library(kableExtra)
library(cowplot)
library(colorblindr)
library(RColorBrewer)
library(ggbeeswarm)
library(ggrepel)
library(magrittr)
library(openxlsx)

library(SummarizedExperiment)
library(edgeR)
library(DESeq2)
library(scuttle)
library(scran)
library(pheatmap)
library(PCAtools)
library(org.Hs.eg.db)
library(ComplexHeatmap)
library(circlize)
library(enrichplot)
library(NbClust)
```

Load ggplot theme
```{r ggplot-theme}
source("./configuration/rmarkdown/ggplot_theme.R")
```

Set color palettes
```{r, echo = FALSE}
source("./configuration/rmarkdown/color_palettes.R")
```

Load custom functions
```{r load-functions}
source('./code/R-functions/dge_wrappers.r')
source('./code/R-functions/dge_report.r')
source('./code/R-functions/gse_omnibus.r')
source('./code/R-functions/gse_report.r')
source('./code/R-functions/subchunkify.R')
```

Load MSigDB gene sets
```{r load-msigdb}
gmt_files_symbols <- list(
  msigdb.c2.cp = paste0(params$msigdb_dir, '/', params$msigdb_vs, '/c2.cp.', params$msigdb_vs, '.symbols.gmt')
)
```

Clean files generated in previous runs
```{r clean-files}
rmd_file <- current_input()
if(!is.null(rmd_file)) {
  figures_dir <- file.path('./docs/figure',rmd_file)
  if(dir.exists(figures_dir)) {
    unlink(file.path(figures_dir, "*"))
  }
}
```



## Configure analyses

### Data wrangling
Load SE object and remove outliers
```{r dw-load-se}
se <- readRDS(params$se_path)
```

Modify sample annotation
```{r dw-coldata}
colData(se) %<>% data.frame %>%
  mutate(
    sample_type_grouped = sample_type,
    sample_type_grouped = ifelse(sample_type == 'ctc_cluster_heterotypic_dn', 
                                 'ctc_cluster_heterotypic', 
                                 sample_type_grouped),
    sample_type_grouped_4l = ifelse(sample_type == 'ctc_cluster_wbc', 
                                 'ctc_cluster_heterotypic', 
                                 sample_type_grouped),
    sample_type_grouped_4l = ifelse(sample_type == 'cto_heterotypic', 
                                 'cto', 
                                 sample_type_grouped),
    sample_type_grouped_4l = ifelse(sample_type == 'cto_wbc', 
                                 'cto', 
                                 sample_type_grouped),
    sample_type_grouped_4l = ifelse(sample_type == 'ctc_single_pool', 
                                 'ctc_single', 
                                 sample_type_grouped),
    sample_type_grouped_3l = ifelse(sample_type_grouped == 'ctc_cluster_heterotypic', 
                                 'ctc_cluster', 
                                 sample_type_grouped),
    sample_type_grouped_3l = ifelse(sample_type_grouped == 'cto_heterotypic', 
                                 'cto', 
                                 sample_type_grouped),
    sample_type_grouped_3l = ifelse(sample_type_grouped == 'cto_wbc', 
                                 'cto', 
                                 sample_type_grouped),
    sample_type_grouped_3l = ifelse(sample_type_grouped == 'ctc_single_pool', 
                                 'ctc_single', 
                                 sample_type_grouped)
  ) %>%
  DataFrame
```


Filtering out low-abundance features. Keeping those features with at least 10 counts in at least 50% of samples of the smallest group.
```{r dw-subset-feature-abundance}
use_rows <- filterByExpr(se, group=se$sample_type, min.count = 10, min.total.count = 15, min.prop = 0.5)
se <- se[use_rows,]
```

TMM normalization
```{r dw-tmm}
norm_method <- "TMM"
exprs_mat_for_norm <- assay(se, 'counts')
norm_factors <- calcNormFactors(exprs_mat_for_norm, method = norm_method)
lib_size <- colSums(exprs_mat_for_norm)/10^6
se$size_factor <- (norm_factors * lib_size)
se <- logNormCounts(se, size.factors =se$size_factor)
se <- logNormCounts(se, size.factors =se$size_factor, log = FALSE)
```

CPM calculation
```{r dw-cpm}
assay(se, 'cpm') <- edgeR::cpm(assay(se, 'normcounts'))
assay(se, 'logcpm') <- edgeR::cpm(assay(se, 'normcounts'), log = TRUE)
```


### Define of comparisons

#### Process data
```{r conf-comparisons}
x <- colData(se) %>% data.frame

# Define groups
ctc_single_samples <- x %>% 
  filter(sample_type %in% c('ctc_single', 'ctc_single_pool')) %>% 
  pull(sample_alias)
ctc_cluster_samples <- x %>% 
  filter(sample_type == 'ctc_cluster') %>% 
  pull(sample_alias)
ctc_cluster_comb_samples <- x %>% 
  filter(grepl('ctc_cluster', sample_type)) %>% 
  pull(sample_alias)
ctc_cluster_hetero_samples <- x %>% 
  filter(sample_type %in% c('ctc_cluster_heterotypic', 'ctc_cluster_heterotypic_dn', 'ctc_cluster_wbc')) %>% 
  pull(sample_alias)
cto_samples <- x %>% 
  filter(sample_type %in% c('cto', 'cto_heterotypic', 'cto_wbc')) %>% 
  pull(sample_alias)
```

Define initial comparisons without batch correction
```{r conf-comparisons-unbathced}
comp_list <- list(
  `CTO-over-S` = list(
    CTO = cto_samples,
    Single = ctc_single_samples
  ),
  `CTO-over-Cl` = list(
    CTO = cto_samples,
    Cluster = ctc_cluster_comb_samples
  )
)

comp_list_description <- data.frame(
  comparison = names(comp_list),
  comparison_long = c(
    "CTO-over-Single",
    "CTO-over-Cluster"
  )
)
```

Add comparisons with batch correction. Only using batch levels with samples in both categories and comparisons with more than 2 levels of the batch variable
```{r conf-comparisons-batched}
batch_var <- 'donor'
batch_levels <- unique(x[,batch_var])
for(cn in names(comp_list)) {
  i <- comp_list[[cn]]
  x$batch <- x[,batch_var]
  sample_batch_levels <- x %>% 
    filter(sample_alias %in% i[[1]]) %>% 
    pull(batch_var) %>% 
    unique
  ctrl_batch_levels <- x %>% 
    filter(sample_alias %in% i[[2]]) %>% 
    pull(batch_var) %>% 
    unique
  use_batch_levels <- intersect(sample_batch_levels, ctrl_batch_levels)
  sample_samples <-  x %>% 
    filter(sample_alias %in% i[[1]]) %>% 
    filter(batch %in% use_batch_levels) %>% 
    pull(sample_alias)
  ctrl_samples <-  x %>% 
    filter(sample_alias %in% i[[2]]) %>% 
    filter(batch %in% use_batch_levels) %>% 
    pull(sample_alias)
  if(length(use_batch_levels) > 1 & 
      length(sample_samples) > 0 & 
      length(ctrl_samples) > 0) {
    # Add comparison to list of comparison
    res <- list(
      sample = sample_samples,
      ctrl = ctrl_samples,
      batch = batch_var
    )
    names(res)[1:2] <- names(i)[1:2]
    comp_list[[paste0(cn, '-', batch_var)]] <- res
    # Add comparison long name description
    new_desc <- comp_list_description %>% 
      filter(comparison == cn) %>%
      mutate(comparison_long = paste(comparison_long, 'batch', batch_var),
             comparison = paste0(cn, '-', batch_var))
    comp_list_description <- rbind(comp_list_description, new_desc)
  }  
}
```

Remove comparisons without batch effect
```{r conf-comparisons-bathced-keep}
use_comps <- grep(batch_var, names(comp_list), value = TRUE)
comp_list <- comp_list[use_comps]
```


#### Table of comparisons

We filter comparisons with n < 3 samples in each group
```{r table-comparisons}
i = names(comp_list)[1]
comp_df <- foreach(i = names(comp_list), .combine = rbind) %do% {
  icomp <- comp_list[[i]]
  data.frame(
    comparison = i,
    `Sample Group (n)` =  length(icomp[[1]]),
    `Reference Group (n)` = length(icomp[[2]]),
    check.names = FALSE
  )
}

# filter comparisons were we have at least 3 samples in each group
comp_df %<>% 
  mutate(Evaluated = `Sample Group (n)` >= 3 & `Reference Group (n)` >= 3)
kepp_comp <- comp_df %>% filter(Evaluated) %>% pull(comparison)
comp_list <- comp_list[kepp_comp]

# Add long names
comp_df <- comp_df %>% left_join(comp_list_description) %>% 
  mutate(comparison = comparison_long) %>%
  dplyr::select(-comparison_long)

comp_df %>%
  datatable(., 
            rownames = FALSE, 
            filter = 'top', 
            caption = 'List of comparisons for differential expression. Comparisons with n < 3 samples in each group won\'t be evaluated',
            extensions = 'Buttons', 
            options = list(
              dom = 'Blfrtip',
              buttons = c('csv', 'excel')
            ))

```

## Data generation

### Run differential expression analysis
Get Ensembl gene info
```{r de-ensembl, eval = TRUE}
ensembl <-  biomaRt::useEnsembl(biomart="genes", dataset="hsapiens_gene_ensembl",
                                mirror = 'useast')
gene_desc <- biomaRt::getBM(attributes=c('external_gene_name','description'),
                              filters = 'external_gene_name',
                              values = rowData(se)$gene_name, mart =ensembl) %>%
    dplyr::rename('gene_name' = 'external_gene_name') %>%
    unique

gene_desc <- gene_desc[!duplicated(gene_desc$gene_name),]
saveRDS(gene_desc, file = file.path(dirname(params$se_path), 'ensembl_gene_desc.rds'))
```

```{r de-ensembl-load, eval = TRUE}
gene_desc <- readRDS(file = file.path(dirname(params$se_path), 'ensembl_gene_desc.rds'))
```

Run differential expression analysis using EdgeR QLF
```{r edgeR-QLF-run, eval = TRUE}
use_comp <-  comp_list[[1]]
dge_list <- foreach(use_comp = comp_list) %do%{
  se_cols <- se$sample_alias %in% unlist(use_comp)
  use_se <- se[,se_cols]
  use_se$group <- ifelse(
    use_se$sample_alias %in% use_comp[[1]],
    names(use_comp)[1],
    names(use_comp)[2]
  )
  
  if(is.null(use_comp$batch)) {
    dge <- edgeR_dge(
      use_se,
      # Design configuration for differential expression
      group_var =  'group',
      group_sample = names(use_comp)[1],
      group_ref = names(use_comp)[2],
      design_formula = "~ group",
      coef = 'last',
      # Conversion from SingleCellExperiment/Summarized Experiment to DGEList
      assay_to_DGEList = 'counts',
      assay_to_row_filter = "counts",
      # Feature filtering parameters
      use_filterByExpr = TRUE,
      min_counts = params$min_counts,
      min_present_prop = params$min_present_prop,
      # EdgeR workflow configuration
      run_calcNormFactors = 'TMM',
      estimateDisp_robust = TRUE,
      estimateDisp_trend.method = "locfit",
      glm_approach = "QLF",
      # Output configuration
      adjust_method = 'BH'
      )
    } else {
      design_formula = paste0("~ ", use_comp$batch," + group")
      dge <- edgeR_dge(
        use_se,
        # Design configuration for differential expression
        group_var =  'group',
        group_sample = names(use_comp)[1],
        group_ref = names(use_comp)[2],
        batch_vars = use_comp$batch,
        design_formula = design_formula,
        coef = 'last',
        # Conversion from SingleCellExperiment/Summarized Experiment to DGEList
        assay_to_DGEList = 'counts',
        assay_to_row_filter = "counts",
        # Feature filtering parameters
        use_filterByExpr = TRUE,
        min_counts = params$min_counts,
        min_present_prop = params$min_present_prop,
        # EdgeR workflow configuration
        run_calcNormFactors = 'TMM',
        estimateDisp_robust = TRUE,
        estimateDisp_trend.method = "locfit",
        glm_approach = "QLF",
        # Output configuration
        adjust_method = 'BH'
        )
    }
  
  # Add gene description
  use_res <- dge$results %>%  left_join(., gene_desc, relationship = "many-to-many")
  dge$results <- use_res %>% 
    filter(!duplicated(feature)) %>% 
    mutate(rownames = feature) %>% 
    column_to_rownames('rownames')
  
   return(dge)
}

names(dge_list) <- names(comp_list)

# Save objects
saveRDS(dge_list, file = file.path(params$output_dir, 'dge-edgeR_QLF.rds'))
```

### Run functional enrichment
```{r gse-run, eval = TRUE}
dge_list <- readRDS(file.path(params$output_dir, 'dge-edgeR_QLF.rds'))
dge <-  dge_list[[2]]

gse_list <- foreach(dge = dge_list) %do%{
  gse_res <- gse_omnibus(
    feature_names = dge$results$gene_name,
    p = dge$results$FDR,
    fc = dge$results$logFC,
    fc_thrs = 0.5,
    gmt_files = gmt_files_symbols,
    
    annot_db = 'org.Hs.eg.db',
    organism = 'hsapiens',
    kegg_organism = 'hsa',
    reactome_organism = 'human',
    
    run_enricher = FALSE,
    run_enrichGO = FALSE,
    run_enrichKEGG = FALSE,
    run_enrichReactome = FALSE,
    
    run_GSEA = TRUE,
    run_gseGO = TRUE,
    run_gseKEGG = FALSE,
    run_gseReactome = FALSE,
    
    go_ontologies = c('BP', 'MF'),
    go_simplify_p = 0.05,
    args_gse = list(minGSSize = 10, maxGSSize = 250, 
                    pvalueCutoff = 1, 
                    BPPARAM = BiocParallel::MulticoreParam(1))
  )
  
  return(gse_res)
  
}
names(gse_list) <- names(dge_list)

# Save objects
saveRDS(gse_list, file = file.path(params$output_dir,'gse_omnibus-edgeR_QLF.rds'))
```


### Load data
```{r dge-gse-load}
dge_list <- readRDS(file = file.path(params$output_dir, 'dge-edgeR_QLF.rds'))
gse_list <- readRDS(file = file.path(params$output_dir, 'gse_omnibus-edgeR_QLF.rds'))
```

