---
title: "Analysis of PDAC CTC patient samples"
subtitle: "Variant calling"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
params:
  date: '`r format(Sys.Date(), "%B %d, %Y")`'
  data_dir: ./data/rnaseq/patients-ctc/haplotype_caller
  se_path: ./data/rnaseq/patients-ctc/se_qc.rds
  output_dir: ./output/rnaseq/patients-ctc/variant_calling
---

## Load libraries, additional functions and data

Setup environment
```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(results='asis', echo=TRUE, message=FALSE, warning=FALSE, 
                      error=FALSE, fig.align = 'center', dpi = 600, dev = c("png", "pdf"), 
                      engine.opts = list(bash = "-l"))

options(stringsAsFactors = FALSE)

use_seed <- 1100101
set.seed(use_seed)

dir.create(params$output_dir, recursive = TRUE, showWarnings = FALSE)
```

Load packages
```{r load-libraries}
library(tidyverse)
library(showtext)
library(foreach)
library(DT)
library(knitr)
library(kableExtra)
library(cowplot)
library(colorblindr)
library(RColorBrewer)
library(ggbeeswarm)
library(ggrepel)
library(magrittr)
library(janitor)
library(VariantAnnotation)
library(ComplexHeatmap)
library(circlize)
library(edgeR)
library(scuttle)
library(HoneyBADGER)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(biomaRt)
library(ggvenn)
```

Load ggplot theme
```{r ggplot-theme}
source("./configuration/rmarkdown/ggplot_theme.R")
```

Set color palettes
```{r, echo = FALSE}
source("./configuration/rmarkdown/color_palettes.R")
```

Load custom functions
```{r load-functions}
source('./code/R-functions/gse_omnibus.r')
source('./code/R-functions/gse_report.r')
source('./code/R-functions/subchunkify.R')
```


Clean files generated in previous runs
```{r clean-files}
rmd_file <- current_input()
if(!is.null(rmd_file)) {
  figures_dir <- file.path('./docs/figure',rmd_file)
  if(dir.exists(figures_dir)) {
    unlink(file.path(figures_dir, "*"))
  }
}
```

## Data processing

Load data SummarizedExperiment object
```{r load-se}
se <- readRDS(params$se_path)
```

Load donor VCF files
```{r load-vcf, eval = FALSE}
donor_levels <- unique(se$donor)
rna_vcf <- foreach(donor = donor_levels) %do% {
  vcf_file <- file.path(
    params$data_dir,
    paste0(donor, '.ann.dbsnp.cosmic.vcf.gz')
    )
  res <- readVcf(vcf_file, 'GRCh38')
  names(res) %<>% gsub(';', '.', .)
  res
}
names(rna_vcf) <- donor_levels

saveRDS(rna_vcf, file = file.path(params$output_dir, 'rna_vcf.rds'))
```

Load donor TSV files
```{r load-tsv, eval = FALSE}
donor_levels <- unique(se$donor)
rna_tsv <- foreach(donor = donor_levels) %do% {
  tsv_file <- file.path(
    params$data_dir,
    paste0(donor, '.ann.dbsnp.cosmic.tsv.gz')
    )
  ## load data
  res <- read_tsv(tsv_file, comment = '#')
  res_col_spc <- read_tsv(tsv_file, comment = '#', col_types=cols_only(SAMPLE_COUNT = "n", GENE = "c"))
  res$SAMPLE_COUNT <- res_col_spc$SAMPLE_COUNT
  res$GENE <- res_col_spc$GENE
  rm(res_sample_count)
  
  ## modify colnames
  colnames(res) %<>% 
    gsub('\\.\\.\\..*', '', .) %>% 
    gsub('\\[.*\\]', '', .) %>% 
    make.names(unique = TRUE)
  ## modify data
  res %<>% 
    ## add ID to is.na(ID)==TRUE
    mutate(ID = ifelse(
        is.na(ID), 
        paste0(CHROM, '.', POS, '_', REF, '.',ALT), 
        ID)
      ) %>% 
    ## in ID, convert ';' to '.' to match VCF ID
    mutate(
      ID = gsub(';', '.', ID),
      IDc = paste(ANN.GENE, ANN.HGVS_C),
      IDp = paste(ANN.GENE, ANN.HGVS_P),
      ANN.IMPACT = factor(ANN.IMPACT, levels = c('HIGH', 'MODERATE', 'LOW', 'MODIFIER'))
    )

  # Annotate somatic variants
  res %<>% 
    mutate(somatic = ifelse(!G5 & SAMPLE_COUNT > 1, TRUE, FALSE))
  
  res
}

names(rna_tsv) <- donor_levels


saveRDS(rna_tsv, file = file.path(params$output_dir, 'rna_tsv.rds'))

```

Load processed files
```{r load-processed-data}
rna_vcf <- readRDS(file.path(params$output_dir, 'rna_vcf.rds'))
rna_tsv <- readRDS(file.path(params$output_dir, 'rna_tsv.rds'))
```


Load PDAC genes from cBioportal
```{r load-pdac-genes}
pdac_cbio <- read_tsv("./data/external/pdac_cbioportal_mutated_genes.txt") %>% 
  dplyr::rename(gene = Gene) %>% 
  mutate(cbio_freq = gsub("\\%", "", Freq) %>% as.numeric) %>%
  dplyr::select(gene, cbio_freq, `Is Cancer Gene (source: OncoKB)`) |> 
  filter(`Is Cancer Gene (source: OncoKB)` == 'Yes' & !is.na(cbio_freq))
```


## Tables : Putative somatic variants
Criteria for somatic mutation definition:

  - More than 5% minor allele frequency in 1+ populations
  
  - Present in COSMIC with two or more occurrences
  
```{r rna-somatic-table}
# for(donor in names(rna_tsv)) {
rna_tsv_somatic <- foreach(donor = names(rna_tsv), .combine = rbind) %do% {
  res <- rna_tsv[[donor]] %>% 
    mutate(donor = donor) %>% 
    filter(somatic) %>%
    dplyr::rename(cosmic_sample_count = SAMPLE_COUNT) %>% 
    dplyr::select(donor, everything())
  
  # add number of sampels mutated in the study
  use_snpmat <- genotypeToSnpMatrix(rna_vcf[[donor]])$genotypes %>% data.frame
  use_snpmat_n <- use_snpmat[,res$ID] %>% 
    mutate_all(as.numeric)
  add_counts <- data.frame(sample_count = colSums(use_snpmat_n > 1)) %>% 
    rownames_to_column('ID')
  
  res %>% 
    left_join(add_counts)

}

res_tab <- rna_tsv_somatic %>%
  dplyr::select(
    donor, GENE, QUAL, sample_count, ANN.HGVS_C, ANN.HGVS_P, 
    cosmic_sample_count, ID, ANN.EFFECT, 
    ANN.IMPACT, TIER, CAF) %>% 
  rename_with(function(x) gsub("ANN.", "", x)) %>% 
  clean_names() %>% 
  mutate(impact_n = as.numeric(impact)) %>% 
  left_join(pdac_cbio)


res_tab %>% 
  mutate(caf = ifelse(is.na(caf), 0, caf)) %>% 
  datatable(., 
            rownames = FALSE, 
            filter = 'top', 
            caption = 'Potential somatic mutations identified with RNA-seq',
            extensions = 'Buttons', 
            options = list(
              dom = 'Blfrtip',
              buttons = c('csv', 'excel')
            ))
```

## Figures : Putative somatic variants

Criteria for somatic mutation definition:

  - More than 5% minor allele frequency in 1+ populations
  
  - Present in COSMIC with two or more occurrences
  
  - Effect : High or Moderate impact
  
  - Gene present in the list of cBioPortal selected PDAC genes as cancer gene in OncoKB
  
  - Remove HLA genes

```{r rna-somatic-txomic-data}
samples_to_remove <- colnames(se)[se$outlier_discard]
donor_levels <- unique(se$donor)
rna_somatic_var_data <- foreach(donor = donor_levels) %do% {
  use_vcf <- rna_vcf[[donor]]
  use_snpmat <- genotypeToSnpMatrix(use_vcf)
  use_var_annot <- rna_tsv[[donor]]
  
  # Extract Genotypes
  use_geno <- use_snpmat$genotypes %>% data.frame
  
  # Filter variants 
  use_variants <- use_var_annot %>% 
    filter(somatic) %>%
    filter(GENE %in% pdac_cbio$gene) %>%
    filter(ANN.IMPACT %in% c('HIGH', 'MODERATE')) %>%
    filter(!grepl("^HLA", GENE)) %>%
    pull(ID) %>% unique
  
  if(length(use_variants) == 0) {
    NULL
  } else{
    # Extract DP (Approximate read depth) 
    use_dp <- geno(use_vcf)$DP
    use_dp[is.na(use_dp)] <- 0
    use_dp <- use_dp[colnames(use_snpmat$genotypes),] %>% data.frame
    use_dp <- t(use_dp)
    
    # Extract AD (allelic depths) for the ref and alt alleles in the order listed 
    use_ad0 <- geno(use_vcf)$AD
    use_ad0 <- use_ad0[use_variants,, drop = FALSE]
    use_ad <- apply(use_ad0, 2, function(xi) {
      x <- xi[[1]]
      res_i <- foreach(x = xi) %do% {
        if(length(x) >= 2) {
          res <- unlist(x) %>% paste(collapse = ',')
        } else {
          res <- NA
        }
        if(grepl("^NA", res))
          res <- NA
        res
      }
      unlist(res_i)
    })
    if(!is.matrix(use_ad)) {
      use_ad <- as.matrix(use_ad) %>% t()
    }
    rownames(use_ad) <- rownames(use_ad0)
    use_ad <- use_ad %>% data.frame
    use_ad <- t(use_ad)
    
    # Extract Allelic proportion from AD (allelic depths) for the ref and alt alleles in the order listed 
    use_ad0 <- geno(use_vcf)$AD
    use_ad0 <- use_ad0[use_variants,, drop = FALSE]
    xi <- use_ad0[,2]
    use_ap <- apply(use_ad0, 2, function(xi) {
      x <- xi[[1]]
      res_i <- foreach(x = xi) %do% {
        if(length(x) >= 2) {
          x[2] / sum(x)
        } else {
          NA
        }
      }
      unlist(res_i)
    })
    if(!is.matrix(use_ap)) {
      use_ap <- as.matrix(use_ap) %>% t()
    }
    rownames(use_ap) <- rownames(use_ad0)
    use_ap <- use_ap %>% data.frame
    use_ap <- t(use_ap)
    
    # Filter variants 
    use_geno <- use_geno[,use_variants, drop = FALSE]
    use_ad <- use_ad[,use_variants, drop = FALSE]
    use_ap <- use_ap[,use_variants, drop = FALSE]
    use_dp <- use_dp[,use_variants, drop = FALSE]
    
    # Sort Annot by column order in use_geno
    use_var_annot <- use_var_annot[match(colnames(use_geno), use_var_annot$ID),,drop = FALSE]
    # colnames(use_geno)[colnames(use_geno) != use_var_annot$ID]
    
    # # Filter variants by QUAL
    # use_variants <- use_var_annot %>% filter(QUAL >= 20) %>% 
    #   pull(ID) %>% unique
    # # table(use_variants %in% colnames(use_geno))
    # use_geno <- use_geno[,use_variants]
    
    # Sort Annot by column order in use_geno
    # use_var_annot <- use_var_annot[match(colnames(use_geno), use_var_annot$ID),]
    # colnames(use_geno)[colnames(use_geno) != use_var_annot$ID]
    
    # New variant names
    gene_info <- use_var_annot %>% 
      dplyr::select(ID, ANN.GENE, ANN.HGVS_C, ANN.HGVS_P) %>% 
      unique %>% 
      mutate(
        newID = paste(ANN.GENE, ANN.HGVS_C)
      ) %>% 
      filter(ID %in% colnames(use_geno))
    
    # gene_info %>% filter(ID == 'rs1799943.COSV61525225')
    
    # Some ID are repeated in the annotation (eg.:same variant in overlapping genes)
    # Duplicate columns from use_geno to match annotation
    use_geno <- use_geno[,gene_info$ID, drop = FALSE]
    colnames(use_geno) <- gene_info$newID
    use_ad <- use_ad[,gene_info$ID, drop = FALSE]
    colnames(use_ad) <- gene_info$newID
    use_ap <- use_ap[,gene_info$ID, drop = FALSE]
    colnames(use_ap) <- gene_info$newID
    use_dp <- use_dp[,gene_info$ID, drop = FALSE]
    colnames(use_dp) <- gene_info$newID
    
    # # Remove samples
    # samples_to_keep <- setdiff(rownames(use_geno), samples_to_remove)
    # use_geno <- use_geno[samples_to_keep,]
    # use_ad <- use_ad[samples_to_keep,]
    # use_ap <- use_ap[samples_to_keep,]
    # use_dp <- use_dp[samples_to_keep,]
    # 
    res <- list(
      geno = use_geno,
      ad = use_ad,
      ap = use_ap,
      dp = use_dp
    )
    
    rm(use_vcf, use_snpmat, use_var_annot, use_geno)
    
    res
  }
}
names(rna_somatic_var_data) <- donor_levels
```

```{r rna-somatic-txomic-plot-oncoprint, echo=FALSE, results='asis'}
knitr::opts_knit$set(
  base.dir = "docs"
  )

donor_levels <- unique(se$donor)
op_list <- list()
for(donor in donor_levels) {
  use_geno <- rna_somatic_var_data[[donor]]$geno
  if(!is.null(use_geno)) {
    mat <- t(use_geno)
    mat[mat == '00'] <- ''
    colnames(mat) <- gsub(".*-", "", colnames(mat))
    colnames(mat) <- gsub("^SA_", "", colnames(mat))
    mat <- mat[,colnames(mat) %in% colnames(se), drop = FALSE]
    
    #Set col roder
    column_ord <- apply(mat, 2, function(x) sum(x %in% c("02", "03"))) %>% 
      sort(decreasing = TRUE) %>% names
    # Set row order
    # mut_prop <- apply(mat, 1, function(x) sum(x %in% c("02", "03")) / sum(x != ''))
    mut_n <- apply(mat, 1, function(x) sum(x %in% c("02", "03")))
    row_ord <- sort(mut_n, decreasing = TRUE) %>% names
    
    # col = c("01" = "#4575b4", "02" = "#fdae61", "03" = "#d73027")
    col = c("01" = "#e0e0e0", "02" = palette_OkabeIto[1], "03" = palette_OkabeIto[2])
    alter_fun = list(
      background = alter_graphic("rect", fill = "white"),
      `01` = alter_graphic("rect", fill = col["01"]),
      `02` = alter_graphic("rect", fill = col["02"]),
      `03` = alter_graphic("rect", fill = col["03"])
    )
    
    column_title = paste0(donor, ": Somatic RNA-seq variants")
    heatmap_legend_param = list(title = "Alterations", at = c("01", "02", "03"), 
                                labels = c("0/0", "0/1", "1/1"))
    
    op <- oncoPrint(mat,
                    alter_fun = alter_fun, 
                    col = col, 
                    column_title = column_title, 
                    heatmap_legend_param = heatmap_legend_param,
                    show_column_names = TRUE,
                    column_order = column_ord,
                    row_order = row_ord)
    
    op_list[[donor]] <- op
    # draw(op)
    subchunkify(op, 
                fig.width = 7.5, 
                fig.height = (7.5*0.3) + nrow(mat)*0.3,
                dpi = 600,
                prefix = 'varcall-oncoprint',
                suffix = donor )
    cat("\n\n")
  }
}
```
