---
title: "Overall survival in TCGA-PDAC3 project"
subtitle: "CTC hotspots module genes in PDAC-3"
author: "Francesc Castro-Giner"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
params:
  date: '`r format(Sys.Date(), "%B %d, %Y")`'
  data_dir: ./data/resources/UCSCXena/SummarizedExperiment
  donor: PDAC-3
  output_dir: ./output/rnaseq-spatial/pdac3/seurat_hs/survival/bc_group_cluster
  findmarkers_file: ./output/rnaseq-spatial/pdac3/seurat_hs/results_findMarkers_bc_group_cluster.rds
  time_var: DSS.time
  event_var: DSS
  min_event_n: 3
  min_surv_n: 20
  max_time_years: 5
---

## Load libraries, additional functions and data

Setup environment
```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(results='asis', echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.align = 'center', fig.width = 3.5, fig.asp = 0.618, dpi = 600, dev = c("png", "pdf"), engine.opts = list(bash = "-l"))

options(stringsAsFactors = FALSE)

use_seed <- 1100101

set.seed(use_seed)

dir.create(params$output_dir, recursive = TRUE, showWarnings = FALSE)
```

Load packages
```{r load-libraries}
library(tidyverse)
library(knitr)
library(foreach)
library(SummarizedExperiment)
library(survival)
library(gtsummary)
library(ggsurvfit)
library(survminer)
library(meta)
library(magrittr)
library(GSVA)
library(kableExtra)
library(DT)
```

Load ggplot theme
```{r ggplot-theme, include=TRUE}
source("./configuration/rmarkdown/ggplot_theme.R")
```
Set color palettes
```{r, echo = FALSE}
source("./configuration/rmarkdown/color_palettes.R")
```

Clean files generated in previous runs
```{r clean-files}
rmd_file <- current_input()
if(!is.null(rmd_file)) {
  figures_dir <- file.path('./docs/figure',rmd_file)
  if(dir.exists(figures_dir)) {
    unlink(file.path(figures_dir, "*"))
  }
}
```

## Data wrangling

Generate signatures based on CTC spatial distribution. Here we use the findMarkers results for each group (CTC1, CTC2 , and ST) in the spatial analysis of PDAC3. For each group,  gene signature was defined as top-100 genes upregulated in each group (FDR < 0.0001 and logFC > 1; genes ranked by FDR). 
```{r gene-spatial-signatures-3g-generate}
topn <- 100
p_thresh <- 0.0001
logfc_thresh <- 1
min_pct_1 <- 0.10
gene_sets <- list()

chunk_suffix <- 'bc_group_cluster'
dge_list <- readRDS(params$findmarkers_file)
use_dge <-  dge_list[[1]]
gene_sets <- foreach(use_dge = dge_list) %do% {
  use_dge$result %>%
    filter(pct.1 >= min_pct_1 ) %>% 
    filter(p_val_adj < p_thresh & 
             avg_log2FC > logfc_thresh) %>% 
    arrange(p_val_adj) %>% 
    head(topn) %>%
    pull(gene)
}
names(gene_sets) <- names(dge_list)
```

Define list of projects
<!-- https://gdc.cancer.gov/resources-tcga-users/tcga-code-tables/tcga-study-abbreviations -->
```{r define-projects}
# project_id <- list.files(
#   path = params$data_dir, 
#   pattern = 'TCGA-[A-Z]+.rds$'
#   ) %>%
#   gsub(".rds", "", .)
# 
# tcga_solid <- c('ACC', 'BLCA', 'BRCA', 'CESC', 'CHOL', 'COAD', 'ESCA', 'GBM', 
#                   'HNSC', 'KICH', 'KIRC', 'KIRP', 'LGG', 'LIHC', 'LUAD', 'LUSC', 
#                   'MESO', 'OV', 'PAAD', 'PCPG', 'PRAD', 'READ', 'SARC', 'SKCM', 
#                   'STAD', 'TGCT', 'THCA', 'THYM', 'UCEC', 'UCS', 'UVM') %>% 
#   paste('TCGA', ., sep = '-')
# 
# project_id <- intersect(project_id, tcga_solid)
project_id <- 'TCGA-PAAD'
```

Generate signature score per patient
```{r generate-signature-score, eval = FALSE}
gsva_res <- foreach(i = project_id) %do% {
  cat("###", i, "\n")
  se <- readRDS(file.path(params$data_dir, paste0(i, '.rds')))

  # Calculate CPM
  cpm_mat <- scuttle::calculateCPM(se, assay.type = "normcounts")
  logcpm_mat <- log2(cpm_mat + 1)
  rm(cpm_mat)

  # Remove duplicated samples: https://ucsc-xena.gitbook.io/project/how-do-i/how-do-i-remove-duplicate-samples-from-a-km-plot
  #logcpm_mat <- logcpm_mat[, se$sample_type == 'Primary Tumor']

  # Signature gene list
  gene_sets_idx <- lapply(gene_sets, function(x) rownames(se)[rowData(se)$gene_name %in% x])

  # Generate GSVA scores
  gsva_params <- gsvaParam(logcpm_mat,
                           gene_sets_idx,
                           kcdf="Gaussian",
                           maxDiff = TRUE)

  gsva_res_i <- gsva(gsva_params)
  
  # saveRDS(gsva_res_i, file = file.path(params$output_dir, paste0(i,'_gsva.rds')))
  rm(se)
  rm(logcpm_mat)
  gsva_res_i %>% t %>% data.frame(check.names = FALSE)
}
names(gsva_res) <- project_id
saveRDS(gsva_res, file = file.path(params$output_dir, 'gsva_res.rds'))
```

Load GSVA results
```{r load-gsva-res}
gsva_res <- readRDS(file.path(params$output_dir, 'gsva_res.rds'))
# use_data <- foreach(i = project_id) %do% {
#   # Extract sample annotation and add gene expression and signature score
#   se <- readRDS(file.path(params$data_dir, paste0(i, '.rds')))
#   sample_annot <- colData(se) %>% data.frame %>% 
#     bind_cols(gsva_res[[i]]) 
#   rm(se)
#   sample_annot
# }
# names(use_data) <- project_id
```


## Run survival analysis
```{r conf-survival-analysis}
selected_modules <- names(gene_sets)
# project_id <- list.files(
#   path = params$data_dir, 
#   pattern = 'TCGA-[A-Z]+.rds$'
#   ) %>%
#   gsub(".rds", "", .)
# 
# tcga_solid <- c('ACC', 'BLCA', 'BRCA', 'CESC', 'CHOL', 'COAD', 'ESCA', 'GBM', 
#                   'HNSC', 'KICH', 'KIRC', 'KIRP', 'LGG', 'LIHC', 'LUAD', 'LUSC', 
#                   'MESO', 'OV', 'PAAD', 'PCPG', 'PRAD', 'READ', 'SARC', 'SKCM', 
#                   'STAD', 'TGCT', 'THCA', 'THYM', 'UCEC', 'UCS', 'UVM') %>% 
#   paste('TCGA', ., sep = '-')
# 
# project_id <- intersect(project_id, tcga_solid)
project_id <- 'TCGA-PAAD'
```

```{r, eval = FALSE}
foreach(i = project_id) %do% {
   se <- readRDS(file.path(params$data_dir, paste0(i, '.rds')))
  
  # Remove duplicated samples: https://ucsc-xena.gitbook.io/project/how-do-i/how-do-i-remove-duplicate-samples-from-a-km-plot
  sample_annot <- colData(se) %>% data.frame %>% 
    filter(sample_type == 'Primary Tumor')
  'Metastasis_nature2012' %in% names(sample_annot)
  
}
```

Load raw data. We set up the follow-up limit and configure survival variable. Remove studies with less than `r params$min_event_n` events and `r params$min_surv_n` patients with survival data
```{r load-data, eval = FALSE}
i <- project_id[1]
data_se <- foreach(i = project_id) %do% {
  se <- readRDS(file.path(params$data_dir, paste0(i, '.rds')))
  
  # Remove duplicated samples: https://ucsc-xena.gitbook.io/project/how-do-i/how-do-i-remove-duplicate-samples-from-a-km-plot
  sample_annot <- colData(se) %>% data.frame %>% 
    filter(sample_type == 'Primary Tumor')
  
  # sample_annot$Metastasis_Coded_nature2012 %>% table
  # sample_annot$Metastasis_nature2012 %>% table
  # sample_annot$distant_metastasis_present_ind2  %>% table
  # sample_annot$metastatic_site_at_diagnosis  %>% table
  # sample_annot$new_neoplasm_event_type %>% table
  # sample_annot$pathologic_stage %>% table
  
  # Filter genes and samples
  res <- se[, rownames(sample_annot)]
  
  # Set survival variables and follow up limit
  sample_annot <- colData(res) %>% data.frame()
  sample_annot$status <- sample_annot[[params$event_var]]
  sample_annot$time <- sample_annot[[params$time_var]]/365.25
    
  sample_annot <- sample_annot %>% 
    mutate(
      status = ifelse(time >= params$max_time_years, 0, status),
      time = ifelse(time >= params$max_time_years, params$max_time_years, time)
    )
  
  # Add modules scores
  sample_annot %<>% 
    bind_cols(gsva_res[[i]][rownames(sample_annot),]) 
  
  # Add back colData
  colData(res) <- DataFrame(sample_annot)
    
  # Number of events
  n_surv_data <- sum(!is.na(res$status))
  n_events <- sum(res$status, na.rm = TRUE)

  if(ncol(res) > 0 & n_events >= params$min_event_n & n_surv_data >= params$min_surv_n) {
    res
  } else {
    NULL
  }
}
names(data_se) <- project_id

# Remove NULL elements
data_se <- compact(data_se)
```

Run survival analysis dichotonomous
```{r run-survival-quantile-analysis, eval = FALSE}
i <- project_id[3]
surv_q3 <- foreach(i = names(data_se), .combine = rbind) %do% {
  se <- data_se[[i]]
  
  # Add gene-expression data
  g <- selected_modules[1]
  foreach(g = selected_modules, .combine = rbind) %do% {
    
    # Generate clinical sample data with survival information
    sample_annot <- colData(se) %>% data.frame()
    sample_annot$ml_score <- sample_annot[[g]]
    use_quantiles <- quantile(sample_annot[[g]], probs = c(0.33, 0.66), na.rm = FALSE)

    # sample_annot$status <- sample_annot[[params$event_var]]
    # sample_annot$time <- sample_annot[[params$time_var]]/365.25
    sample_annot %<>% 
      mutate(
        gene_q = ifelse(ml_score >= use_quantiles[2], "T3", NA),
        gene_q = ifelse(ml_score <= use_quantiles[1], "T1", gene_q)
      ) %>% 
      filter(!is.na(gene_q))
    
    # Set follow up limit
    # sample_annot <- sample_annot %>% 
    #   mutate(
    #     status = ifelse(time >= params$max_time_years, 0, status),
    #     time = ifelse(time >= params$max_time_years, params$max_time_years, time)
    # )
    
    # Cox-PH results
    cox.res <- coxph(Surv(time, status) ~ gene_q, data = sample_annot) %>%
      tbl_regression(exp = TRUE)
    
    data.frame(
        project_id = i,
        module_name = g,
        HR = cox.res$table_body$estimate[3],
        p.value = cox.res$table_body$p.value[3],
        std.error = cox.res$table_body$std.error[3],
        conf.low = cox.res$table_body$conf.low[3],
        conf.high = cox.res$table_body$conf.high[3],
        n.low = sum(sample_annot$gene_q == "Q1"),
        n.high = sum(sample_annot$gene_q == "Q2")
      )
  }

}

saveRDS(surv_q3, file = file.path(params$output_dir, 'surv_q3.rds'))

```

Generate survfit object for dichotonomous survival analysis
```{r run-surfit-quantile-analysis, eval = FALSE}
survfit_km_q3 <- foreach(i = names(data_se)) %do% {
  se <- data_se[[i]]
  
  # Add gene-expression data
  g <- selected_modules[1]
  res_by_genes <- foreach(g = selected_modules) %do% {
    # Extract gene expression data
    gene_id <- rownames(se)[rowData(se)$gene_name == g]
    gene_exp <- assay(se, 'logcounts')[gene_id,]
    
    # Generate clinical sample data with survival information
    sample_annot <- colData(se) %>% data.frame()
    sample_annot$ml_score <- sample_annot[[g]]
    use_quantiles <- quantile(sample_annot[[g]], probs = c(0.33, 0.66), na.rm = FALSE)
    # sample_annot$status <- sample_annot[[params$event_var]]
    # sample_annot$time <- sample_annot[[params$time_var]]/365.25
    sample_annot %<>% 
      mutate(
        gene_q = ifelse(ml_score >= use_quantiles[2], "T3", NA),
        gene_q = ifelse(ml_score <= use_quantiles[1], "T1", gene_q)
      ) %>% 
      filter(!is.na(gene_q))
    
    # Set follow up limit
    # sample_annot <- sample_annot %>% 
    #   mutate(
    #     status = ifelse(time >= params$max_time_years, 0, status),
    #     time = ifelse(time >= params$max_time_years, params$max_time_years, time)
    # )
    
    # KM plot
    survfit2(Surv(time, status) ~ gene_q, data = sample_annot) %>% 
      ggsurvfit() +
      add_risktable() +
      labs(
        x = "Time (years)",
        title = i
      ) +
      theme_project + 
      theme(legend.position="bottom")
  }
  names(res_by_genes) <- selected_modules
  res_by_genes

}
names(survfit_km_q3) <- names(data_se)
saveRDS(survfit_km_q3, file = file.path(params$output_dir, 'survfit_km_q3.rds'))

```

Generate survfit object for dichotonomous survival analysis, combining all the signatures into a single plot
```{r run-surfit-quantile-analysis-combined, eval = FALSE}
survfit_km_combined_q3 <- foreach(i = names(data_se)) %do% {
  se <- data_se[[i]]
  
  sample_annot <- colData(se) %>% data.frame()
  # Add gene-expression data
  g <- selected_modules[1]
  
  for(g in selected_modules){
    gene_id <- rownames(se)[rowData(se)$gene_name == g]
    gene_exp <- assay(se, 'logcounts')[gene_id,]
    use_quantiles <- quantile(sample_annot[[g]], probs = c(0.33, 0.66), na.rm = FALSE)
    
    q_var <- paste0(g, "_q")
    sample_annot[,q_var] <- ifelse(sample_annot[[g]]  >= use_quantiles[2], "T3", NA)
    sample_annot[,q_var] <- ifelse(sample_annot[[g]]  <= use_quantiles[1], "T1", sample_annot[,q_var])
  }
  
  q_vars <- paste0(selected_modules, "_q")
  
  df_long <- sample_annot %>%
    pivot_longer(
      cols = all_of(q_vars),
      names_to = "predictor",
      values_to = "group"
    ) %>% 
    mutate(
      group = factor(group, levels = c("T1", "T3")),
      predictor = gsub("_q", "", predictor)
    ) %>% 
    filter(!is.na(group))
  
  # Fit KM curve with interaction predictor:group
  fit <- survfit(Surv(time, status) ~ predictor + group, data = df_long)
  
  return(
    list(
      fit = fit,
      data = df_long
    )
  )
}
names(survfit_km_combined_q3) <- names(data_se)
saveRDS(survfit_km_combined_q3, file = file.path(params$output_dir, 'survfit_km_combined_q3.rds'))

```


Generate survfit object for dichotonomous survival analysis, combining all the CTC signatures into a single plot
```{r run-surfit-quantile-analysis-ctc-combined, eval = FALSE}
use_modules <- c('CTC1', 'CTC2')
survfit_km_ctc_combined_q3 <- foreach(i = names(data_se)) %do% {
  se <- data_se[[i]]
  
  sample_annot <- colData(se) %>% data.frame()
  # Add gene-expression data
  g <- use_modules[1]
  
  for(g in use_modules){
    gene_id <- rownames(se)[rowData(se)$gene_name == g]
    gene_exp <- assay(se, 'logcounts')[gene_id,]
    use_quantiles <- quantile(sample_annot[[g]], probs = c(0.33, 0.66), na.rm = FALSE)
    
    q_var <- paste0(g, "_q")
    sample_annot[,q_var] <- ifelse(sample_annot[[g]]  >= use_quantiles[2], "T3", NA)
    sample_annot[,q_var] <- ifelse(sample_annot[[g]]  <= use_quantiles[1], "T1", sample_annot[,q_var])
  }
  
  q_vars <- paste0(use_modules, "_q")
  
  df_long <- sample_annot %>%
    pivot_longer(
      cols = all_of(q_vars),
      names_to = "predictor",
      values_to = "group"
    ) %>% 
    mutate(
      group = factor(group, levels = c("T1", "T3")),
      predictor = gsub("_q", "", predictor)
    ) %>% 
    filter(!is.na(group))
  
  # Fit KM curve with interaction predictor:group
  fit <- survfit(Surv(time, status) ~ predictor + group, data = df_long)
  
  return(
    list(
      fit = fit,
      data = df_long
    )
  )
}
names(survfit_km_ctc_combined_q3) <- names(data_se)
saveRDS(survfit_km_ctc_combined_q3, file = file.path(params$output_dir, 'survfit_km_ctc_combined_q3.rds'))
```


Load survival data
```{r load-survival-data}
surv_q3 <- readRDS(file.path(params$output_dir, 'surv_q3.rds'))
survfit_km_q3 <- readRDS(file.path(params$output_dir, 'survfit_km_q3.rds'))
survfit_km_combined_q3 <- readRDS(file.path(params$output_dir, 'survfit_km_combined_q3.rds'))
survfit_km_ctc_combined_q3 <- readRDS(file.path(params$output_dir, 'survfit_km_ctc_combined_q3.rds'))
```

Adapt survival data to be used with the template
```{r adapt-survival-data}
surv_q3 <- surv_q3 %>% 
  select(-project_id) %>% 
  dplyr::rename(project_id = module_name)
```


Run metanalysis for quantitative survival analysis
<!-- https://bookdown.org/MathiasHarrer/Doing_Meta_Analysis_in_R/ -->
```{r run-survival-quant-meta-analysis}
## Survival with Q3 expression
meta_surv_q3 <-  metagen(
    HR = log(surv_q3$HR), 
    lower = log(surv_q3$conf.low), 
    upper = log(surv_q3$conf.high),
    sm = "HR", fixed=F, random=T,
    studlab = surv_q3$project_id,
    method.tau = "DL", ## method to calculate Tau
    method.random.ci = "classic" ## method to calculate estimator's CI
  )

x <- summary(meta_surv_q3)
meta_surv_q3_summary <- x$data %>% 
  mutate(
    project_id = 'Summary',
    gene_name = .studlab,
    HR = exp(x$random$TE),
    p.value = x$random$p,
    std.error = x$random$seTE,
    conf.low = exp(x$random$lower),
    conf.high = exp(x$random$upper)
  ) %>% 
  select(
    project_id, gene_name, HR, p.value, std.error, conf.low, conf.high
  )

meta_surv_q3_weight <- data.frame(
    project_id = x$studlab,
    gene_name = 'Module-Up',
    w.random = round(100*x$w.random / sum(x$w.random), 1)
  )

```


## Module scores

### Genes included
```{r m-gene-list}
use_data <- foreach(i = names(gene_sets), .combine = rbind) %do% {
  data.frame(module = i, gene = gene_sets[[i]])
}
use_data %>% datatable(., 
          rownames = FALSE, 
          filter = 'top', 
          caption = 'Genes included for each module',
          extensions = 'Buttons', 
          options = list(
            dom = 'Blfrtip',
            buttons = c('csv', 'excel')
          ))

```


## Quantile 33 versus 66

Cox proportional-hazards model to assess the association between gene expression and overall survival (OS). Expression data was trichotomized into three quantiles (Q1, Q2, Q3) based on the 33rd and 66th percentiles.

```{r coxph-q3-perm}
chunk_prefix <-'q3'
use_cox_data <- surv_q3
use_meta <- meta_surv_q3
```

### Forest plot with ggplot simplified
```{r q3-coxph-ggplot-forestplot-simplified, fig.width = 2, fig.asp = 0.4}
chunk_data <- use_cox_data %>%
  mutate(
    project_id = factor(project_id, levels = c('CTC2', 'CTC1', 'ST'))
  )

project_ord <- chunk_data$project_id %>% levels
  

use_meta_sum <- summary(use_meta)

meta_weight <- data.frame(
  project_id = gsub("TCGA-", "", use_meta_sum$studlab),
  w.fixed = use_meta_sum$w.fixed,
  w.pct = 100* use_meta_sum$w.fixed / sum(use_meta_sum$w.fixed)
)

chunk_data %<>% left_join(meta_weight)


chunk_data %<>%
  mutate(
    project_id = factor(project_id, levels= c('Total', '', project_ord)),
    project_type = ifelse(project_id == 'Total', 'Total', 'Project'),
    FDR = p.adjust(p.value, method = 'BH'),
    logP = -log10(p.value),
    logP = ifelse(logP > 10, 10 , logP),
    logFDR = -log10(FDR),
    logFDR = ifelse(logFDR > 10, 10 , logFDR)
  )

chunk_data %>%
  ggplot(aes(HR, project_id,
             # size = w.pct,
             shape = project_type)
         ) +
  geom_point(size = 3) +
  geom_linerange(aes(xmin=conf.low, xmax=conf.high), 
                 linewidth = 1/2,
                 show.legend = FALSE) +
  geom_vline(xintercept = 1, linetype = 'dotted') +
  scale_shape_manual(values = c(Project = 16, Total = 18)) +
  scale_size_continuous(range = c(1, 3)
                        # breaks = c(0.1, 0.5, 1, 2, 5, 10),
                        ) +
  scale_x_continuous(trans = 'log2',
                     breaks = c(0.1, 0.5, 1, 2, 5, 10),
                     labels = as.character(c(0.1, 0.5, 1, 2, 5, 10))) +
  theme(
    axis.line.y = element_blank(),
    axis.ticks.y = element_blank()
  ) +
  labs(
    x = 'Hazard ratio',
    y = ''
    # size = expression(paste("-lo", g[10],"(P value)")),
    # size = "Weight %",
    ) +
  guides(
    shape = 'none'
  )
 
```


### Kaplan-Meier curves

#### CTC1
```{r coxph-q3-km-CTC1, fig.asp = 1}
g <- 'CTC1'
km_list <- map(survfit_km_q3, g)
print(km_list)
cat("\n\n")
```


#### CTC2

```{r coxph-q3-km-CTC2,fig.asp = 1}
g <- 'CTC2'
km_list <- map(survfit_km_q3, g)

print(km_list)
cat("\n\n")
```


#### ST

```{r coxph-q3-km-ST,fig.asp = 1}
g <- 'ST'
km_list <- map(survfit_km_q3, g)
print(km_list)
cat("\n\n")
```


#### Combined CTC
```{r coxph-q3-km-combined-ctc, fig.width = 3.5, fig.asp = 0.8}
km_res <- survfit_km_ctc_combined_q3$`TCGA-PAAD`

res <- ggsurvplot(
  fit = km_res$fit,
  data = km_res$data,
  color = "predictor",            
  linetype = "group",
  linewidth = linewidth_default,
  palette = bc_group_cluster_palette,
  legend.title = "",
  xlab = "Time",
  ylab = "Survival Probability",
  ggtheme = theme_project,
  censor = FALSE,
  risk.table = T,
  tables.height = 0.40,
  pval = TRUE, 
  pval.size = geom_text_size,
  fontsize = geom_text_size,
  legend.size = geom_text_size
) 

res$plot <- res$plot + 
  scale_linetype_manual(values = c(T3 = "solid", T1 = "dashed"))

print(res)

cat("\n\n")
```
