---
title: "PDAC-3 spatial transcriptomics"
subtitle: "TME distance to barcodes groups"
author: "Francesc Castro-Giner"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
params:
  date: '`r format(Sys.Date(), "%B %d, %Y")`'
  donor: PDAC-3
  output_dir: ./output/rnaseq-spatial/pdac3
  seurat_dir: ./data/rnaseq-spatial/pdac3
  ctc_met_bc_se_file: ./data/barcodes/pdac3-ctc-mets/se.rds
  min_counts: 5
  min_present_prop: 0.25
  msigdb_dir: ./data/resources/MSigDB
  msigdb_vs: v2023.1.Hs
  max_dist: 1000
  crop_dist: 1000
  ncores: 4
---



## Configuration

Load libraries, additional functions and data

Setup environment
```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(results='asis', echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.align = 'center', fig.width = 3.5, fig.asp = 0.618, dpi = 600, dev = c("png", "pdf"), fig.showtext = FALSE, engine.opts = list(bash = "-l"))

options(stringsAsFactors = FALSE)

use_seed <- 1100101
set.seed(use_seed)

dir.create(params$output_dir, recursive = TRUE, showWarnings = FALSE)
```

Load packages
```{r load-libraries}
library(tidyverse)
library(foreach)
library(DT)
library(knitr)
library(patchwork)
library(kableExtra)
library(RColorBrewer)
library(ggbeeswarm)
library(ggrepel)
library(magrittr)
library(UpSetR)
library(GGally)
library(openxlsx)
library(ggpubr)
library(rstatix)
library(arsenal)
#
library(Seurat)
library(SummarizedExperiment)
library(semla)
library(edgeR)
library(scran)
library(ComplexHeatmap)
library(circlize)
library(mgcv)
library(gratia)
```

Load ggplot theme
```{r ggplot-theme}
source("./configuration/rmarkdown/ggplot_theme.R")
```

Set color palettes
```{r, echo = FALSE}
source("./configuration/rmarkdown/color_palettes.R")
```

Load custom functions
```{r load-functions}
source('./code/R-functions/semla_functions.r')
# source('./code/R-functions/dge_wrappers.r')
source('./code/R-functions/dge_report.r')
source('./code/R-functions/gse_omnibus.r')
source('./code/R-functions/gse_report.r')
source('./code/R-functions/subchunkify.R')
```

Load MSigDB gene sets
```{r load-msigdb}
gmt_files_symbols <- list(
  msigdb.h = paste0(params$msigdb_dir, '/', params$msigdb_vs, '/h.all.', params$msigdb_vs, '.symbols.gmt'),
  msigdb.c2.cp = paste0(params$msigdb_dir, '/', params$msigdb_vs, '/c2.cp.', params$msigdb_vs, '.symbols.gmt')
)
```

Clean files generated in previous runs
```{r clean-files}
rmd_file <- current_input()
if(!is.null(rmd_file)) {
  figures_dir <- file.path('./docs/figure',rmd_file)
  if(dir.exists(figures_dir)) {
    unlink(file.path(figures_dir, "*"))
  }
}
```


## Data processing (copied from analysis/pdac3-visium-bc-reanalysis.Rmd)

### Load initial data

Load CTC barcodes SummarizedExperiment
```{r dw-load-ctc-bc-se}
ctc_met_bc_se <- readRDS(params$ctc_met_bc_se_file)
ctc_met_bc_se <- ctc_met_bc_se[,ctc_met_bc_se$total_bc>0]
assay(ctc_met_bc_se, 'cpm') <- edgeR::cpm(assay(ctc_met_bc_se, 'norm_counts'))
assay(ctc_met_bc_se, 'logcpm') <- edgeR::cpm(assay(ctc_met_bc_se, 'norm_counts'), log = TRUE)

colData(ctc_met_bc_se) <- colData(ctc_met_bc_se) %>% data.frame %>% 
  mutate(
    sample_type_grouped = ifelse(grepl('^ctc_cluster', sample_type), 'ctc_cluster', sample_type),
    sample_type_grouped = ifelse(sample_type == 'ctc_single_pool', 'ctc_single', sample_type_grouped),
    sample_type_grouped = ifelse(sample_type %in% c('liver', 'lung'), 'metastasis', sample_type_grouped),
    use_anatomical_origin = ifelse(anatomical_origin == 'cardiac_puncture', 'blood', anatomical_origin),
    sample_type_origin = paste(sample_type_grouped, use_anatomical_origin, sep = '-'),
    sample_type_origin = gsub('-lung|-liver', '', sample_type_origin)
  ) %>% 
  DataFrame

```

Load human seurat data
```{r dw-load-seurat}
seu <- readRDS(file.path(params$seurat_dir, 'seurat_hs/seurat_merged_semla_geneNMF.rds'))

# Change rownames
rownames(seu@assays[['Spatial']]) <-  gsub('GRCh38-', '',rownames(seu@assays[['Spatial']]))

i <- 'SCT'
rownames(seu@assays[[i]]@counts) <-  gsub('GRCh38-', '', rownames(seu@assays[[i]]@counts))
rownames(seu@assays[[i]]@data) <-  gsub('GRCh38-', '', rownames(seu@assays[[i]]@data))
rownames(seu@assays[[i]]@scale.data) <-  gsub('GRCh38-', '', rownames(seu@assays[[i]]@scale.data))
seu@assays[[i]]@var.features <- gsub('GRCh38-', '', seu@assays[[i]]@var.features)
```

List of individual seurat files with full features
```{r dw-list-seurat}
seurat_files <- list.files(file.path(params$seurat_dir, "seurat"), pattern = "S[0-9]+.*.rds", full.names = TRUE)
```


### Load processed data
```{r dw-load-processed-data}
bc_spot_freq <- readRDS(file.path(params$output_dir, 'seurat/bc_spot_freq.rds'))
bc_mat <- readRDS(file.path(params$seurat_dir, "seurat_bc", "bc_matrix.rds"))
bc_freq <- readRDS(file.path(params$output_dir, 'seurat/bc_freq.rds'))
seu_full_list <- readRDS(file.path(params$seurat_dir, "seurat", 'seurat_merged_list.rds'))
seu_bc <- readRDS(file.path(params$output_dir, 'seurat_hs/seurat_bc_hotspots.rds'))

# Load hotspot data
hotspot_bc <- readRDS(file.path(params$output_dir, 'seurat_bc/semla_hotspots.rds'))
caf_hotspots <- readRDS(file.path(params$output_dir, "hotspots_tme", "spots_caf_hotspot_2.rds"))
caf_hotspot_r_dist <- readRDS(file.path(params$output_dir, "hotspots_tme", "caf_hotspot_r_dist.rds"))
lyve1_hotspots <- readRDS(file.path(params$output_dir, "hotspots_tme", "spots_lyve1_hotspot_1.rds"))
lyve1_hotspot_r_dist <- readRDS(file.path(params$output_dir, "hotspots_tme", 'lyve1_hotspot_r_dist.rds'))
pecam1_hotspots <- readRDS(file.path(params$output_dir, "hotspots_tme", "spots_pecam1_hotspot_1.rds")) 
pecam1_hotspot_r_dist <- readRDS(file.path(params$output_dir, "hotspots_tme", 'pecam1_hotspot_r_dist.rds'))
```


Add annotation to Seurat object
```{r dw-add-annotation-seurat}
# Add hotspots distances
seu[[]] %<>% 
  rownames_to_column('barcode') %>% 
  left_join(caf_hotspot_r_dist, by = 'barcode') %>%
  left_join(lyve1_hotspot_r_dist, by = 'barcode') %>%
  left_join(pecam1_hotspot_r_dist, by = 'barcode') %>%
  column_to_rownames('barcode')


# Add CAF spots: Filtering criteria must follow the same approach as in chunk::caf-distance-conf
use_bc <- caf_hotspots %>%
  filter(hotspot != 'singleton') %>%
  filter(hotspot_size >= 3) %>%
  pull(sample_spot)

seu$caf_hotspots <- ifelse(colnames(seu) %in% use_bc, 'yes', NA)

# Add Lyve1 spots: Filtering criteria must follow the same approach as in chunk::caf-distance-conf
use_bc <- lyve1_hotspots %>%
  filter(hotspot != 'singleton') %>%
  filter(hotspot_size >= 3) %>%
  pull(sample_spot)

seu$lyve1_hotspots <- ifelse(colnames(seu) %in% use_bc, 'yes', NA)

# Add Pecam1 spots: Filtering criteria must follow the same approach as in chunk::caf-distance-conf
use_bc <- pecam1_hotspots %>%
  filter(hotspot != 'singleton') %>%
  filter(hotspot_size >= 3) %>%
  pull(sample_spot)

seu$pecam1_hotspots <- ifelse(colnames(seu) %in% use_bc, 'yes', NA)

# Add hotspots defined based on barcodes. We use the same criteria as in chunk::dw-bc-hotspots-to-seurat
st_spots <- hotspot_bc$st_specific %>% 
  filter(hotspot != 'singleton') %>% 
  filter(hotspot_size >= 5) %>%
  pull(sample_spot)
ctc_spots <- hotspot_bc$ctc_top %>% 
  filter(hotspot != 'singleton') %>% 
  filter(hotspot_size >= 5) %>%
  pull(sample_spot)

seu[[]] %<>% 
  rownames_to_column('sample_spot') %>%
  mutate(
    bc_group = ifelse(sample_spot %in% st_spots, 'ST', ifelse(sample_spot %in% ctc_spots, 'CTC', 'other'))
  ) %>% 
  column_to_rownames('sample_spot')

```

Create a Seurat object combining CAF hotspots and BC group annotation. We will use this Seurat object for the spatial distribution analysis.
```{r dw-seurat-bc-caf}
# Seurat object combining CAF and BC group annotation
seu_bc_caf <- seu

# Add bc_group_cluster information
bc_group_cluster_spots <- seu_bc[[]] %>% rownames_to_column('sample_spot') %>% 
  select(sample_spot, bc_group_cluster, clusters)

# Add Acta2 expression
# Rename groups
seu_bc_caf[[]] %<>% 
  rownames_to_column('sample_spot') %>% 
  left_join(bc_group_cluster_spots) %>% 
  mutate(
    group_hotspots = paste(bc_group, caf_hotspots, sep = '_'),
    group_hotspots = gsub("_NA", "", group_hotspots),
    group_hotspots = gsub("NA_", "", group_hotspots),
    group_hotspots = gsub("yes", "CAF", group_hotspots),
    group_hotspots = gsub("other_CAF", "CAF", group_hotspots),
    
    group_hotspots_extended = paste(bc_group_cluster, caf_hotspots, sep = '_'),
    group_hotspots_extended = gsub("_NA", "", group_hotspots_extended),
    group_hotspots_extended = gsub("NA_", "", group_hotspots_extended),
    group_hotspots_extended = gsub("yes", "CAF", group_hotspots_extended),
    group_hotspots_extended = ifelse(group_hotspots_extended == 'NA', 'other', group_hotspots_extended)
  ) %>% 
  column_to_rownames('sample_spot')

# Remove spots with no annotation
seu_bc_caf <- seu_bc_caf[,seu_bc_caf$group_hotspots !='other']
```





## Acta2 proximity

Load Acta2 expression
```{r pcam1-proximity}
acta2_exprs <- readRDS(file.path(params$output_dir, 'caf/acta2_expression.rds'))
```

Showing radial distance from CTC groups and ST spots to nearest Acta2 hotspots
```{r bc_group_clusters-acta2-proximity-conf}
use_data <- seu_bc[[]] %>% 
  filter(!is.na(bc_group_cluster))  %>%
  rownames_to_column('barcode') %>%
  left_join(acta2_exprs) %>% 
  mutate(acta2_exprs = acta2 >= 2)
```


### Summary statistics
```{r bc_group_clusters-acta2-proximity-table}
use_data %>% 
  group_by(bc_group_cluster) %>% 
  summarise(
    median_dist_2d = median(r_dist_caf_hotspot_2d, na.rm = TRUE)%>% round(2),
    median_dist_3d = median(r_dist_caf_hotspot_3d_transformed, na.rm = TRUE)%>% round(2),
    mean_dist_2d = mean(r_dist_caf_hotspot_2d, na.rm = TRUE)%>% round(2),
    sd_dist_2d = sd(r_dist_caf_hotspot_2d, na.rm = TRUE)%>% round(2),
    mean_dist_3d = mean(r_dist_caf_hotspot_3d_transformed, na.rm = TRUE)%>% round(2),
    sd_dist_3d = sd(r_dist_caf_hotspot_3d_transformed, na.rm = TRUE) %>% round(2),
    acta2_exprs_prop = round(sum(acta2_exprs, na.rm = TRUE) / n(), 2),
  ) %>% 
  kbl(caption = 'Summary statistics for CAF distance') %>%
  kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```



### Co-localization

#### Barplot
Percentage of spots with Acta2 expression (UMI counts >= 1) in each subgroup
```{r bc_group_clusters-acta2-colocalization, fig.width = 0.984, fig.asp = 1}
use_data %>% 
  group_by(bc_group_cluster) %>%
  summarise(
    acta2_exprs_prop = round(sum(acta2_exprs, na.rm = TRUE) / n(), 2)
  ) %>%
  ggplot(aes(bc_group_cluster, acta2_exprs_prop, fill = bc_group_cluster)) +
  geom_col()+
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)),
                     labels = scales::percent) +
  scale_fill_manual(values = bc_group_cluster_palette) +
  labs(
    x = '',
      y = 'Spots %',
    fill = ''
  ) +
  guides(fill = 'none') +
  theme_facet  +
  theme(
    # plot.margin = unit(c(-0.1, 0.01, -0.05, 0.01), "inches"),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank()
  )
```

#### Expression distribution
```{r bc_group_clusters-acta2-coloc-score, fig.width= 1.2, fig.asp = 0.738}

median_stats <- use_data %>% 
  group_by(bc_group_cluster) %>%
  summarise(
    median_expression = median(acta2, na.rm = TRUE),
    iqr_expression = iqr(acta2, na.rm = TRUE),
    spots_n = sum(acta2 >= 1, na.rm = TRUE),
    spots_prop = sum(acta2 >= 1, na.rm = TRUE) / n(),
  )

stats_table <- use_data %>% 
  wilcox_test(acta2 ~ bc_group_cluster) %>% 
  left_join(median_stats, by = c('group1' = 'bc_group_cluster'))

median_stats %>% 
  kbl(caption = 'Summary statistics for Acta2 expression') %>%
  kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)

stats_table %>% 
  kbl(caption = 'Summary statistics for Acta2 expression') %>%
  kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```


### Distance: Volin plot
```{r bc_group_clusters-acta2-proximity-violin, fig.width= 1.2, fig.asp = 0.738}
dist_var <- 'r_dist_caf_hotspot_3d_transformed'
use_data_plot <- use_data %>% 
  mutate(
    r_dist = ifelse(.[[dist_var]] < 0, 0, .[[dist_var]])
    )

# Calculate automatic upper limit to crop values (not applied at the moment)
upper_limit_box <- use_data %>%  
  mutate(
    r_dist = ifelse(.[[dist_var]] < 0, 0, .[[dist_var]])
    ) %>% 
  group_by(bc_group_cluster) %>% 
  summarise(
    q3 = quantile(r_dist, .75),
    iqr = iqr(r_dist)
    # upper_limit = 1.5*IQR + q3
  ) %>% 
  mutate(
    upper_limit_whisker = 1.5*iqr + q3,
    upper_limit_cropped = 0.75*iqr + q3
  )
cropped_val <- upper_limit_box$upper_limit_cropped %>% max()


# Define a fixed value to crop distance values
cropped_val <- 3
y_axis_breaks <- scales::breaks_extended(n = 5)(c(0, cropped_val))
y_axis_labels <- as.character(y_axis_breaks)
y_axis_labels[length(y_axis_labels)] <- paste0('≥', y_axis_labels[length(y_axis_labels)])

# Comapre means to add p-values to the plot. We will use Wilcoxon test, which is a non-parametric test that does not assume normal distribution of the data. We use manual calculation instea of adding stat_compare_means to the plot because we are using cropped vales for plotting
stat.test <- compare_means(
  r_dist ~ bc_group_cluster, 
  data = use_data_plot,
  method = "wilcox.test"
) %>% 
  mutate(
    y.position = cropped_val + cropped_val*0.1
  )

use_data_plot %>%
  mutate(
    r_dist = r_dist / 1000,
    r_dist = ifelse(r_dist >= cropped_val, cropped_val, r_dist)
  ) %>%
  ggplot(aes(bc_group_cluster, r_dist,fill = bc_group_cluster)) +
  geom_violin(scale = 1, rel_min_height = 0.01, alpha = 0.7, linewidth = linewidth_default/2) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA, linewidth = linewidth_default/2) +
  scale_fill_manual(values = bc_group_cluster_palette) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.25)),
                       breaks = y_axis_breaks,
                       labels = y_axis_labels
    ) +
  stat_pvalue_manual(
    stat.test, 
    size = geom_text_size * 0.8,
    label = "p.signif", 
    tip.length = 0.01, 
    step.increase = 0.2,
    vjust = 0.5,
    bracket.size = linewidth_default/2
  ) +
    labs(
      title = '',
      x = '',
      y = 'Distance (mm)'#,
      # caption = paste0('Distance cropped at ',cropped_val)
    ) +
    theme(
      legend.position = "none",
      plot.margin = unit(c(-0.1, 0.01, -0.05, 0.01), "inches"),
      axis.ticks.x = element_blank(),
      axis.text.x = element_blank()
    )
```


## Lyve1 proximity

Load Lyve1 expression
```{r lyve1-proximity}
lyve1_exprs <- readRDS(file.path(params$output_dir, 'hotspots_tme/lyve1_expression.rds'))
```

Showing radial distance from CTC groups and ST spots to nearest Lyve1 hotspots
```{r bc_group_clusters-lyve1-proximity-conf}
use_data <- seu_bc[[]] %>% 
  filter(!is.na(bc_group_cluster))  %>%
  rownames_to_column('barcode') %>%
  left_join(lyve1_exprs) %>% 
  mutate(lyve1_exprs = lyve1 >= 1)
```


### Summary statistics
```{r bc_group_clusters-lyve1-proximity-table}
use_data %>% 
  group_by(bc_group_cluster) %>% 
  summarise(
    median_dist_2d = median(r_dist_lyve1_hotspot_2d, na.rm = TRUE)%>% round(2),
    median_dist_3d = median(r_dist_lyve1_hotspot_3d_transformed, na.rm = TRUE)%>% round(2),
    mean_dist_2d = mean(r_dist_lyve1_hotspot_2d, na.rm = TRUE)%>% round(2),
    sd_dist_2d = sd(r_dist_lyve1_hotspot_2d, na.rm = TRUE)%>% round(2),
    mean_dist_3d = mean(r_dist_lyve1_hotspot_3d_transformed, na.rm = TRUE)%>% round(2),
    sd_dist_3d = sd(r_dist_lyve1_hotspot_3d_transformed, na.rm = TRUE) %>% round(2)
  ) %>% 
  kbl(caption = 'Summary statistics for Lyve1 distance') %>%
  kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```



### Co-localization

#### Barplot
Percentage of spots with Lyve1 expression (UMI counts >= 1) in each subgroup
```{r bc_group_clusters-lyve1-colocalization, fig.width = 0.984, fig.asp = 1}
use_data %>% 
  group_by(bc_group_cluster) %>%
  summarise(
    lyve1_exprs_prop = round(sum(lyve1_exprs, na.rm = TRUE) / n(), 2)
  ) %>%
  ggplot(aes(bc_group_cluster, lyve1_exprs_prop, fill = bc_group_cluster)) +
  geom_col()+
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)),
                     labels = scales::percent) +
  scale_fill_manual(values = bc_group_cluster_palette) +
  labs(
    x = '',
      y = 'Spots %',
    fill = ''
  ) +
  guides(fill = 'none') +
  theme_facet  +
  theme(
    # plot.margin = unit(c(-0.1, 0.01, -0.05, 0.01), "inches"),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank()
  )
```

#### Expression distribution
```{r bc_group_clusters-lyve1-coloc-score, fig.width= 1.2, fig.asp = 0.738}

median_stats <- use_data %>% 
  group_by(bc_group_cluster) %>%
  summarise(
    median_expression = median(lyve1, na.rm = TRUE),
    iqr_expression = iqr(lyve1, na.rm = TRUE),
    spots_n = sum(lyve1 >= 1, na.rm = TRUE),
    spots_prop = sum(lyve1 >= 1, na.rm = TRUE) / n(),
  )

stats_table <- use_data %>% 
  wilcox_test(lyve1 ~ bc_group_cluster) %>% 
  left_join(median_stats, by = c('group1' = 'bc_group_cluster'))

median_stats %>% 
  kbl(caption = 'Summary statistics for Lyve1 expression') %>%
  kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)

stats_table %>% 
  kbl(caption = 'Summary statistics for Lyve1 expression') %>%
  kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```


### Distance: Volin plot
```{r bc_group_clusters-lyve1-proximity-violin, fig.width= 1.2, fig.asp = 0.738}
dist_var <- 'r_dist_lyve1_hotspot_3d_transformed'
use_data_plot <- use_data %>% 
  mutate(
    r_dist = ifelse(.[[dist_var]] < 0, 0, .[[dist_var]])
    )

# Calculate automatic upper limit to crop values (not applied at the moment)
upper_limit_box <- use_data %>%  
  mutate(
    r_dist = ifelse(.[[dist_var]] < 0, 0, .[[dist_var]])
    ) %>% 
  group_by(bc_group_cluster) %>% 
  summarise(
    q3 = quantile(r_dist, .75),
    iqr = iqr(r_dist)
    # upper_limit = 1.5*IQR + q3
  ) %>% 
  mutate(
    upper_limit_whisker = 1.5*iqr + q3,
    upper_limit_cropped = 0.75*iqr + q3
  )
cropped_val <- upper_limit_box$upper_limit_cropped %>% max()


# Define a fixed value to crop distance values
cropped_val <- 3
y_axis_breaks <- scales::breaks_extended(n = 5)(c(0, cropped_val))
y_axis_labels <- as.character(y_axis_breaks)
y_axis_labels[length(y_axis_labels)] <- paste0('≥', y_axis_labels[length(y_axis_labels)])

# Comapre means to add p-values to the plot. We will use Wilcoxon test, which is a non-parametric test that does not assume normal distribution of the data. We use manual calculation instea of adding stat_compare_means to the plot because we are using cropped vales for plotting
stat.test <- compare_means(
  r_dist ~ bc_group_cluster, 
  data = use_data_plot,
  method = "wilcox.test"
) %>% 
  mutate(
    y.position = cropped_val + cropped_val*0.1
  )

use_data_plot %>%
  mutate(
    r_dist = r_dist / 1000,
    r_dist = ifelse(r_dist >= cropped_val, cropped_val, r_dist)
  ) %>%
  ggplot(aes(bc_group_cluster, r_dist,fill = bc_group_cluster)) +
  geom_violin(scale = 1, rel_min_height = 0.01, alpha = 0.7, linewidth = linewidth_default/2) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA, linewidth = linewidth_default/2) +
  scale_fill_manual(values = bc_group_cluster_palette) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.25)),
                       breaks = y_axis_breaks,
                       labels = y_axis_labels
    ) +
  stat_pvalue_manual(
    stat.test, 
    size = geom_text_size * 0.8,
    label = "p.signif", 
    tip.length = 0.01, 
    step.increase = 0.2,
    vjust = 0.5,
    bracket.size = linewidth_default/2
  ) +
    labs(
      title = '',
      x = '',
      y = 'Distance (mm)'#,
      # caption = paste0('Distance cropped at ',cropped_val)
    ) +
    theme(
      legend.position = "none",
      plot.margin = unit(c(-0.1, 0.01, -0.05, 0.01), "inches"),
      axis.ticks.x = element_blank(),
      axis.text.x = element_blank()
    )
```


## Pecam1 proximity

Load Pecam1 expression
```{r pecam1-proximity}
pecam1_exprs <- readRDS(file.path(params$output_dir, 'hotspots_tme/pecam1_expression.rds'))
```

Showing radial distance from CTC groups and ST spots to nearest Pecam1 hotspots
```{r bc_group_clusters-pecam1-proximity-conf}
use_data <- seu_bc[[]] %>% 
  filter(!is.na(bc_group_cluster))  %>%
  rownames_to_column('barcode') %>%
  left_join(pecam1_exprs) %>% 
  mutate(pecam1_exprs = pecam1 >= 1)
```


### Summary statistics
```{r bc_group_clusters-pecam1-proximity-table}
use_data %>% 
  group_by(bc_group_cluster) %>% 
  summarise(
    median_dist_2d = median(r_dist_pecam1_hotspot_2d, na.rm = TRUE)%>% round(2),
    median_dist_3d = median(r_dist_pecam1_hotspot_3d_transformed, na.rm = TRUE)%>% round(2),
    mean_dist_2d = mean(r_dist_pecam1_hotspot_2d, na.rm = TRUE)%>% round(2),
    sd_dist_2d = sd(r_dist_pecam1_hotspot_2d, na.rm = TRUE)%>% round(2),
    mean_dist_3d = mean(r_dist_pecam1_hotspot_3d_transformed, na.rm = TRUE)%>% round(2),
    sd_dist_3d = sd(r_dist_pecam1_hotspot_3d_transformed, na.rm = TRUE) %>% round(2)
  ) %>% 
  kbl(caption = 'Summary statistics for Pecam1 distance') %>%
  kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```



### Co-localization

#### Barplot
Percentage of spots with Pecam1 expression (UMI counts >= 1) in each subgroup
```{r bc_group_clusters-pecam1-colocalization, fig.width = 0.984, fig.asp = 1}
use_data %>% 
  group_by(bc_group_cluster) %>%
  summarise(
    pecam1_exprs_prop = round(sum(pecam1_exprs, na.rm = TRUE) / n(), 2)
  ) %>%
  ggplot(aes(bc_group_cluster, pecam1_exprs_prop, fill = bc_group_cluster)) +
  geom_col()+
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)),
                     labels = scales::percent) +
  scale_fill_manual(values = bc_group_cluster_palette) +
  labs(
    x = '',
      y = 'Spots %',
    fill = ''
  ) +
  guides(fill = 'none') +
  theme_facet  +
  theme(
    # plot.margin = unit(c(-0.1, 0.01, -0.05, 0.01), "inches"),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank()
  )
```

#### Expression distribution
```{r bc_group_clusters-pecam1-coloc-score, fig.width= 1.2, fig.asp = 0.738}

median_stats <- use_data %>% 
  group_by(bc_group_cluster) %>%
  summarise(
    median_expression = median(pecam1, na.rm = TRUE),
    iqr_expression = iqr(pecam1, na.rm = TRUE),
    spots_n = sum(pecam1 >= 1, na.rm = TRUE),
    spots_prop = sum(pecam1 >= 1, na.rm = TRUE) / n(),
  )

stats_table <- use_data %>% 
  wilcox_test(pecam1 ~ bc_group_cluster) %>% 
  left_join(median_stats, by = c('group1' = 'bc_group_cluster'))

median_stats %>% 
  kbl(caption = 'Summary statistics for Pecam1 expression') %>%
  kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)

stats_table %>% 
  kbl(caption = 'Summary statistics for Pecam1 expression') %>%
  kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```


### Distance: Volin plot
```{r bc_group_clusters-pecam1-proximity-violin, fig.width= 1.2, fig.asp = 0.738}
dist_var <- 'r_dist_pecam1_hotspot_3d_transformed'
use_data_plot <- use_data %>% 
  mutate(
    r_dist = ifelse(.[[dist_var]] < 0, 0, .[[dist_var]])
    )

# Calculate automatic upper limit to crop values (not applied at the moment)
upper_limit_box <- use_data %>%  
  mutate(
    r_dist = ifelse(.[[dist_var]] < 0, 0, .[[dist_var]])
    ) %>% 
  group_by(bc_group_cluster) %>% 
  summarise(
    q3 = quantile(r_dist, .75),
    iqr = iqr(r_dist)
    # upper_limit = 1.5*IQR + q3
  ) %>% 
  mutate(
    upper_limit_whisker = 1.5*iqr + q3,
    upper_limit_cropped = 0.75*iqr + q3
  )
cropped_val <- upper_limit_box$upper_limit_cropped %>% max()


# Define a fixed value to crop distance values
cropped_val <- 3
y_axis_breaks <- scales::breaks_extended(n = 5)(c(0, cropped_val))
y_axis_labels <- as.character(y_axis_breaks)
y_axis_labels[length(y_axis_labels)] <- paste0('≥', y_axis_labels[length(y_axis_labels)])

# Comapre means to add p-values to the plot. We will use Wilcoxon test, which is a non-parametric test that does not assume normal distribution of the data. We use manual calculation instea of adding stat_compare_means to the plot because we are using cropped vales for plotting
stat.test <- compare_means(
  r_dist ~ bc_group_cluster, 
  data = use_data_plot,
  method = "wilcox.test"
) %>% 
  mutate(
    y.position = cropped_val + cropped_val*0.1
  )

use_data_plot %>%
  mutate(
    r_dist = r_dist / 1000,
    r_dist = ifelse(r_dist >= cropped_val, cropped_val, r_dist)
  ) %>%
  ggplot(aes(bc_group_cluster, r_dist,fill = bc_group_cluster)) +
  geom_violin(scale = 1, rel_min_height = 0.01, alpha = 0.7, linewidth = linewidth_default/2) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA, linewidth = linewidth_default/2) +
  scale_fill_manual(values = bc_group_cluster_palette) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.25)),
                       breaks = y_axis_breaks,
                       labels = y_axis_labels
    ) +
  stat_pvalue_manual(
    stat.test, 
    size = geom_text_size * 0.8,
    label = "p.signif", 
    tip.length = 0.01, 
    step.increase = 0.2,
    vjust = 0.5,
    bracket.size = linewidth_default/2
  ) +
    labs(
      title = '',
      x = '',
      y = 'Distance (mm)'#,
      # caption = paste0('Distance cropped at ',cropped_val)
    ) +
    theme(
      legend.position = "none",
      plot.margin = unit(c(-0.1, 0.01, -0.05, 0.01), "inches"),
      axis.ticks.x = element_blank(),
      axis.text.x = element_blank()
    )
```


## Macropahge and Monocyte weight from RCTD
```{r deconv-proximity-conf}
rctd_dist <- readRDS(file.path(params$output_dir, 'seurat_mm', 'rctd_multi_df_tabmuris_r_dist.rds'))
rctd_multi_df <- readRDS(file.path(params$output_dir, 'seurat_mm', 'rctd_multi_df_tabmuris_filtered.rds'))
bc_group_dist <- readRDS(file.path(params$output_dir, 'seurat_hs', 'bc_group_cluster_r_dist.rds'))
```

### Score distribution
```{r myeloid-colocalization-conf}
use_weight_df <- rctd_multi_df %>% 
  filter(confidence) %>%
  select(barcode, cell_type_grouped, weight) %>% 
  group_by(barcode, cell_type_grouped) %>%
  summarise(weight = max(weight)) %>% 
  filter(cell_type_grouped %in% c('macrophage', 'monocyte')) %>% 
  pivot_wider(names_from = cell_type_grouped, values_from = weight)

use_data <- seu_bc[[]] %>% 
  rownames_to_column('barcode') %>% 
  left_join(use_weight_df , by = 'barcode') %>% 
  mutate(
    macrophage = ifelse(is.na(macrophage), FALSE, macrophage),
    monocyte = ifelse(is.na(monocyte), FALSE, monocyte)
  ) %>% 
  filter(!is.na(bc_group_cluster))
```

#### Table
```{r myeloid-colocalization-table}
use_score_vars <- c('monocyte', 'macrophage')

use_formula <- as.formula(paste('bc_group_cluster ~', paste0("`", use_score_vars, "`", collapse = ' + ')))
tab1 <- tableby(use_formula, data=use_data,
                control = tableby.control(test = TRUE, total = FALSE))
summary(tab1, pfootnote=TRUE, title='Distribution of Monocyte and Macrophage deconvolution weights in BC group spots')

```

#### Macrophage score
```{r macrophage-colocalization-dist, fig.width = 0.984, fig.asp = 1}
use_comparisons <- combn(levels(seu_bc$bc_group_cluster), 2, simplify = FALSE)
use_data %>% 
    ggplot(aes(bc_group_cluster, macrophage, fill = bc_group_cluster)) +
    geom_violin(scale = 'width', alpha = 0.7, 
                linewidth = linewidth_default/2, 
                trim = FALSE,
                adjust = 2
    ) +
    geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA, linewidth = linewidth_default/2) +
    scale_fill_manual(values = bc_group_cluster_palette) +
  stat_compare_means(
    comparisons = use_comparisons,
    bracket.size = linewidth_default/2,
    size = geom_text_size * 0.8
    )+
    labs(
      title = '',
      x = '',
      y = 'Macrophage score'#,
      # caption = paste0('Distance cropped at ',cropped_val)
    ) +
    theme(
      legend.position = "none",
      plot.margin = unit(c(0, 0.01, -0.05, 0.01), "inches"),
      axis.ticks.x = element_blank(),
      axis.text.x = element_blank()
    )
  
```

#### Monocyte score
```{r monocyte-colocalization-dist, fig.width = 0.984, fig.asp = 1}
use_comparisons <- combn(levels(seu_bc$bc_group_cluster), 2, simplify = FALSE)
use_data %>% 
    ggplot(aes(bc_group_cluster, monocyte, fill = bc_group_cluster)) +
    geom_violin(scale = 'width', alpha = 0.7, 
                linewidth = linewidth_default/2, 
                trim = FALSE,
                adjust = 2
    ) +
    geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA, linewidth = linewidth_default/2) +
    scale_fill_manual(values = bc_group_cluster_palette) +
  stat_compare_means(
    comparisons = use_comparisons,
    bracket.size = linewidth_default/2,
    size = geom_text_size * 0.8
    )+
    labs(
      title = '',
      x = '',
      y = 'Monocyte score'#,
      # caption = paste0('Distance cropped at ',cropped_val)
    ) +
    theme(
      legend.position = "none",
      plot.margin = unit(c(0, 0.01, -0.05, 0.01), "inches"),
      axis.ticks.x = element_blank(),
      axis.text.x = element_blank()
    )
  
```



## CAF subtypes proximity
```{r caf_sub-proximity-conf}
caf_mk <- readRDS(file.path(params$output_dir, "caf", 'caf_subtypes_markers.rds'))
fb_subtypes_data <- readRDS(file.path(params$output_dir, "caf", 'ucell_fibroblast_subtypes.rds'))
fb_subtypes_dist <- readRDS(file.path(params$output_dir, 'caf/caf_subtypes_r_dist.rds'))
bc_group_dist <- readRDS(file.path(params$output_dir, 'seurat_hs', 'bc_group_cluster_r_dist.rds'))
```

Showing radial distance from CTC and ST spots to nearest cell type hotspots

```{r bc_group_clusters-caf_sub-proximity-conf}
use_data <- seu_bc[[]] %>% 
  filter(!is.na(bc_group_cluster)) %>% 
  rownames_to_column('barcode') %>%
  left_join(fb_subtypes_dist, by = 'barcode')
rdist_vars_2d <- grep('r_dist_cafst_2d' , colnames(use_data), value = TRUE)
rdist_vars_3d <- grep('r_dist_cafst_3d' , colnames(use_data), value = TRUE)
```

### Co-localization

We define co-localization as spots at a radial distance ≤ 0 microns from a cell type hotspot. This means that the spot is either within the hotspot or exactly at the boundary of the hotspot.

```{r bc_group_clusters-caf_sub-proximity-coloc-table}
use_data_co <- use_data %>% 
  dplyr::select(barcode, bc_group_cluster, all_of(rdist_vars_2d)) %>% 
  mutate(across(all_of(rdist_vars_2d), ~ .x  <= 0))

use_labels <- gsub('r_dist_cafst_2d_', '', rdist_vars_2d) %>% 
  set_names(rdist_vars_2d)
labels(use_data_co)  <- gsub('r_dist_cafst_2d_', '', rdist_vars_2d) %>% 
  set_names(rdist_vars_2d)

use_formula <- as.formula(paste('bc_group_cluster ~', paste0("`", rdist_vars_2d, "`", collapse = ' + ')))
tab1 <- tableby(use_formula, data=use_data_co,
                control = tableby.control(test = TRUE, total = FALSE))
summary(tab1, pfootnote=TRUE, title='Co-localization of cell types')


tab1 <- tableby(use_formula, data=use_data_co, cat.simplify = "label",
                control = tableby.control(test = TRUE, total = FALSE, cat.stats = c("countpct")))
summary(tab1, pfootnote=TRUE, title='Co-localization of cell types (simplified)')

```



#### Signature scores

```{r bc_group_clusters-caf_sub-proximity-coloc-score, fig.width= 1.2, fig.asp = 0.738}
var_ucell <- paste0(names(caf_mk$mm), '_mm_ucell') %>% grep('^FB', ., value = TRUE, invert = TRUE)

use_data_plot <- use_data %>% 
  left_join(fb_subtypes_data %>% select(barcode, all_of(var_ucell)), by = 'barcode') %>%
  dplyr::select(barcode, bc_group_cluster, all_of(var_ucell)) %>% 
  pivot_longer(cols = all_of(var_ucell), names_to = 'cell_type', values_to = 'score') %>% 
  mutate(cell_type = gsub('_mm_ucell', '', cell_type))

use_comparisons <- combn(levels(use_data_plot$bc_group_cluster), 2, simplify = FALSE)


median_stats <- use_data_plot %>% 
  group_by(cell_type, bc_group_cluster) %>%
  summarise(
    median_score = median(score),
    iqr_score = iqr(score)
  )

stats_table <- use_data_plot %>% 
  group_by(cell_type) %>% 
  wilcox_test(score ~ bc_group_cluster) %>% 
  left_join(median_stats, by = c('cell_type', 'group1' = 'bc_group_cluster')) %>% 
  dplyr::rename(
    group1_median_score = median_score,
    group1_iqr_score = iqr_score
  ) %>% 
  left_join(median_stats, by = c('cell_type', 'group2' = 'bc_group_cluster')) %>% 
  dplyr::rename(
    group2_median_score = median_score,
    group2_iqr_score = iqr_score
  ) %>% 
  mutate(
    median_diff = group1_median_score - group2_median_score
  )

stats_table %>% 
  kbl(caption = 'Summary statistics for CAF distance') %>%
  kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
  
ct <- 'myCAF'
for(ct in unique(use_data_plot$cell_type)) {
  
  cat("#####", ct, "\n\n")
  res <- use_data_plot %>% 
    filter(cell_type == ct) %>% 
    mutate(
      score = ifelse(score >= 0.5, 0.5, score)
    ) %>%
    ggplot(aes(bc_group_cluster, score, fill = bc_group_cluster)) +
    geom_violin(scale = 'width', alpha = 0.7, 
                linewidth = linewidth_default/2, 
                trim = FALSE,
                adjust = 2
    ) +
    geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA, linewidth = linewidth_default/2) +
    scale_fill_manual(values = bc_group_cluster_palette) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.25))) +
    stat_compare_means(comparisons = use_comparisons, 
                       size = geom_text_size * 0.8,
                       label = "p.signif",
                       bracket.size = linewidth_default/2,
                       step.increase = 0.3,
                       vjust = 0.5) +
    # facet_wrap(~cell_type, scales = 'free_y') +
    labs(
      x = '',
      y = 'Score',
      title = ''
    ) +
    theme(
      legend.position = "none",
      plot.margin = unit(c(-0.1, 0.01, -0.05, 0.01), "inches"),
      axis.ticks.x = element_blank(),
      axis.text.x = element_blank()
      ) #+
    # theme_facet
  
  print(res)
  cat("\n\n")
}

```


### Score - Distances curves

#### Individual plot
```{r bc_group_clusters-caf_sub-score-distance-individual, fig.width= 1.2, fig.asp = 0.738}
use_score_vars <- paste0(caf_mk$mm %>% names, '_mm_ucell') %>% grep('^FB', ., value = TRUE, invert = TRUE)

use_data <- fb_subtypes_data %>% 
  left_join(bc_group_dist) %>% 
  pivot_longer(cols = starts_with('r_dist_'), 
               names_to = 'bc_group_cluster', 
               values_to = 'r_dist') %>% 
  mutate(
    bc_group_cluster = gsub('r_dist_', '', bc_group_cluster),
    bc_group_cluster = factor(bc_group_cluster, levels = c('ST', 'CTC1', 'CTC2'))
  ) %>% 
  dplyr::select(barcode, bc_group_cluster, r_dist, all_of(use_score_vars)) %>% 
  pivot_longer(cols = all_of(use_score_vars), names_to = 'cell_type', values_to = 'score') %>%
  mutate(cell_type = gsub('_mm_ucell', '', cell_type))


ct <- 'myCAF'
for(ct in unique(use_data$cell_type)) {
  
  cat("#####", ct, "\n\n")
  res <- use_data %>% 
    filter(cell_type == ct) %>% 
    filter(r_dist < params$max_dist) %>% 
    mutate(
      r_dist = ifelse(r_dist < 0, 0, r_dist)
    ) %>%
    ggplot(aes(r_dist/1000, score, color = bc_group_cluster)) +
    geom_smooth(method = "gam", 
                formula = y ~ s(x, bs = "cs"),
                se = FALSE,
                linewidth = linewidth_default*2) +
    scale_color_manual(values = bc_group_cluster_palette) +
    scale_x_continuous(
      expand = expansion(mult = c(0, 0.15)),
      limits = c(0, params$crop_dist/1000)
    ) +
    labs(
      title = '',
      x = 'Distance (mm)',
      y = 'Score',
      color = ''
    )  +
    theme(
      legend.position = "none",
      plot.margin = unit(c(-0.1, 0.01, 0, 0.01), "inches")
      ) 
  
  print(res)
  cat("\n\n")
}

```


#### Statistical test

Use Generalized Additive Models (GAMs) to evaluate relationship between expression and distance across the groups. We used Cubic Regression Splines with shrinkage to allow for automated smoothing parameter selection and to penalize non-significant wiggles toward a zero-effect null space. All models were fitted using Restricted Maximum Likelihood (REML).

```{r bc_group_clusters-caf_sub-score-distance-test}
use_data <- fb_subtypes_data %>% 
  left_join(bc_group_dist) %>% 
  pivot_longer(cols = starts_with('r_dist_'), 
               names_to = 'bc_group_cluster', 
               values_to = 'r_dist') %>% 
  mutate(
    bc_group_cluster = gsub('r_dist_', '', bc_group_cluster),
    bc_group_cluster = factor(bc_group_cluster, levels = c('ST', 'CTC1', 'CTC2'))
  ) %>% 
  dplyr::select(barcode, bc_group_cluster, r_dist, all_of(use_score_vars)) %>% 
  pivot_longer(cols = all_of(use_score_vars), names_to = 'cell_type', values_to = 'score') %>%
  mutate(cell_type = gsub('_mm_ucell', '', cell_type)) %>% 
  filter(r_dist < params$max_dist) %>% 
  mutate(r_dist = ifelse(r_dist < 0, 0, r_dist), 
         r_dist = r_dist / 1000) 


# Global assessment, comparing null and full model
global_test_combined <- foreach(ct = unique(use_data$cell_type), .combine = rbind) %do% {
  use_data_test <- use_data %>% 
    filter(cell_type == ct)

  m_null <- gam(score ~ s(r_dist, bs = "cs"), data = use_data_test, method = "REML")
  m_full <- gam(score ~ bc_group_cluster + s(r_dist, by = bc_group_cluster, bs = "cs"), 
                data = use_data_test, method = "REML")
  global_test <- anova(m_null, m_full, test = "F") %>% 
    tidy
  global_test[2,] %>% 
    mutate(term = ct)
  
}
cat("Global Comparison: The overall effect of Condition was assessed by comparing a full model (allowing condition-specific smooths) against a null model (a single smooth for all data) using an Analysis of Deviance (F-test):\n\n")
global_test_combined %>% 
  mutate(
    p.value = format.pval(p.value, 2)
    ) %>% 
  kbl(caption = 'Global test for differences in expression-distance curves') %>%
  kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)


# Specific differences
data_ordered <- use_data
# We convert Condition to an "ordered" factor. 
# This tells mgcv to treat the first level as a reference and others as differences.
data_ordered$bc_group_cluster <- as.ordered(data_ordered$bc_group_cluster)
contrasts(data_ordered$bc_group_cluster) <- "contr.treatment" # Use treatment contrasts
gam_specific_combined <- foreach(ct = unique(data_ordered$cell_type)) %do% {
  use_data_test <- data_ordered %>% 
    filter(cell_type == ct)
  gam(score ~ bc_group_cluster + s(r_dist, bs = "cs") + 
        s(r_dist, by = bc_group_cluster, bs = "cs"), 
      data = use_data_test, method = "REML")
}
names(gam_specific_combined) <- unique(data_ordered$cell_type)

# Extract the parametric table (Intercepts and Magnitude differences)
cat("Pairwise Comparison: Parametric coefficients. This table shows the differences in magnitude (the average vertical position of your curves). \n\n")
parametric_table <- foreach(i = names(gam_specific_combined), .combine = rbind) %do% {
  gam_specific_combined[[i]] %>% 
    tidy(parametric = TRUE) %>% 
    mutate(
      term = c('Intercept', levels(data_ordered$bc_group_cluster)[-1]),
      p.value = format.pval(p.value, 2)
    ) %>%
    filter(term != 'Intercept') %>%
    mutate(cell_type = i)
}

parametric_table %>% 
  dplyr::select(cell_type, everything())  %>% 
  kbl(caption = 'Parametric (magnitude) differences by group, using as reference ST') %>%
  kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)


# Extract the Smooth table (Shape/Slope differences)
cat("Pairwise Comparison: Smooth terms. This table shows the differences in shape of the relationship between expression and distance. \n\n")
smooth_table <- foreach(i = names(gam_specific_combined), .combine = rbind) %do% {
  gam_specific_combined[[i]] %>% 
    tidy(parametric = FALSE) %>% 
    mutate(
      term = c('Intercept', levels(data_ordered$bc_group_cluster)[-1]),
      p.value = format.pval(p.value, 2)
    ) %>%
    filter(term != 'Intercept') %>%
    mutate(cell_type = i)
}
smooth_table %>% 
  dplyr::select(cell_type, everything()) %>%
  kbl(caption = 'Smooth (shape) differences by group, using as reference ST') %>%
  kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```








## Macrophage subtypes

Analysis of gene-signature scores for macrophage subtypes derived from Khaliq (2024).
```{r mc-conf}
khaliq_signatures <- readRDS(file.path('./output/rnaseq-spatial/pdac3/seurat_mm', 'ucell_khaliq2024.rds')) %>% 
  dplyr::rename(barcode = sample_barcode)

mc_signatures <- khaliq_signatures %>% 
  dplyr::select(barcode, `M1 signature_UCell`, `Tumor-associated Macrophages_UCell`) %>% 
  dplyr::rename(M1_UCell = `M1 signature_UCell`, TAM_UCell = `Tumor-associated Macrophages_UCell`) %>% 
  mutate(
    m1_score_khaliq = log2((M1_UCell + 0.01)/ (TAM_UCell + 0.01)),
    tam_score_khaliq = log2((TAM_UCell + 0.01)/ ( M1_UCell+ 0.01)),
    m_subtype_khaliq = case_when(
      m1_score_khaliq >= 3 ~ 'M1',
      m1_score_khaliq <= -3 ~ 'TAM',
      TRUE ~ 'Intermediate/Mixed'
    ) 
  ) 
    
seu_bc[[]] %<>% 
  rownames_to_column('barcode') %>% 
  left_join(mc_signatures) %>% 
  column_to_rownames('barcode')


bc_group_dist <- readRDS(file.path(params$output_dir, 'seurat_hs', 'bc_group_cluster_r_dist.rds'))
```

### Score distribution

Score distribution in spots with different BC groups

#### Table

```{r mc-colocalization--table}

use_data <- seu_bc[[]] %>% 
  filter(!is.na(bc_group_cluster)) %>% 
  dplyr::rename(
    `M1 score` = M1_UCell,
    `TAM score` = TAM_UCell,
    `TAM/M1 score` = tam_score_khaliq
  )
  
use_score_vars <- c('M1 score', 'TAM score', 'TAM/M1 score')

use_formula <- as.formula(paste('bc_group_cluster ~', paste0("`", use_score_vars, "`", collapse = ' + ')))
tab1 <- tableby(use_formula, data=use_data,
                control = tableby.control(test = TRUE, total = FALSE))
summary(tab1, pfootnote=TRUE, title='Distribution of MC scores in BC group spots')

```


#### M1 score
```{r mc-colocalization-m1, fig.width = 0.984, fig.asp = 1}
use_comparisons <- combn(levels(seu_bc$bc_group_cluster), 2, simplify = FALSE)
seu_bc[[]] %>%
  filter(!is.na(bc_group_cluster)) %>% 
    ggplot(aes(bc_group_cluster, M1_UCell, fill = bc_group_cluster)) +
    geom_violin(scale = 'width', alpha = 0.7, 
                linewidth = linewidth_default/2, 
                trim = FALSE,
                adjust = 2
    ) +
    geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA, linewidth = linewidth_default/2) +
    scale_fill_manual(values = bc_group_cluster_palette) +
  stat_compare_means(
    comparisons = use_comparisons,
    bracket.size = linewidth_default/2,
    size = geom_text_size * 0.8
    )+
    labs(
      title = '',
      x = '',
      y = 'M1 score'#,
      # caption = paste0('Distance cropped at ',cropped_val)
    ) +
    theme(
      legend.position = "none",
      plot.margin = unit(c(0, 0.01, -0.05, 0.01), "inches"),
      axis.ticks.x = element_blank(),
      axis.text.x = element_blank()
    )
  
```



#### TAM score
```{r mc-colocalization-tam, fig.width = 0.984, fig.asp = 1}
use_comparisons <- combn(levels(seu_bc$bc_group_cluster), 2, simplify = FALSE)
seu_bc[[]] %>%
  filter(!is.na(bc_group_cluster)) %>% 
    ggplot(aes(bc_group_cluster, TAM_UCell, fill = bc_group_cluster)) +
    geom_violin(scale = 'width', alpha = 0.7, 
                linewidth = linewidth_default/2, 
                trim = FALSE,
                adjust = 2
    ) +
    geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA, linewidth = linewidth_default/2) +
    scale_fill_manual(values = bc_group_cluster_palette) +
    # scale_y_continuous(expand = expansion(mult = c(0, 0.25)),
    #                    breaks = y_axis_breaks,
    #                    labels = y_axis_labels
    # ) +
  stat_compare_means(
    comparisons = use_comparisons,
    bracket.size = linewidth_default/2,
    size = geom_text_size * 0.8
    )+
    labs(
      title = '',
      x = '',
      y = 'TAM score'#,
      # caption = paste0('Distance cropped at ',cropped_val)
    ) +
    theme(
      legend.position = "none",
      plot.margin = unit(c(0, 0.01, -0.05, 0.01), "inches"),
      axis.ticks.x = element_blank(),
      axis.text.x = element_blank()
    )
  
```





## ECM

Load ECM UCell results
```{r ecm-ucell}
ecm_ucell <- readRDS(file.path(params$output_dir, 'seurat_mm/ucell_ecm.rds'))
ucell_vars <- setdiff(names(ecm_ucell), 'sample_barcode')
use_data <- seu_bc[[]] %>% 
  filter(!is.na(bc_group_cluster))  %>%
  rownames_to_column('barcode') %>%
  left_join(ecm_ucell, by = c('barcode' = 'sample_barcode'))

```


### Summary statistics
```{r bc_group_clusters-ecm-table}
use_formula <- as.formula(paste0("bc_group_cluster ~ ", paste(ucell_vars, collapse = '+')))
tab1 <- tableby(use_formula, 
                data= use_data, 
                total = FALSE,
                control= tableby.control(
                  numeric.stats=c("mean", "sd", "median", "q1q3")
                )
)
summary(tab1)
```

### Ucell scores distribution
```{r bc_group_clusters-ecm-score-dist, fig.width= 1.2, fig.asp = 0.738}
ct <- ucell_vars[1]
for(ct in ucell_vars) {
  cat("####", ct, "\n\n")
  
  stat.test <- compare_means(
    as.formula(paste0(ct,  "~ bc_group_cluster")), 
    data = use_data,
    method = "wilcox.test"
  ) %>% 
    mutate(
      y.position = 1 + 1*0.2
    )

  
  res_plot <- use_data %>% 
    ggplot(aes(bc_group_cluster, .data[[ct]], fill = bc_group_cluster)) +
    geom_violin(scale = 'width', alpha = 0.7, 
                linewidth = linewidth_default/2, 
                trim = FALSE,
                adjust = 2
    ) +
    geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA, linewidth = linewidth_default/2) +
    scale_fill_manual(values = bc_group_cluster_palette) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.25)), breaks = c(0, 0.5, 1)) +
    stat_pvalue_manual(
      stat.test, 
      label = "p.signif", 
      tip.length = 0.01, 
      step.increase = 0.15, 
      size = geom_text_size * 0.8,
      bracket.size = linewidth_default/2
    ) +
    labs(
      x = '',
      y = 'Score',
      title = ''
    ) +
    theme(
      legend.position = "none",
      plot.margin = unit(c(-0.1, 0.01, -0.05, 0.01), "inches"),
      axis.ticks.x = element_blank(),
      axis.text.x = element_blank()
    )
  print(res_plot)
  cat("\n\n")
}
```


### Distances curves (max `r params$max_dist`` microns)

```{r bc_group_clusters-ecm-expression-distance-max_dist, fig.width= 1.2, fig.asp = 0.738}
ct <- ucell_vars[1]
for(ct in ucell_vars) {
  cat("####", ct, "\n\n")
  
  res_plot <-  use_data %>% 
    mutate(r_dist = r_dist_caf_hotspot_3d_transformed) %>% 
    filter(r_dist < params$max_dist) %>%
    mutate(
      r_dist = ifelse(r_dist < 0, 0, r_dist)
    ) %>%
    ggplot(aes(r_dist/1000, .data[[ct]], color = bc_group_cluster)) +
    geom_smooth(method = "gam", 
                formula = y ~ s(x, bs = "cs"),
                se = FALSE,
                linewidth = linewidth_default*2) +
    scale_color_manual(values = bc_group_cluster_palette) +
    scale_x_continuous(
      expand = expansion(mult = c(0, 0.15))
    ) +
    labs(
      title = '',
      x = 'Distance (mm)',
      y = 'Score',
      color = ''
    )  +
    theme(
      legend.position = "none",
      plot.margin = unit(c(-0.1, 0.01, 0, 0.01), "inches")
    )
  print(res_plot)
  cat("\n\n")

}


```

#### Statistical test

Use Generalized Additive Models (GAMs) to evaluate relationship between expression and distance across the groups. We used Cubic Regression Splines with shrinkage to allow for automated smoothing parameter selection and to penalize non-significant wiggles toward a zero-effect null space. All models were fitted using Restricted Maximum Likelihood (REML).
```{r bc_group_clusters-ecm-expression-distanc-max_dist-test}
require(mgcv)   # For fitting GAMs
require(gratia) # For plotting and difference smooths

for(ct in ucell_vars) {
  cat("####", ct, "\n\n")
  use_data_test <- use_data %>% 
    mutate(r_dist = r_dist_caf_hotspot_3d_transformed) %>% 
    filter(r_dist < params$max_dist) %>% 
    mutate(r_dist = r_dist / 1000,
           r_dist = ifelse(r_dist < 0, 0, r_dist)) %>% 
    mutate(Expression = .data[[ct]])
  
  # Global assessment, comparing null and full model
  m_null <- gam(Expression ~ s(r_dist, bs = "cs"), data = use_data_test, method = "REML")
  m_full <- gam(Expression ~ bc_group_cluster + s(r_dist, by = bc_group_cluster, bs = "cs"), 
                data = use_data_test, method = "REML")
  global_test <- anova(m_null, m_full, test = "F")
  
  cat("Global Comparison: The overall effect of Condition was assessed by comparing a full model (allowing condition-specific smooths) against a null model (a single smooth for all data) using an Analysis of Deviance (F-test):\n\n")
  global_test %>% 
    broom::tidy() %>% 
    mutate(
      term = c('Null model', 'Full model'),
      p.value = format.pval(p.value, 2)
    ) %>% 
    kbl(caption = 'Global test for differences in expression-distance curves') %>%
    kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>% 
    print()
  
  # Specific differences
  # We convert Condition to an "ordered" factor. 
  # This tells mgcv to treat the first level as a reference and others as differences.
  data_ordered <- use_data_test
  data_ordered$bc_group_cluster <- as.ordered(data_ordered$bc_group_cluster)
  contrasts(data_ordered$bc_group_cluster) <- "contr.treatment" # Use treatment contrasts
  
  m_diff <- gam(Expression ~ bc_group_cluster + s(r_dist, bs = "cs") + 
                  s(r_dist, by = bc_group_cluster, bs = "cs"), 
                data = data_ordered, method = "REML")
  
  # Extract the parametric table (Intercepts and Magnitude differences)
  cat("Pairwise Comparison: Parametric coefficients. This table shows the differences in magnitude (the average vertical position of your curves). \n\n")
  
  parametric_table <- tidy(m_diff, parametric = TRUE)
  parametric_table %>% 
    mutate(
      term = c('Intercept', levels(data_ordered$bc_group_cluster)[-1]),
      p.value = format.pval(p.value, 2)
    ) %>%
    kbl(caption = 'Parametric (magnitude) differences by group.') %>%
    kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>% 
    print()
  
  # Extract the Smooth table (Shape/Slope differences)
  cat("Pairwise Comparison: Smooth terms. This table shows the differences in shape of the relationship between expression and distance. \n\n")
  smooth_table <- tidy(m_diff, parametric = FALSE)
  smooth_table %>% 
    mutate(
      term = c('Intercept', levels(data_ordered$bc_group_cluster)[-1]),
      p.value = format.pval(p.value, 2)
    ) %>%
    kbl(caption = 'Smooth (shape) differences by group.') %>%
    kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>% 
    print()
  
  cat("\n\n")
  
}
```



### Distances curves (max 200 microns)

```{r bc_group_clusters-ecm-expression-distance-max100, fig.width= 1.2, fig.asp = 0.738}
ct <- ucell_vars[1]
for(ct in ucell_vars) {
  cat("####", ct, "\n\n")
  
  res_plot <-  use_data %>% 
    mutate(r_dist = r_dist_caf_hotspot_3d_transformed) %>% 
    filter(r_dist < 200) %>%
    mutate(
      r_dist = ifelse(r_dist < 0, 0, r_dist)
    ) %>%
    ggplot(aes(r_dist/1000, .data[[ct]], color = bc_group_cluster)) +
    geom_smooth(method = "gam", 
                formula = y ~ s(x, bs = "cs"),
                se = FALSE,
                linewidth = linewidth_default*2) +
    scale_color_manual(values = bc_group_cluster_palette) +
    scale_x_continuous(
      expand = expansion(mult = c(0, 0.15))
    ) +
    labs(
      title = '',
      x = 'Distance (mm)',
      y = 'Score',
      color = ''
    )  +
    theme(
      legend.position = "none",
      plot.margin = unit(c(-0.1, 0.01, 0, 0.01), "inches")
    )
  print(res_plot)
  cat("\n\n")

}


```

#### Statistical test

Use Generalized Additive Models (GAMs) to evaluate relationship between expression and distance across the groups. We used Cubic Regression Splines with shrinkage to allow for automated smoothing parameter selection and to penalize non-significant wiggles toward a zero-effect null space. All models were fitted using Restricted Maximum Likelihood (REML).
```{r bc_group_clusters-ecm-expression-distanc-max100-test}
require(mgcv)   # For fitting GAMs
require(gratia) # For plotting and difference smooths

for(ct in ucell_vars) {
  cat("####", ct, "\n\n")
  use_data_test <- use_data %>% 
    mutate(r_dist = r_dist_caf_hotspot_3d_transformed) %>% 
    filter(r_dist < 200) %>% 
    mutate(r_dist = r_dist / 1000,
           r_dist = ifelse(r_dist < 0, 0, r_dist)) %>% 
    mutate(Expression = .data[[ct]])
  
  # Global assessment, comparing null and full model
  m_null <- gam(Expression ~ s(r_dist, bs = "cs"), data = use_data_test, method = "REML")
  m_full <- gam(Expression ~ bc_group_cluster + s(r_dist, by = bc_group_cluster, bs = "cs"), 
                data = use_data_test, method = "REML")
  global_test <- anova(m_null, m_full, test = "F")
  
  cat("Global Comparison: The overall effect of Condition was assessed by comparing a full model (allowing condition-specific smooths) against a null model (a single smooth for all data) using an Analysis of Deviance (F-test):\n\n")
  global_test %>% 
    broom::tidy() %>% 
    mutate(
      term = c('Null model', 'Full model'),
      p.value = format.pval(p.value, 2)
    ) %>% 
    kbl(caption = 'Global test for differences in expression-distance curves') %>%
    kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>% 
    print()
  
  # Specific differences
  # We convert Condition to an "ordered" factor. 
  # This tells mgcv to treat the first level as a reference and others as differences.
  data_ordered <- use_data_test
  data_ordered$bc_group_cluster <- as.ordered(data_ordered$bc_group_cluster)
  contrasts(data_ordered$bc_group_cluster) <- "contr.treatment" # Use treatment contrasts
  
  m_diff <- gam(Expression ~ bc_group_cluster + s(r_dist, bs = "cs") + 
                  s(r_dist, by = bc_group_cluster, bs = "cs"), 
                data = data_ordered, method = "REML")
  
  # Extract the parametric table (Intercepts and Magnitude differences)
  cat("Pairwise Comparison: Parametric coefficients. This table shows the differences in magnitude (the average vertical position of your curves). \n\n")
  
  parametric_table <- tidy(m_diff, parametric = TRUE)
  parametric_table %>% 
    mutate(
      term = c('Intercept', levels(data_ordered$bc_group_cluster)[-1]),
      p.value = format.pval(p.value, 2)
    ) %>%
    kbl(caption = 'Parametric (magnitude) differences by group.') %>%
    kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>% 
    print()
  
  # Extract the Smooth table (Shape/Slope differences)
  cat("Pairwise Comparison: Smooth terms. This table shows the differences in shape of the relationship between expression and distance. \n\n")
  smooth_table <- tidy(m_diff, parametric = FALSE)
  smooth_table %>% 
    mutate(
      term = c('Intercept', levels(data_ordered$bc_group_cluster)[-1]),
      p.value = format.pval(p.value, 2)
    ) %>%
    kbl(caption = 'Smooth (shape) differences by group.') %>%
    kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>% 
    print()
  
  cat("\n\n")
  
}
```

