---
title: "PDAC-3 spatial transcriptomics"
subtitle: "Analysis of distance to edge an center for each barcode group"
author: "Francesc Castro-Giner"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
params:
  date: '`r format(Sys.Date(), "%B %d, %Y")`'
  donor: PDAC-3
  output_dir: ./output/rnaseq-spatial/pdac3
  seurat_dir: ./data/rnaseq-spatial/pdac3
  qupath_edge_dir: ./data/slide-images/pdac3-qupath/edge_annotation
---



## Configuration

Load libraries, additional functions and data

Setup environment
```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(results='asis', echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.align = 'center', fig.width = 3.5, fig.asp = 0.618, dpi = 600, dev = c("png", "pdf"), fig.showtext = FALSE, engine.opts = list(bash = "-l"))

options(stringsAsFactors = FALSE)

use_seed <- 1100101
set.seed(use_seed)

dir.create(params$output_dir, recursive = TRUE, showWarnings = FALSE)
```

Load packages
```{r load-libraries}
library(tidyverse)
library(foreach)
library(knitr)
library(magrittr)
library(ggpubr)
library(arsenal)
library(Seurat)
library(SummarizedExperiment)
library(semla)
library(sf)
library(dbscan)


library(plotly)
```

Load ggplot theme
```{r ggplot-theme}
source("./configuration/rmarkdown/ggplot_theme.R")
```

Set color palettes
```{r, echo = FALSE}
source("./configuration/rmarkdown/color_palettes.R")
```

Clean files generated in previous runs
```{r clean-files}
rmd_file <- current_input()
if(!is.null(rmd_file)) {
  figures_dir <- file.path('./docs/figure',rmd_file)
  if(dir.exists(figures_dir)) {
    unlink(file.path(figures_dir, "*"))
  }
}
```

Load custom functions
```{r load-functions}
source('./code/R-functions/semla_functions.r')
```


## Data processing

Load human seurat data and get sample_alias to sampleID mapping
```{r dw-load-seurat}
seu <- readRDS(file.path(params$seurat_dir, 'seurat_hs/seurat_merged_semla_geneNMF.rds'))
# seu <- UpdateSeuratObject(seu)
# saveRDS(seu, file.path(params$seurat_dir, 'seurat_hs/seurat_merged_semla_geneNMF.rds'))
sample_alias_to_sampleID <- GetCoordinates(seu) %>% 
  left_join(
    seu[[]] %>% 
      rownames_to_column('barcode') %>% 
      select(barcode, sample_alias)
  ) %>% 
  select(sample_alias, sampleID) %>% unique
```

Load final data from barcode analysis report
```{r dw-load-processed-data}
seu_bc <- readRDS(file.path(params$output_dir, 'seurat_hs/seurat_bc_hotspots.rds'))
seu_bc <- UpdateSeuratObject(seu_bc)
# Load hotspot data
hotspot_bc <- readRDS(file.path(params$output_dir, 'seurat_bc/semla_hotspots.rds'))
```

Add annotation to Seurat object
```{r dw-add-annotation-seurat}
# Add hotspots defined based on barcodes. We use the same criteria as in chunk::dw-bc-hotspots-to-seurat
st_spots <- hotspot_bc$st_specific %>% 
  filter(hotspot != 'singleton') %>% 
  filter(hotspot_size >= 5) %>%
  pull(sample_spot)
ctc_spots <- hotspot_bc$ctc_top %>% 
  filter(hotspot != 'singleton') %>% 
  filter(hotspot_size >= 5) %>%
  pull(sample_spot)

seu[[]] %<>% 
  rownames_to_column('sample_spot') %>%
  mutate(
    bc_group = ifelse(sample_spot %in% st_spots, 'ST', ifelse(sample_spot %in% ctc_spots, 'CTC', 'other'))
  ) %>% 
  column_to_rownames('sample_spot')

```

Extract xyz coordinates and convert to microns
```{r dw-add-xy-um}
use_seu <- seu

# Get pxl to microns conversion factor. We will use the median of the pixel spot diameter across the different slices to calculate the conversion factor, and we will use the physical spot diameter of 55 microns provided by 10x as reference for the conversion. 
physical_spot_um <- 55
use_coords_list <- semla:::.get_coords_list(use_seu)
spatnet_list <- lapply(use_coords_list, GetSpatialNetwork, maxDist = 50)
i <- spatnet_list[[1]]

um_per_pxl <- foreach(i = spatnet_list) %do% {
  pixel_spot_diameter <- i[[1]] %>% group_by(from) %>% summarise(min_distance = min(distance)) %>% pull(min_distance) %>% median
  physical_spot_um / pixel_spot_diameter
} %>% unlist %>% median


# Convert_coordinates
xy_coord <- use_seu@tools$Staffli@meta_data %>%  
  select(barcode, pxl_col_in_fullres_transformed, pxl_row_in_fullres_transformed) %>%
  set_names(c('barcode', 'x', 'y')) %>% 
  mutate(
    x = x * um_per_pxl,
    y = y * um_per_pxl
  )

xyz_coord <- use_seu[[]] %>% 
  mutate(
    z = z_um
  ) %>% 
  rownames_to_column('barcode') %>%
  left_join(xy_coord, by = 'barcode') %>% 
  dplyr::select(barcode, x, y ,z)

# Add coordinates to seurat objects
seu[[]] %<>% 
  rownames_to_column('barcode') %>%
  left_join(xyz_coord, by = 'barcode') %>%
  column_to_rownames('barcode')
seu_bc[[]] %<>% 
  rownames_to_column('barcode') %>%
  left_join(xyz_coord, by = 'barcode') %>%
  column_to_rownames('barcode')

```


Load the edge annotation from QuPath for each image, and transform the coordinates
```{r dw-load-qupath}
require(sf)
qupath_edge_files <- list.files(params$qupath_edge_dir, pattern = "*.geojson", full.names = TRUE)

# Load Transformations 
rigid_transform <- readRDS(file.path(file.path(params$seurat_dir, 'seurat', 'rigid_transform.rds')))

# Add image width and height to the transformations tibble
st_object <- GetStaffli(seu)
rigid_transform %<>% 
  left_join(st_object@image_info %>% 
              select(sampleID, full_width, full_height) %>% 
              mutate(sampleID = as.numeric(sampleID))
  )


# Load and transform edge annotations for each sample. We will use the semla-naive transformation, which applies the scale, rotation and translation parameters to the coordinates.
i <- 'S2_Y14Y20_338_B1'
i <- 'S1_Y14Y20_336_D1'
edge_coords <- foreach(i = sample_alias_to_sampleID$sample_alias) %do% {
  # Remove serialnumber from the image name, because TIFF file names used by QuPath doesn't contain them
  i_short <- (str_split(i, "_") %>% unlist)[-2]
  i_short_qupath <- paste0(i_short[1], "_", i_short[2], "-", i_short[3])
  # i_short_sample_id <- paste0(i_short[1], "_", i_short[3])
  
  # Load the QuPath annotation. coordinates are in fullres pixels at this point.
  annot_file <- file.path(params$qupath_edge_dir, paste0("PDAC3_", i_short_qupath, ".geojson"))
  if(file.exists(annot_file)){
    qupath_annotation <- st_read(
      annot_file,
      quiet = TRUE)
    
    # Convert your QuPath annotation to an sf line (if it's a boundary)
    # st_cast helps if you imported it as a Polygon but only want the edge
    boundary_line <- st_cast(qupath_annotation, "LINESTRING")
    
    # Extract raw coordinates from the sf boundary object
    # st_coordinates returns a matrix with X, Y, and potentially L1/L2 (path IDs)
    boundary_coords <- as.data.frame(st_coordinates(boundary_line)) %>%
      dplyr::rename(x = X, y = Y)
    
    # Extract transformations for the sample used
    sample_id <- sample_alias_to_sampleID %>% 
      filter(sample_alias == i) %>% 
      pull(sampleID)
    tr <- rigid_transform %>% filter(sampleID == sample_id) 
    
    
    # Prepare parameters for the semla-native transformation, as in semla::CoordAndImageTransform
    center <- c(tr$full_width/2, tr$full_height/2)
    xy_offset <- c(
      round(tr$tr_x*tr$full_width), 
      round(tr$tr_y*tr$full_height)
    )
    
    # Apply the semla-native transformation to the QuPath boundary coordinates
    transformed_coords <- CoordTransform(
      xy_coords = boundary_coords[,c('x', 'y')],
      angle = tr$angle,
      scalefactor = tr$scalefactor,
      center = center,
      xy_offset = xy_offset
    )
    
    # Create a new boundary coordinates dataframe with the transformed coordinates
    boundary_coords_t <- data.frame(
      x = transformed_coords$tr_x,
      y = transformed_coords$tr_y
    )
    
    return(boundary_coords_t)
  } else {
    return(NULL)
  }
}
names(edge_coords) <- sample_alias_to_sampleID$sample_alias
edge_coords <- compact(edge_coords)

```

Convert edge coordinates to microns and add z coordinates
```{r dw-edge-to-microns-3d}
# Get pxl to microns conversion factor. We will use the median of the pixel spot diameter across the different slices to calculate the conversion factor, and we will use the physical spot diameter of 55 microns provided by 10x as reference for the conversion. 
physical_spot_um <- 55
use_coords_list <- semla:::.get_coords_list(seu)
spatnet_list <- lapply(use_coords_list, GetSpatialNetwork, maxDist = 50)
i <- spatnet_list[[1]]

um_per_pxl <- foreach(i = spatnet_list) %do% {
  pixel_spot_diameter <- i[[1]] %>% group_by(from) %>% summarise(min_distance = min(distance)) %>% pull(min_distance) %>% median
  physical_spot_um / pixel_spot_diameter
} %>% unlist %>% median

# Get z_coordinates
z_coord <- read_csv('./data/external/spatial_tx_annotation/pdac3_st_z_coordinates.csv', comment = "#") %>% select(sample_alias, z_um)

# Convert_coordinates
edge_coords_um <- foreach(i = names(edge_coords)) %do% {
  use_coords <- edge_coords[[i]]
  use_coords$x <- use_coords$x * um_per_pxl
  use_coords$y <- use_coords$y * um_per_pxl
  use_coords$z <- z_coord %>% filter(sample_alias == i) %>% pull(z_um) 
  return(use_coords)
}
names(edge_coords_um) <- names(edge_coords)
```


Calculate distance to centroid for each point. We will use the centroids calculated in the previous step to calculate the distance from each point to the centroid of its corresponding slice. 
```{r dw-centroid-dist}
centroid_dist <- foreach(i = unique(xyz_coord$z)) %do% {
  centroid_xy <- xyz_coord %>% 
    filter(z == i) %>% 
    summarise(
      x = mean(x),
      y = mean(y)
    )
  coord_i <- xyz_coord %>% 
    filter(z == i) %>% 
    mutate(
      dist_center = sqrt((x - centroid_xy$x)^2 + (y - centroid_xy$y)^2)
    )
} %>% bind_rows()

# Add distance to seurat objects
seu[[]] %<>% 
  rownames_to_column('barcode') %>%
  left_join(centroid_dist, by = 'barcode') %>%
  column_to_rownames('barcode')
seu_bc[[]] %<>% 
  rownames_to_column('barcode') %>%
  left_join(centroid_dist, by = 'barcode') %>%
  column_to_rownames('barcode')

```

Calculate distance to the closest edge for each point. 
```{r dw-edge-dist}
require(dbscan)

# Remove  "S1_Y14Y20_336_D1". Simran:  I had issues with samples in section S1_336_D1 and a part of it actually came off. So, please remove this from the edge analysis since we don’t have the entire section on the slide.
xyz_coord_f <- xyz_coord %>% filter(!grepl('S1_Y14Y20_336_D1', barcode))

# Calculate 3d distances
query_coords <- bind_rows(edge_coords_um)
knn_res <- kNN(x = query_coords[, c("x", "y", "z")] |> as.matrix(),
               query = xyz_coord_f[, c("x", "y", "z")] |> as.matrix(),
               k = 1)
edge_dist_3d <- data.frame(
  barcode = xyz_coord_f$barcode,
  dist_edge_3d = knn_res$dist[, 1, drop = TRUE]
)


# Calculate 2d distance 
use_xyz <- xyz_coord_f %>% 
  left_join(seu[[]] %>% 
              rownames_to_column('barcode') %>% 
              select(barcode, sample_alias), 
            by = 'barcode')
use_xyz <- split(use_xyz, use_xyz$sample_alias)
edge_dist_2d <- foreach(i = names(use_xyz), .combine = rbind) %do% {
  knn_res <- kNN(x = edge_coords_um[[i]][, c("x", "y")] |> as.matrix(),
               query = use_xyz[[i]][, c("x", "y")] |> as.matrix(),
               k = 1)
  data.frame(
    barcode = use_xyz[[i]]$barcode,
    dist_edge_2d = knn_res$dist[, 1, drop = TRUE]
  )
}

# Add distance to seurat objects
seu[[]] %<>% 
  rownames_to_column('barcode') %>%
  left_join(edge_dist_3d, by = 'barcode') %>%
  left_join(edge_dist_2d, by = 'barcode') %>%
  column_to_rownames('barcode')
seu_bc[[]] %<>% 
  rownames_to_column('barcode') %>%
  left_join(edge_dist_3d, by = 'barcode') %>%
  left_join(edge_dist_2d, by = 'barcode') %>%
  column_to_rownames('barcode')
```



## Tissue edge plots

Plot annotated tissue edges on top of the spatial coordinates for each sample. This will allow us to visually inspect the quality of the edge annotation and the transformation applied to the coordinates.
```{r edge-plots, fig.width=12, fig.asp=1.3}
edge_plots <- foreach(i = sample_alias_to_sampleID$sample_alias) %do% {
  if(i %in% names(edge_coords)) {
  boundary_coords <- edge_coords[[i]]
  sample_id <- sample_alias_to_sampleID %>% 
    filter(sample_alias == i) %>% 
    pull(sampleID)
  
  res <- GetCoordinates(seu) %>% 
    filter(sampleID == sample_id) %>% 
    ggplot(
      aes(x = pxl_col_in_fullres_transformed, y = pxl_row_in_fullres_transformed),
    ) +
    geom_point(size = 0.1, alpha = 0.5) +
    geom_point(data = boundary_coords, 
               aes(x = x, y = y),
               color = 'red') +
    
    theme_minimal() +
    labs(
      x = 'x',
      y = 'y',
      title = i)
  
  
   # we need to reverse the y-axis because the coordinates are in image space, where (0,0) is the top-left corner and y increases downwards. In ggplot2, the default is Cartesian space, where (0,0) is the bottom-left corner and y increases upwards. By adding scale_y_reverse(), we flip the y-axis so that it matches the image coordinate system, allowing the points to be plotted in their correct spatial locations relative to the tissue edge.
   res <- res + scale_y_reverse()
   
   return(res)
  } else {
    return(NULL)
  }
}
edge_plots <- compact(edge_plots)
patchwork::wrap_plots(edge_plots, ncol = 5)
```


## 3D visualization of physical distances

### Center
```{r physical-distances-center}
fig <- plot_ly(
  seu[[]] , 
  x = ~x, y = ~y, z = ~z,
  color = ~dist_center, 
  colors = 'YlOrRd',
  # alpha_stroke = 0,
  marker = list(
    opacity = 0.2,
    size = 2
  )
)

fig %>% 
  layout(
    title = 'Distance the center',
    legend= list(itemsizing='constant'),
    scene = list(
      xaxis = list(showgrid = FALSE, zeroline = FALSE,
                   showticklabels=FALSE, title = ''),
      yaxis = list(showgrid = FALSE, zeroline = FALSE,
                   showticklabels=FALSE, title = ''),
      zaxis = list(showgrid = FALSE, zeroline = FALSE,
                   showticklabels=FALSE, title = '')
    )
    ) # keep the legend makers the same size
```

### Edge (3D)
```{r physical-distances-edge-3d}
fig <- plot_ly(
  seu[[]] , 
  x = ~x, y = ~y, z = ~z,
  color = ~dist_edge_3d, 
  colors = 'YlOrRd',
  # alpha_stroke = 0,
  marker = list(
    opacity = 0.2,
    size = 2
  )
)

fig %>% 
  layout(
    title = 'Distance (3D) to the closest edge point',
    legend= list(itemsizing='constant'),
    scene = list(
      xaxis = list(showgrid = FALSE, zeroline = FALSE,
                   showticklabels=FALSE, title = ''),
      yaxis = list(showgrid = FALSE, zeroline = FALSE,
                   showticklabels=FALSE, title = ''),
      zaxis = list(showgrid = FALSE, zeroline = FALSE,
                   showticklabels=FALSE, title = '')
    )
    ) # keep the legend makers the same size
```

### Edge (2dD)
```{r physical-distances-edge-2d}
fig <- plot_ly(
  seu[[]] , 
  x = ~x, y = ~y, z = ~z,
  color = ~dist_edge_2d, 
  colors = 'YlOrRd',
  # alpha_stroke = 0,
  marker = list(
    opacity = 0.2,
    size = 2
  )
)

fig %>% 
  layout(
    title = 'Distance (2D) to the closest edge point',
    legend= list(itemsizing='constant'),
    scene = list(
      xaxis = list(showgrid = FALSE, zeroline = FALSE,
                   showticklabels=FALSE, title = ''),
      yaxis = list(showgrid = FALSE, zeroline = FALSE,
                   showticklabels=FALSE, title = ''),
      zaxis = list(showgrid = FALSE, zeroline = FALSE,
                   showticklabels=FALSE, title = '')
    )
    ) # keep the legend makers the same size
```




## BC goups
### Compare distances
```{r bc-groups-physical-distances-table}
res_tab <- tableby(
  bc_group_cluster ~ dist_center + dist_edge_3d + dist_edge_2d, 
  data = seu_bc[[]],
  control = tableby.control(total=FALSE)
)
summary(
  res_tab, pfootnote=TRUE
)
```

### Violin plot distances
```{r bc-groups-physical-distances-violin_conf}
xvar <- 'bc_group_cluster'
use_palette <- bc_group_cluster_palette
use_data <- seu_bc[[]] %>% 
  filter(!is.na(.[[xvar]]))
use_comparisons <- combn(levels(seu_bc[[]][,xvar]), 2, simplify = FALSE)
```

#### Center
```{r bc-groups-physical-distances-violin-center, fig.asp = 0.718}
ggviolin(use_data, 
          x = xvar, y = "dist_center",
          color = xvar, palette = use_palette,
         add = "boxplot", add.params = list(fill = "white")) + 
  stat_compare_means(comparisons = use_comparisons, label = "p.signif") +
  labs(
    title = 'Distance to the center',
    x = '',
    y = 'Distance (microns)'
  ) +
  theme(legend.position = "none")
```

#### Edge
```{r bc-groups-physical-distances-violin-edge3d, fig.asp = 0.718}
ggviolin(use_data, 
          x = xvar, y = "dist_edge_3d",
          color = xvar, palette = use_palette,
         add = "boxplot", add.params = list(fill = "white")) + 
  stat_compare_means(comparisons = use_comparisons, label = "p.signif") +
  labs(
    title = 'Distance to the edge (3D)',
    x = '',
    y = 'Distance (microns)',
    caption = 'Distance calculated to the closest edge point'
  ) +
  theme(legend.position = "none")
```

