---
title: "Analysis of PDAC-3 and patients"
subtitle: "Intersection of GSEA results across models for sample type"
author: "Francesc Castro-Giner"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
params:
  date: '`r format(Sys.Date(), "%B %d, %Y")`'
  gse_file_regex_xenograft: ./output/rnaseq/DONOR/differential_expression_edger/sample_type/gse_omnibus-edgeR_QLF.rds
  model_xenograft: pdac3-ctc
  gse_file_regex_patient: ./output/rnaseq/DONOR/differential_expression_edger/donor_batch/sample_type/gse_omnibus-edgeR_QLF.rds
  model_patient: patients-ctc
---

## Load libraries, additional functions and data

Setup environment
```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(results='asis', echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.align = 'center', dev = c("png", "pdf"), engine.opts = list(bash = "-l"))

options(stringsAsFactors = FALSE)

use_seed <- 1100101
set.seed(use_seed)
```

Load packages
```{r load-libraries}
library(tidyverse)
library(showtext)
library(foreach)
library(knitr)
library(kableExtra)
library(magrittr)
library(patchwork)
```

Load ggplot theme
```{r ggplot-theme}
source("./configuration/rmarkdown/ggplot_theme.R")
```

Set color palettes
```{r, echo = FALSE}
source("./configuration/rmarkdown/color_palettes.R")
```

Load custom functions
```{r load-functions}
source('./code/R-functions/gse_report.r')
source('./code/R-functions/subchunkify.R')
```


Clean files generated in previous runs
```{r clean-files}
rmd_file <- current_input()
if(!is.null(rmd_file)) {
  figures_dir <- file.path('./docs/figure',rmd_file)
  if(dir.exists(figures_dir)) {
    unlink(file.path(figures_dir, "*"))
  }
}
```


## Data wrangling

Load gse_omnibus files
```{r conf-gse-files}
# Load xengraft models
list_models_x <- str_split(params$model_xenograft, ',', simplify = FALSE) %>% unlist
gse_list_x <- foreach(model = list_models_x) %do% {
  gse_file <- params$gse_file_regex_xenograft %>%
    gsub('DONOR', model, .)
  if(file.exists(gse_file)) {
    readRDS(gse_file)
  } else {
    NULL
  }
}
names(gse_list_x) <- list_models_x

# Load patient models
list_models_p <- str_split(params$model_patient, ',', simplify = FALSE) %>% unlist
gse_list_p <- foreach(model = list_models_p) %do% {
  gse_file <- params$gse_file_regex_patient %>%
    gsub('DONOR', model, .)
  if(file.exists(gse_file)) {
    readRDS(gse_file)
  } else {
    NULL
  }
}
names(gse_list_p) <- list_models_p

gse_list <- c(gse_list_x, gse_list_p)
list_models <- c(list_models_p, list_models_x)
```

Modify xenografts comparison names to be more intuitive
```{r conf-gse-rename}
# Xenografts
use_comps <- c("CTO-over-S-mouse_id", "CTO-over-Cl-mouse_id")
new_comps <- c("CTO-over-Single", "CTO-over-Cluster")
for(i in c('pdac3-ctc')){
  gse_list[[i]] <- gse_list[[i]][use_comps]
  names(gse_list[[i]]) <- new_comps
}

# Patients
use_comps <- c("CTO-over-S-donor", "CTO-over-Cl-donor")
new_comps <- c("CTO-over-Single", "CTO-over-Cluster")
for(i in c('patients-ctc')){
  gse_list[[i]] <- gse_list[[i]][use_comps]
  names(gse_list[[i]]) <- new_comps
}
```

Get full list of comparisons done
```{r conf-gse-comparisons}
comp_list <- foreach(i = gse_list) %do% {
  names(i)
} %>% unlist %>% unique
```

Merge GSE files
```{r conf-gse-merge}
model <- names(gse_list)[1]
comp <- comp_list[1]
sr <- 'GSEA'
gc <- 'msigdb.c2.cp'
gse_list_merged <- foreach(comp = comp_list) %do% {
  res <- foreach(model = list_models, .combine = rbind) %do% {
    x <- gse_list[[model]][[comp]][[sr]][[gc]]
    if(!is.null(x)) {
      return(x@result %>% 
        mutate(
          model = model,
          model = ifelse(model == 'patients-ctc', 'Patient', model),
          model = ifelse(model == 'pdac3-ctc', 'PDAC-3', model),
          model = factor(model, levels = c('PDAC-3', 'Patient'))
          )
      )
    } else {
      return(NULL)
    }
  }
  return(res)
}
names(gse_list_merged) <- comp_list
```

Set group levels manually, for extended plots
```{r conf-gse-groups-levels}
group_levels <- list(
  `CTO-over-Single` = c(
    "Cell cycle",
    "Translation",
    "Metabolism",
    "Cytoskeleton",
    "Immune response",
    "DNA replication",
    "Cellular response to stress",
    "RNA processing",
    "OxPhos and ATP synthesis",
    "MET signalling",
    "ECM degradation",
    "Interferron signalling",
    "Cell-matrix interaction" 
  ),
  `CTO-over-Cluster` = c(
    "Translation",
    "DNA replication",
    "Cell-cell junction",
    "Cytoskeleton",
    "Cell cycle",
    "ECM organization",
    "Met Pathway",
    "Cell-matrix interactions",
    "Metabolism",
    "Immune response",
    "Interferon signalling"
  )
)
```


Add groups to gene sets, generated manually
```{r conf-gse-groups}
# Get gene set grouping 
gsets_grouping_file <- './data/external/pdac3_gsea_gsets_groups_20260408.xlsx' 
gsets_sheets <- readxl::excel_sheets(path = gsets_grouping_file)
gsets_groups <- foreach(i = gsets_sheets) %do% {
  readxl::read_xlsx(gsets_grouping_file, sheet = i)
}
names(gsets_groups) <- gsets_sheets

for(comp in names(gse_list_merged)) {
  gse_list_merged[[comp]] %<>% left_join(gsets_groups[[comp]], by = 'ID') %>% 
    filter(!is.na(Group)) %>% 
    mutate(gene_set_group = factor(Group, levels = group_levels[[comp]]))
}
```



Add group stats
```{r conf-gse-group-stats}
comp <- names(gse_list_merged)[1]
for(comp in names(gse_list_merged)) {
  gs_nsign <-  gse_list_merged[[comp]] %>% 
    group_by(ID) %>%
    summarise(
      n_significant = sum(p.adjust < 0.05),
      n_up = sum(NES > 0),
      avg_fdr = mean(p.adjust),
      avg_nes = mean(NES)
    )
  gse_list_merged[[comp]] %<>% left_join(gs_nsign, by = 'ID')
}

```


## Dot plots by comparison (complete)

```{r gse-dotplots, echo=FALSE, results='asis'}
knitr::opts_knit$set(
  base.dir = "docs"
  )
root_section_number <- 3
section_n <- 1

comp <- names(gse_list_merged)[1]
for(comp in names(gse_list_merged)) {
  h_section_number <- paste(root_section_number, section_n, sep=".")
  cat("<h3> <span class=\"header-section-number\">",h_section_number,"</span>", comp, "</h3>\n\n") #Cause results='asis',we must use raw HTML
  section_n <- section_n + 1
  x <- gse_list_merged[[comp]] %>% 
    arrange(desc(n_up), desc(avg_nes)) %>% 
    mutate(
      is.sign = ifelse(p.adjust < 0.05, 'yes', 'no'),
      log10P = -log10(p.adjust),
      Description = clean_msigdb_names(Description),
      Description = fct_reorder(Description, avg_nes),
      
    )
  if(nrow(x) == 0) {
    cat("No gene sets significant in at least two models.\n\n")
  } else {
    xlim <- c(-max(abs(x$NES)), max(abs(x$NES)))
    res <- x %>% 
      ggplot(aes(x = NES, y = Description, size = log10P, fill = NES)) +
      geom_point(pch = 21, aes(color = is.sign)) +
      facet_grid(cols = vars(model), rows = vars(gene_set_group), scales = 'free_y', space = 'free_y') +
      scale_y_discrete(labels =  default_labeller(80)) +
      geom_vline(xintercept = 0, lty = 3) +
      labs(
        y = '',
        size = expression(paste("lo", g[10],"(Adjusted P value)")),
        color = 'FDR < 0.05'
        # caption = 'Gene sets with FDR<0.05 in at least 2 models',
      ) +
      scale_size_continuous(range=c(0.8, 2)) +
      scale_fill_gradient2(trans = "reverse") +
      scale_color_manual(values = c(no = 'white', yes = 'black')) +
      # scale_color_viridis_c() +
      scale_x_continuous(expand = expansion(mult = c(0.20, 0.20)),
                         limits = xlim) +
      theme(
        panel.border = element_rect(color = "black", fill = NA, linewidth = linewidth_default),
        strip.text.y = element_text(angle = 0, hjust = 0),
        
        axis.line = element_blank(),
        axis.ticks = element_line(color = "black", linewidth = linewidth_default),
        axis.text.y = element_text(size = 4),
        axis.text.x = element_text(size = 4),
        
        legend.text = element_text(size = 4), 
        legend.title = element_text(size = 4),
        legend.key.width = unit(0.2, "cm")
      )
    
    n_gsets <- x$ID %>% unique() %>% length()
    # use_width <-  7.2 + n_gsets * 0.02
    use_width <- 5.1
    # use_asp <-  0.318 + n_gsets * 0.017
    use_asp <-  n_gsets * 0.021
    use_height <- use_width * use_asp
    subchunkify(res,
                fig.width = use_width,
                fig.height = use_height,
                dpi = 600,
                prefix = 'gse-dotplot',
                suffix = comp )
    
    
    cat("\n\n")
  }
  
}
```

## Dot plots by comparison (unnamed)

```{r gse-dotplots-unnamed, echo=FALSE, results='asis'}
knitr::opts_knit$set(
  base.dir = "docs"
  )
root_section_number <- 4
section_n <- 1

comp <- names(gse_list_merged)[1]
for(comp in names(gse_list_merged)) {
  h_section_number <- paste(root_section_number, section_n, sep=".")
  cat("<h3> <span class=\"header-section-number\">",h_section_number,"</span>", comp, "</h3>\n\n") #Cause results='asis',we must use raw HTML
  section_n <- section_n + 1
  x <- gse_list_merged[[comp]] %>% 
    arrange(desc(n_up), desc(avg_nes)) %>% 
    mutate(
      is.sign = ifelse(p.adjust < 0.05, 'yes', 'no'),
      log10P = -log10(p.adjust),
      Description = clean_msigdb_names(Description),
      Description = fct_reorder(Description, avg_nes)
    )
  if(nrow(x) == 0) {
    cat("No gene sets significant in at least two models.\n\n")
  } else {

    # Remove redundancy
    x %<>% 
      group_by(model, gene_set_group, Description) %>% 
      slice_min(avg_fdr, n = 1, with_ties = FALSE)
    
    xlim <- c(-max(abs(x$NES)), max(abs(x$NES)))
    
    res <- x %>% 
      ggplot(aes(x = NES, y = Description, size = log10P, fill = NES)) +
      geom_vline(xintercept = 0, lty = 3) +
      # geom_point(pch = 21, aes(color = is.sign)) +
      geom_point(data = x %>% filter(is.sign == 'yes'), pch = 21, aes(color = is.sign)) +
      geom_point(data = x %>% filter(is.sign == 'no'), pch = 21, stroke = 0) +
      facet_grid(cols = vars(model), rows = vars(gene_set_group), scales = 'free_y', space = 'free_y') +
      labs(
        x = '',
        y = '',
        size = expression(paste("lo", g[10],"(FDR)")),
        color = 'FDR\n< 0.05'
      ) +
      scale_y_discrete(labels =  default_labeller(80)) +
      scale_x_continuous(expand = expansion(mult = c(0.20, 0.20)),
                         limits = xlim) +
      scale_size_continuous(range=c(0.5, 1.1)) +
      scale_fill_gradient2(trans = "reverse") +
      scale_color_manual(values = c(no = 'white', yes = 'black')) +
      theme(
        panel.border = element_rect(color = "black", fill = NA, linewidth = linewidth_default),
        strip.text.y = element_blank(),
        strip.text.x = element_blank(),
        
        axis.line = element_blank(),
        axis.ticks = element_line(color = "black", linewidth = linewidth_default/2),
        axis.text.y = element_blank(),
        axis.text.x = element_blank(),
        
        legend.text = element_text(size = 3), 
        legend.title = element_text(size = 3),
        legend.key.width = unit(0.1, "cm"),
        legend.key.height = unit(0.1, "cm"),
        
        plot.margin = unit(c(0.0217, 0.0217, 0.0217, 0.0217), "inches")
      )
          
    n_gsets <- x$ID %>% unique() %>% length()
    n_gsgroups <- x$gene_set_group %>% unique %>% length
    

    # Calculate width = left margin + total width of panels + total width of spacing + right margin. Everything scaled by conversion factor to get final width in inches. Conversion factor calculater with a rule in the screen
    nmodels <- x$model %>% unique() %>% length()
    # Calculate width: left margin + tickmarks + total width of panels + total width of spacing + right margin
    use_width <- 0.1165 + 0.0173 + (0.18898*nmodels) + (0.035*(nmodels - 1))+ 0.0217
    
    # Calculate height: top margin + tickmarks + bottom margin and legend
    use_height <- 0.0217 + 0.0217 + n_gsets * 0.0492 + (0.035*(nmodels - 1)) + 0.4984


    subchunkify(res +
                  guides(
                    size  = guide_legend(direction = "vertical",
                                         keyheight = unit(0.0289, "inches"),
                                         keywidth = unit(0.0289, "inches")
                                         ),
                    color = "none",
                    fill  = "none"
                    # color = guide_legend(direction = "vertical"),
                    # fill  = guide_colourbar(direction = "vertical")
                  ) +
                  theme(
                    legend.position = "bottom",
                    legend.box = "horizontal"
                  ),
                fig.width = use_width,
                fig.height = use_height,
                dpi = 600,
                prefix = 'gse-sign-dotplot-unnamed',
                suffix = comp )
    
    cat("\n\n")
    
    # Print legend
    res_legend <- get_legend(
      res +
        guides(
          color = guide_legend(direction = "vertical"),
          size  = guide_legend(direction = "vertical"),
          fill  = guide_colourbar(direction = "vertical")
        ) +
        theme(
          legend.position = "bottom",
          legend.box = "horizontal"
        )
    )
    subchunkify(plot_grid(res_legend),
                fig.width = 1.8,
                fig.height = 1,
                dpi = 600,
                prefix = 'gse-sign-dotplot-legend-unnamed',
                suffix = comp )

    cat("\n\n")
  }
  
}
```


## Dot plots by comparison (separate scale)

```{r gse-dotplots-separate, echo=FALSE, results='asis'}
knitr::opts_knit$set(
  base.dir = "docs"
  )
root_section_number <- 3
section_n <- 1

comp <- names(gse_list_merged)[1]
for(comp in names(gse_list_merged)) {
  h_section_number <- paste(root_section_number, section_n, sep=".")
  cat("<h3> <span class=\"header-section-number\">",h_section_number,"</span>", comp, "</h3>\n\n") #Cause results='asis',we must use raw HTML
  section_n <- section_n + 1

  x <- gse_list_merged[[comp]] %>% 
    arrange(desc(n_up), desc(avg_nes)) %>% 
    mutate(
      is.sign = ifelse(p.adjust < 0.05, 'yes', 'no'),
      log10P = -log10(p.adjust),
      Description = clean_msigdb_names(Description),
      Description = fct_reorder(Description, avg_nes)
    )
  if(nrow(x) == 0) {
    cat("No gene sets significant in at least two models.\n\n")
  } else {
    xlim <- c(-max(abs(x$NES)), max(abs(x$NES)))
    
    x_pdac3 <-  x %>% 
      filter(model == 'PDAC-3')
    x_patient <-  x %>% 
      filter(model == 'Patient')
    
    for(i in levels(x$Description)) {
      if(!i %in% x_pdac3$Description ) {
        x_pdac3 %<>% add_row(
          ID = x %>% filter(Description == i) %>% pull(ID) %>% unique(),
          Description = i,
          NES = 0,
          p.adjust = 1,
          model = 'PDAC-3',
          gene_set_group = x %>% filter(Description == i) %>% pull(gene_set_group) %>% unique(),
          n_significant = 0,
          n_up = 0,
          avg_fdr = 1,
          avg_nes = 0,
          log10P = 0,
          is.sign = 'no'
        )
      }
      if(!i %in% x_patient$Description ) {
        x_patient %<>% add_row(
          ID = x %>% filter(Description == i) %>% pull(ID) %>% unique(),
          Description = i,
          NES = 0,
          p.adjust = 1,
          model = 'Patient',
          gene_set_group = x %>% filter(Description == i) %>% pull(gene_set_group) %>% unique(),
          n_significant = 0,
          n_up = 0,
          avg_fdr = 1,
          avg_nes = 0,
          log10P = 0,
          is.sign = 'no'
        )
      }
    }
    
    # Remove redundancy
    x_pdac3 %<>% 
      group_by(gene_set_group, Description) %>% 
      slice_min(pvalue, n = 1, with_ties = FALSE)
    x_patient %<>% 
      group_by(gene_set_group, Description) %>% 
      slice_min(pvalue, n = 1, with_ties = FALSE)
    
    # Refactor Description
    x_pdac3 %<>% 
      mutate(
        Description = as.character(Description),
        Description = factor(Description, levels = levels(x$Description))
        )
    x_patient %<>% 
      mutate(
        Description = as.character(Description),
        Description = factor(Description, levels = levels(x$Description))
        )      
    
    # Generate plots
    res_1 <- x_pdac3 %>% 
      ggplot(aes(x = NES, y = Description, size = log10P, fill = NES)) +
      # geom_point(pch = 21, aes(color = is.sign)) +
      geom_point(pch = 21, stroke = linewidth_default/2) +
      facet_grid(cols = vars(model), rows = vars(gene_set_group), scales = 'free_y', space = 'free_y') +
      scale_y_discrete(labels =  default_labeller(80)) +
      geom_vline(xintercept = 0, lty = 3) +
      labs(
        y = '',
        size = expression(paste("lo", g[10],"(Adjusted P value)")),
        color = 'FDR < 0.05'
        # caption = 'Gene sets with FDR<0.05 in at least 2 models',
      ) +
      scale_size_continuous(range=c(0.8, 2)) +
      scale_fill_gradient2(trans = "reverse") +
      scale_color_manual(values = c(no = 'white', yes = 'black')) +
      # scale_color_viridis_c() +
      scale_x_continuous(expand = expansion(mult = c(0.20, 0.20)),
                         limits = xlim) +
      theme(
        panel.border = element_rect(color = "black", fill = NA, linewidth = linewidth_default),
        strip.text.y = element_text(angle = 0, hjust = 0),
        
        axis.line = element_blank(),
        axis.ticks = element_line(color = "black", linewidth = linewidth_default),
        axis.text.y = element_text(size = 4),
        axis.text.x = element_text(size = 4),
        
        legend.text = element_text(size = 4), 
        legend.title = element_text(size = 4),
        legend.key.width = unit(0.2, "cm"),
        panel.spacing.y = unit(0.08, "lines")
      )
    
    res_2 <- x_patient %>% 
      ggplot(aes(x = NES, y = Description, size = log10P, fill = NES)) +
      # geom_point(pch = 21, aes(color = is.sign)) +
      geom_point(pch = 21, stroke = linewidth_default/2) +
      facet_grid(cols = vars(model), rows = vars(gene_set_group), scales = 'free_y', space = 'free_y') +
      scale_y_discrete(labels =  default_labeller(80)) +
      geom_vline(xintercept = 0, lty = 3) +
      labs(
        y = '',
        size = expression(paste("lo", g[10],"(Adjusted P value)")),
        color = 'FDR < 0.05'
        # caption = 'Gene sets with FDR<0.05 in at least 2 models',
      ) +
      scale_size_continuous(range=c(0.8, 2)) +
      scale_fill_gradient2(trans = "reverse") +
      scale_color_manual(values = c(no = 'white', yes = 'black')) +
      # scale_color_viridis_c() +
      scale_x_continuous(expand = expansion(mult = c(0.20, 0.20)),
                         limits = xlim) +
      theme(
        panel.border = element_rect(color = "black", fill = NA, linewidth = linewidth_default),
        strip.text.y = element_text(angle = 0, hjust = 0),
        
        axis.line = element_blank(),
        axis.ticks = element_line(color = "black", linewidth = linewidth_default),
        axis.text.y = element_text(size = 4),
        axis.text.x = element_text(size = 4),
        
        legend.text = element_text(size = 4), 
        legend.title = element_text(size = 4),
        legend.key.width = unit(0.2, "cm"),
        panel.spacing.y = unit(0.08, "lines")
      )
    
    n_gsets <- x$ID %>% unique() %>% length()
    # use_width <-  7.2 + n_gsets * 0.02
    use_width <- 5.1
    # use_asp <-  0.318 + n_gsets * 0.017
    use_asp <-  n_gsets * 0.021
    use_height <- use_width * use_asp

    subchunkify(res_1,
                fig.width = use_width,
                fig.height = use_height,
                dpi = 600,
                prefix = 'gse-dotplot-separate-pdac3',
                suffix = comp )
    cat("\n\n")
    subchunkify(res_2,
                fig.width = use_width,
                fig.height = use_height,
                dpi = 600,
                prefix = 'gse-dotplot-separate-patient',
                suffix = comp )   
    
    cat("\n\n")
  }
  
}
```



## Dot plots by comparison (unamed and separate scale)

```{r gse-dotplots-unnamed-spearate, echo=FALSE, results='asis'}
knitr::opts_knit$set(
  base.dir = "docs"
  )
root_section_number <- 4
section_n <- 1

comp <- names(gse_list_merged)[1]
for(comp in names(gse_list_merged)) {
  h_section_number <- paste(root_section_number, section_n, sep=".")
  cat("<h3> <span class=\"header-section-number\">",h_section_number,"</span>", comp, "</h3>\n\n") #Cause results='asis',we must use raw HTML
  section_n <- section_n + 1
  x <- gse_list_merged[[comp]] %>% 
    arrange(desc(n_up), desc(avg_nes)) %>% 
    mutate(
      is.sign = ifelse(p.adjust < 0.05, 'yes', 'no'),
      log10P = -log10(p.adjust),
      Description = clean_msigdb_names(Description),
      Description = fct_reorder(Description, avg_nes)
    )
  if(nrow(x) == 0) {
    cat("No gene sets significant in at least two models.\n\n")
  } else {

    # Remove redundancy
    x %<>% 
      group_by(model, gene_set_group, Description) %>% 
      slice_min(avg_fdr, n = 1, with_ties = FALSE)
    
    xlim <- c(-max(abs(x$NES)), max(abs(x$NES)))
    
    x_pdac3 <-  x %>% 
      ungroup() %>%
      filter(model == 'PDAC-3') %>% 
      mutate(model = as.character(model))
    x_patient <-  x %>% 
      ungroup() %>%
      filter(model == 'Patient') %>% 
      mutate(model = as.character(model))
    
    
    for(i in levels(x$Description)) {
      if(!i %in% x_pdac3$Description ) {
        x_pdac3 %<>% add_row(
          ID = x %>% filter(Description == i) %>% pull(ID) %>% unique(),
          Description = i,
          NES = 0,
          p.adjust = 1,
          model = 'PDAC-3',
          gene_set_group = x %>% filter(Description == i) %>% pull(gene_set_group) %>% unique(),
          n_significant = 0,
          n_up = 0,
          avg_fdr = 1,
          avg_nes = 0,
          log10P = 0,
          is.sign = 'no'
        )
      }
      if(!i %in% x_patient$Description ) {
        x_patient %<>% add_row(
          ID = x %>% filter(Description == i) %>% pull(ID) %>% unique(),
          Description = i,
          NES = 0,
          p.adjust = 1,
          model = 'Patient',
          gene_set_group = x %>% filter(Description == i) %>% pull(gene_set_group) %>% unique(),
          n_significant = 0,
          n_up = 0,
          avg_fdr = 1,
          avg_nes = 0,
          log10P = 0,
          is.sign = 'no'
        )
      }
    }
    
    
    # Remove redundancy
    x_pdac3 %<>% 
      group_by(gene_set_group, Description) %>% 
      slice_min(pvalue, n = 1, with_ties = FALSE)
    x_patient %<>% 
      group_by(gene_set_group, Description) %>% 
      slice_min(pvalue, n = 1, with_ties = FALSE)
    
    
    # Refactor Description
    x_pdac3 %<>% 
      mutate(
        Description = as.character(Description),
        Description = factor(Description, levels = levels(x$Description))
      )
    x_patient %<>% 
      mutate(
        Description = as.character(Description),
        Description = factor(Description, levels = levels(x$Description))
      )      
    
    
    res_1 <- x_pdac3 %>%
      ggplot(aes(x = NES, y = Description, size = log10P, fill = NES)) +
      geom_vline(xintercept = 0, lty = 3) +
      geom_point(pch = 21, stroke = linewidth_default/2) +
      # geom_point(data = x_pdac3 %>% filter(is.sign == 'yes'), pch = 21, aes(color = is.sign)) +
      # geom_point(data = x_pdac3 %>% filter(is.sign == 'no'), pch = 21, stroke = 0) +
      facet_grid(cols = vars(model), rows = vars(gene_set_group), scales = 'free_y', space = 'free_y') +
      labs(
        x = '',
        y = '',
        size = expression(paste("lo", g[10],"(FDR)")),
        color = 'FDR\n< 0.05'
      ) +
      scale_y_discrete(labels =  default_labeller(80)) +
      scale_x_continuous(expand = expansion(mult = c(0.20, 0.20)),
                         limits = xlim) +
      scale_size_continuous(range=c(0.5, 1.1)) +
      scale_fill_gradient2(trans = "reverse") +
      scale_color_manual(values = c(no = 'white', yes = 'black')) +
      theme(
        panel.border = element_rect(color = "black", fill = NA, linewidth = linewidth_default),
        strip.text.y = element_blank(),
        strip.text.x = element_blank(),
        
        axis.line = element_blank(),
        axis.ticks = element_line(color = "black", linewidth = linewidth_default/2),
        axis.text.y = element_blank(),
        axis.text.x = element_blank(),
        
        legend.text = element_text(size = 3), 
        legend.title = element_text(size = 3),
        legend.key.width = unit(0.1, "cm"),
        legend.key.height = unit(0.1, "cm"),
        
        plot.margin = unit(c(0.0217, 0.0217, 0.0217, 0.0217), "inches"),
        panel.spacing.y = unit(0.08, "lines")
      )

    res_2 <- x_patient %>%
      ggplot(aes(x = NES, y = Description, size = log10P, fill = NES)) +
      geom_vline(xintercept = 0, lty = 3) +
      geom_point(pch = 21, stroke = linewidth_default/2) +
      # geom_point(data = x_patient %>% filter(is.sign == 'yes'), pch = 21, aes(color = is.sign)) +
      # geom_point(data = x_patient %>% filter(is.sign == 'no'), pch = 21, stroke = 0) +
      facet_grid(cols = vars(model), rows = vars(gene_set_group), scales = 'free_y', space = 'free_y') +
      labs(
        x = '',
        y = '',
        size = expression(paste("lo", g[10],"(FDR)")),
        color = 'FDR\n< 0.05'
      ) +
      scale_y_discrete(labels =  default_labeller(80)) +
      scale_x_continuous(expand = expansion(mult = c(0.20, 0.20)),
                         limits = xlim) +
      scale_size_continuous(range=c(0.5, 1.1)) +
      scale_fill_gradient2(trans = "reverse") +
      scale_color_manual(values = c(no = 'white', yes = 'black')) +
      theme(
        panel.border = element_rect(color = "black", 
                                    fill = NA, 
                                    linewidth = linewidth_default),
        strip.text.y = element_blank(),
        strip.text.x = element_blank(),
        
        axis.line = element_blank(),
        axis.ticks = element_line(color = "black", 
                                  linewidth = linewidth_default/2),
        axis.text.y = element_blank(),
        axis.text.x = element_blank(),
        
        legend.text = element_text(size = 3), 
        legend.title = element_text(size = 3),
        legend.key.width = unit(0.1, "cm"),
        legend.key.height = unit(0.1, "cm"),
        
        plot.margin = unit(c(0.0217, 0.0217, 0.0217, 0.0217), "inches"),
        panel.spacing.y = unit(0.08, "lines")
      )
        
        
    n_gsets <- x$ID %>% unique() %>% length()
    n_gsgroups <- x$gene_set_group %>% unique %>% length
    

    # Calculate width = left margin + total width of panels + total width of spacing + right margin. Everything scaled by conversion factor to get final width in inches. Conversion factor calculater with a rule in the screen
    # nmodels <- x$model %>% unique() %>% length()
    nmodels <- 1

    # Calculate width: left margin + tickmarks + total width of panels + total width of spacing + right margin
    use_width <- 0.1165 + 0.0173 + (0.18898*nmodels) + (0.035*(nmodels - 1))+ 0.0217
    
    # Calculate height: top margin + tickmarks + bottom margin and legend
    use_height <- 0.0217 + 0.0217 + n_gsets * 0.0492 + (0.035*(n_gsgroups - 1)) + 0.4984

    
    # Print PDAC-3
    subchunkify(res_1 +
                  guides(
                    size  = guide_legend(direction = "vertical",
                                         keyheight = unit(0.0289, "inches"),
                                         keywidth = unit(0.0289, "inches")
                                         ),
                    color = "none",
                    fill  = "none"
                    # color = guide_legend(direction = "vertical"),
                    # fill  = guide_colourbar(direction = "vertical")
                  ) +
                  theme(
                    legend.position = "bottom",
                    legend.box = "horizontal"
                  ),
                fig.width = use_width,
                fig.height = use_height,
                dpi = 600,
                prefix = 'gse-sign-dotplot-unnamed-separated-pdac3',
                suffix = comp )
    
    cat("\n\n")
    
    # Print legend
    res_legend <- get_legend(
      res_1 +
        guides(
          color = guide_legend(direction = "vertical"),
          size  = guide_legend(direction = "vertical"),
          fill  = guide_colourbar(direction = "vertical")
        ) +
        theme(
          legend.position = "bottom",
          legend.box = "horizontal"
        )
    )
    subchunkify(plot_grid(res_legend),
                fig.width = 1.8,
                fig.height = 1,
                dpi = 600,
                prefix = 'gse-sign-dotplot-legend-unnamed-separated-pdac3',
                suffix = comp )

    cat("\n\n")
    
      
    # Print Patient
    subchunkify(res_2 +
                  guides(
                    size  = guide_legend(direction = "vertical",
                                         keyheight = unit(0.0289, "inches"),
                                         keywidth = unit(0.0289, "inches")
                                         ),
                    color = "none",
                    fill  = "none"
                    # color = guide_legend(direction = "vertical"),
                    # fill  = guide_colourbar(direction = "vertical")
                  ) +
                  theme(
                    legend.position = "bottom",
                    legend.box = "horizontal"
                  ),
                fig.width = use_width,
                fig.height = use_height,
                dpi = 600,
                prefix = 'gse-sign-dotplot-unnamed-separated-patient',
                suffix = comp )
    
    cat("\n\n")
    
    # Print legend
    res_legend <- get_legend(
      res_2 +
        guides(
          color = guide_legend(direction = "vertical"),
          size  = guide_legend(direction = "vertical"),
          fill  = guide_colourbar(direction = "vertical")
        ) +
        theme(
          legend.position = "bottom",
          legend.box = "horizontal"
        )
    )
    subchunkify(plot_grid(res_legend),
                fig.width = 1.8,
                fig.height = 1,
                dpi = 600,
                prefix = 'gse-sign-dotplot-legend-unnamed-separated-patient',
                suffix = comp )

    cat("\n\n")
  }
  
}
```







## Dot plots simplified (complete)

```{r gse-dotplots-simplified, echo=FALSE, results='asis'}
knitr::opts_knit$set(
  base.dir = "docs"
  )
root_section_number <- 3
section_n <- 1

comp <- names(gse_list_merged)[1]
for(comp in names(gse_list_merged)) {
  h_section_number <- paste(root_section_number, section_n, sep=".")
  cat("<h3> <span class=\"header-section-number\">",h_section_number,"</span>", comp, "</h3>\n\n") #Cause results='asis',we must use raw HTML
  section_n <- section_n + 1

  x <- gse_list_merged[[comp]] %>% 
    arrange(desc(n_up), desc(avg_nes)) %>% 
    mutate(
      is.sign = ifelse(p.adjust < 0.05, 'yes', 'no'),
      log10P = -log10(p.adjust),
      Description = clean_msigdb_names(Description),
      Description = fct_reorder(Description, avg_nes)
    )
  if(nrow(x) == 0) {
    cat("No gene sets significant in at least two models.\n\n")
  } else {
    xlim <- c(-max(abs(x$NES)), max(abs(x$NES)))
    
    x_pdac3 <-  x %>% 
      filter(model == 'PDAC-3')
    x_patient <-  x %>% 
      filter(model == 'Patient')
    
    for(i in levels(x$Description)) {
      if(!i %in% x_pdac3$Description ) {
        x_pdac3 %<>% add_row(
          ID = x %>% filter(Description == i) %>% pull(ID) %>% unique(),
          Description = i,
          NES = 0,
          p.adjust = 1,
          model = 'PDAC-3',
          gene_set_group = x %>% filter(Description == i) %>% pull(gene_set_group) %>% unique(),
          n_significant = 0,
          n_up = 0,
          avg_fdr = 1,
          avg_nes = 0,
          log10P = 0,
          is.sign = 'no'
        )
      }
      if(!i %in% x_patient$Description ) {
        x_patient %<>% add_row(
          ID = x %>% filter(Description == i) %>% pull(ID) %>% unique(),
          Description = i,
          NES = 0,
          p.adjust = 1,
          model = 'Patient',
          gene_set_group = x %>% filter(Description == i) %>% pull(gene_set_group) %>% unique(),
          n_significant = 0,
          n_up = 0,
          avg_fdr = 1,
          avg_nes = 0,
          log10P = 0,
          is.sign = 'no'
        )
      }
    }
    
    # PDAC3: Select representative for each gene set group and model, to avoid redundancy in the plot.
    x_median <- x_pdac3 %>% 
      group_by(gene_set_group) %>% 
      summarise(
        group_n = n(),
        group_nes = median(NES)
      ) %>% 
      mutate(
        group_sign = ifelse(group_nes > 0, 'activated', 'supressed')
      ) %>% 
      arrange(group_nes)
    x_pdac3 <- x_pdac3 %>% 
      left_join(x_median) %>% 
      filter(.sign == group_sign) %>% 
      group_by(Group) %>%
      slice_min(p.adjust, n = 1, with_ties = FALSE)
    
    # Patient: Select representative for each gene set group and model, to avoid redundancy in the plot.
    x_median <- x_patient %>% 
      group_by(gene_set_group) %>% 
      summarise(
        group_n = n(),
        group_nes = median(NES)
      ) %>% 
      mutate(
        group_sign = ifelse(group_nes > 0, 'activated', 'supressed')
      ) %>% 
      arrange(group_nes)
    x_patient<- x_patient %>% 
      left_join(x_median) %>% 
      filter(.sign == group_sign) %>% 
      group_by(Group) %>%
      slice_min(p.adjust, n = 1, with_ties = FALSE)

    
    # Refactor Description
    group_levels <- x_pdac3 %>% arrange(NES) %>% pull(Group)
    x_pdac3 %<>% 
      mutate(
        gene_set_group = factor(gene_set_group, levels = group_levels)
        )
    x_patient %<>% 
      mutate(
        gene_set_group = factor(gene_set_group, levels = group_levels)
        )      
    
    # Generate plots
    res_1 <- x_pdac3 %>% 
      ggplot(aes(x = NES, y = gene_set_group, size = log10P, fill = NES)) +
      # geom_point(pch = 21, aes(color = is.sign)) +
      geom_point(pch = 21, stroke = linewidth_default/2) +
      facet_grid(cols = vars(model), scales = 'free_y', space = 'free_y') +
      scale_y_discrete(labels =  default_labeller(80)) +
      geom_vline(xintercept = 0, lty = 3) +
      labs(
        y = '',
        size = expression(paste("lo", g[10],"(Adjusted P value)")),
        color = 'FDR < 0.05'
        # caption = 'Gene sets with FDR<0.05 in at least 2 models',
      ) +
      scale_size_continuous(range=c(0.8, 2)) +
      scale_fill_gradient2(trans = "reverse") +
      scale_color_manual(values = c(no = 'white', yes = 'black')) +
      # scale_color_viridis_c() +
      scale_x_continuous(expand = expansion(mult = c(0.20, 0.20)),
                         limits = xlim) +
      theme(
        panel.border = element_rect(color = "black", fill = NA, linewidth = linewidth_default),
        strip.text.y = element_text(angle = 0, hjust = 0),
        
        axis.line = element_blank(),
        axis.ticks = element_line(color = "black", linewidth = linewidth_default),
        axis.text.y = element_text(size = 4),
        axis.text.x = element_text(size = 4),
        
        legend.text = element_text(size = 4), 
        legend.title = element_text(size = 4),
        legend.key.width = unit(0.2, "cm"),
        panel.spacing.y = unit(0.08, "lines")
      )
    
    res_2 <- x_patient %>% 
      ggplot(aes(x = NES, y = gene_set_group, size = log10P, fill = NES)) +
      # geom_point(pch = 21, aes(color = is.sign)) +
      geom_point(pch = 21, stroke = linewidth_default/2) +
      facet_grid(cols = vars(model), scales = 'free_y', space = 'free_y') +
      scale_y_discrete(labels =  default_labeller(80)) +
      geom_vline(xintercept = 0, lty = 3) +
      labs(
        y = '',
        size = expression(paste("lo", g[10],"(Adjusted P value)")),
        color = 'FDR < 0.05'
        # caption = 'Gene sets with FDR<0.05 in at least 2 models',
      ) +
      scale_size_continuous(range=c(0.8, 2)) +
      scale_fill_gradient2(trans = "reverse") +
      scale_color_manual(values = c(no = 'white', yes = 'black')) +
      # scale_color_viridis_c() +
      scale_x_continuous(expand = expansion(mult = c(0.20, 0.20)),
                         limits = xlim) +
      theme(
        panel.border = element_rect(color = "black", fill = NA, linewidth = linewidth_default),
        strip.text.y = element_text(angle = 0, hjust = 0),
        
        axis.line = element_blank(),
        axis.ticks = element_line(color = "black", linewidth = linewidth_default),
        axis.text.y = element_text(size = 4),
        axis.text.x = element_text(size = 4),
        
        legend.text = element_text(size = 4), 
        legend.title = element_text(size = 4),
        legend.key.width = unit(0.2, "cm"),
        panel.spacing.y = unit(0.08, "lines")
      )
    
    n_gsets <- x_patient$Group %>% unique() %>% length()
    # use_width <-  7.2 + n_gsets * 0.02
    use_width <- 2.2
    # use_asp <-  0.318 + n_gsets * 0.017
    # use_asp <-  n_gsets * 0.05
    use_asp <-  n_gsets * 0.051
    use_height <- use_width * use_asp
    

    subchunkify(res_1,
                fig.width = use_width,
                fig.height = use_height,
                dpi = 600,
                prefix = 'gse-dotplot-simplified-pdac3',
                suffix = comp )
    cat("\n\n")
    subchunkify(res_2,
                fig.width = use_width,
                fig.height = use_height,
                dpi = 600,
                prefix = 'gse-dotplot-simplified-patient',
                suffix = comp )   
    
    cat("\n\n")
  }
  
}
```


## Dot plots simplified (unamed)

```{r gse-dotplots-simplified-unnamed, echo=FALSE, results='asis'}
knitr::opts_knit$set(
  base.dir = "docs"
  )
root_section_number <- 4
section_n <- 1

comp <- names(gse_list_merged)[1]
for(comp in names(gse_list_merged)) {
  h_section_number <- paste(root_section_number, section_n, sep=".")
  cat("<h3> <span class=\"header-section-number\">",h_section_number,"</span>", comp, "</h3>\n\n") #Cause results='asis',we must use raw HTML
  section_n <- section_n + 1
  x <- gse_list_merged[[comp]] %>% 
    arrange(desc(n_up), desc(avg_nes)) %>% 
    mutate(
      is.sign = ifelse(p.adjust < 0.05, 'yes', 'no'),
      log10P = -log10(p.adjust),
      Description = clean_msigdb_names(Description),
      Description = fct_reorder(Description, avg_nes)
    )
  if(nrow(x) == 0) {
    cat("No gene sets significant in at least two models.\n\n")
  } else {

    # Remove redundancy
    x %<>% 
      group_by(model, gene_set_group, Description) %>% 
      slice_min(avg_fdr, n = 1, with_ties = FALSE)
    
    xlim <- c(-max(abs(x$NES)), max(abs(x$NES)))
    
    x_pdac3 <-  x %>% 
      ungroup() %>%
      filter(model == 'PDAC-3') %>% 
      mutate(model = as.character(model))
    x_patient <-  x %>% 
      ungroup() %>%
      filter(model == 'Patient') %>% 
      mutate(model = as.character(model))
    
    
    for(i in levels(x$Description)) {
      if(!i %in% x_pdac3$Description ) {
        x_pdac3 %<>% add_row(
          ID = x %>% filter(Description == i) %>% pull(ID) %>% unique(),
          Description = i,
          NES = 0,
          p.adjust = 1,
          model = 'PDAC-3',
          gene_set_group = x %>% filter(Description == i) %>% pull(gene_set_group) %>% unique(),
          n_significant = 0,
          n_up = 0,
          avg_fdr = 1,
          avg_nes = 0,
          log10P = 0,
          is.sign = 'no'
        )
      }
      if(!i %in% x_patient$Description ) {
        x_patient %<>% add_row(
          ID = x %>% filter(Description == i) %>% pull(ID) %>% unique(),
          Description = i,
          NES = 0,
          p.adjust = 1,
          model = 'Patient',
          gene_set_group = x %>% filter(Description == i) %>% pull(gene_set_group) %>% unique(),
          n_significant = 0,
          n_up = 0,
          avg_fdr = 1,
          avg_nes = 0,
          log10P = 0,
          is.sign = 'no'
        )
      }
    }
    
    
    # PDAC3: Select representative for each gene set group and model, to avoid redundancy in the plot.
    x_median <- x_pdac3 %>% 
      group_by(gene_set_group) %>% 
      summarise(
        group_n = n(),
        group_nes = median(NES)
      ) %>% 
      mutate(
        group_sign = ifelse(group_nes > 0, 'activated', 'supressed')
      ) %>% 
      arrange(group_nes)
    x_pdac3 <- x_pdac3 %>% 
      left_join(x_median) %>% 
      filter(.sign == group_sign) %>% 
      group_by(Group) %>%
      slice_min(p.adjust, n = 1, with_ties = FALSE)
    
    # Patient: Select representative for each gene set group and model, to avoid redundancy in the plot.
    x_median <- x_patient %>% 
      group_by(gene_set_group) %>% 
      summarise(
        group_n = n(),
        group_nes = median(NES)
      ) %>% 
      mutate(
        group_sign = ifelse(group_nes > 0, 'activated', 'supressed')
      ) %>% 
      arrange(group_nes)
    x_patient<- x_patient %>% 
      left_join(x_median) %>% 
      filter(.sign == group_sign) %>% 
      group_by(Group) %>%
      slice_min(p.adjust, n = 1, with_ties = FALSE)

    
    # Refactor Description
    group_levels <- x_pdac3 %>% arrange(NES) %>% pull(Group)
    x_pdac3 %<>% 
      mutate(
        gene_set_group = factor(gene_set_group, levels = group_levels)
        )
    x_patient %<>% 
      mutate(
        gene_set_group = factor(gene_set_group, levels = group_levels)
        )      
    
    
    res_1 <- x_pdac3 %>%
      ggplot(aes(x = NES, y = gene_set_group, size = log10P, fill = NES)) +
      geom_vline(xintercept = 0, lty = 3) +
      geom_point(pch = 21, stroke = linewidth_default/2) +
      # geom_point(data = x_pdac3 %>% filter(is.sign == 'yes'), pch = 21, aes(color = is.sign)) +
      # geom_point(data = x_pdac3 %>% filter(is.sign == 'no'), pch = 21, stroke = 0) +
      facet_grid(cols = vars(model), scales = 'free_y', space = 'free_y') +
      labs(
        x = '',
        y = '',
        # title = 'PDAC-3',
        size = expression(paste("lo", g[10],"(FDR)")),
        color = 'FDR\n< 0.05'
      ) +
      scale_y_discrete(labels =  default_labeller(80)) +
      scale_x_continuous(expand = expansion(mult = c(0.20, 0.20)),
                         limits = xlim) +
      scale_size_continuous(range=c(0.5, 1.1)) +
      scale_fill_gradient2(trans = "reverse") +
      scale_color_manual(values = c(no = 'white', yes = 'black')) +
      theme(
        panel.border = element_rect(color = "black", fill = NA, linewidth = linewidth_default),
        strip.text.y = element_blank(),
        strip.text.x = element_blank(),
        
        axis.line = element_blank(),
        axis.ticks = element_line(color = "black", linewidth = linewidth_default/2),
        axis.text.y = element_blank(),
        axis.text.x = element_blank(),
        
        legend.text = element_text(size = 3), 
        legend.title = element_text(size = 3),
        legend.key.width = unit(0.1, "cm"),
        legend.key.height = unit(0.1, "cm"),
        
        plot.margin = unit(c(0.0217, 0.0217, 0.0217, 0.0217), "inches"),
        panel.spacing.y = unit(0.08, "lines")
      )

    res_2 <- x_patient %>%
      ggplot(aes(x = NES, y = gene_set_group, size = log10P, fill = NES)) +
      geom_vline(xintercept = 0, lty = 3) +
      geom_point(pch = 21, stroke = linewidth_default/2) +
      # geom_point(data = x_patient %>% filter(is.sign == 'yes'), pch = 21, aes(color = is.sign)) +
      # geom_point(data = x_patient %>% filter(is.sign == 'no'), pch = 21, stroke = 0) +
      facet_grid(cols = vars(model), scales = 'free_y', space = 'free_y') +
      labs(
        x = '',
        y = '',
        # title = 'Patient',
        size = expression(paste("lo", g[10],"(FDR)")),
        color = 'FDR\n< 0.05'
      ) +
      scale_y_discrete(labels =  default_labeller(80)) +
      scale_x_continuous(expand = expansion(mult = c(0.20, 0.20)),
                         limits = xlim) +
      scale_size_continuous(range=c(0.5, 1.1)) +
      scale_fill_gradient2(trans = "reverse") +
      scale_color_manual(values = c(no = 'white', yes = 'black')) +
      theme(
        panel.border = element_rect(color = "black", 
                                    fill = NA, 
                                    linewidth = linewidth_default),
        strip.text.y = element_blank(),
        strip.text.x = element_blank(),
        
        axis.line = element_blank(),
        axis.ticks = element_line(color = "black", 
                                  linewidth = linewidth_default/2),
        axis.text.y = element_blank(),
        axis.text.x = element_blank(),
        
        legend.text = element_text(size = 3), 
        legend.title = element_text(size = 3),
        legend.key.width = unit(0.1, "cm"),
        legend.key.height = unit(0.1, "cm"),
        
        plot.margin = unit(c(0.0217, 0.0217, 0.0217, 0.0217), "inches"),
        panel.spacing.y = unit(0.08, "lines")
      )
        
        
    n_gsets <- x_patient$ID %>% unique() %>% length()
    n_gsgroups <- x_patient$gene_set_group %>% unique %>% length
    
    
    
    # Calculate width = left margin + total width of panels + total width of spacing + right margin. Everything scaled by conversion factor to get final width in inches. Conversion factor calculater with a rule in the screen
    # nmodels <- x$model %>% unique() %>% length()
    nmodels <- 1
    
    # Calculate width: left margin + tickmarks + total width of panels + total width of spacing + right margin
    # use_width <- 0.1165 + 0.0173 + (0.18898*nmodels) + (0.035*(nmodels - 1))+ 0.0217
    use_width <- 0.1165 + 0.0173 + (0.18898*nmodels) + (0.035*(nmodels - 1))+ 0.0217
    
    # Calculate height: top margin + tickmarks + bottom margin and legend
    use_height <- 0.0217 + 0.0217 + (n_gsets*1.05) * 0.0492 + (0.015*(n_gsgroups - 1)) + 0.4984
    
    
    # Print PDAC-3
    subchunkify(res_1 +
                  guides(
                    size  = guide_legend(direction = "vertical",
                                         keyheight = unit(0.0289, "inches"),
                                         keywidth = unit(0.0289, "inches")
                                         ),
                    color = "none",
                    fill  = "none"
                    # color = guide_legend(direction = "vertical"),
                    # fill  = guide_colourbar(direction = "vertical")
                  ) +
                  theme(
                    legend.position = "bottom",
                    legend.box = "horizontal"
                  ),
                fig.width = use_width,
                fig.height = use_height,
                dpi = 600,
                prefix = 'gse-dotplot-simplified-unnamed-pdac3',
                suffix = comp )
    
    cat("\n\n")
    
    # Print legend
    res_legend <- get_legend(
      res_1 +
        guides(
          color = guide_legend(direction = "vertical"),
          size  = guide_legend(direction = "vertical"),
          fill  = guide_colourbar(direction = "vertical")
        ) +
        theme(
          legend.position = "bottom",
          legend.box = "horizontal"
        )
    )
    subchunkify(plot_grid(res_legend),
                fig.width = 1.8,
                fig.height = 1,
                dpi = 600,
                prefix = 'gse-dotplot-legend-simplified-unnamed-pdac3',
                suffix = comp )

    cat("\n\n")
    
      
    # Print Patient
    subchunkify(res_2 +
                  guides(
                    size  = guide_legend(direction = "vertical",
                                         keyheight = unit(0.0289, "inches"),
                                         keywidth = unit(0.0289, "inches")
                                         ),
                    color = "none",
                    fill  = "none"
                    # color = guide_legend(direction = "vertical"),
                    # fill  = guide_colourbar(direction = "vertical")
                  ) +
                  theme(
                    legend.position = "bottom",
                    legend.box = "horizontal"
                  ),
                fig.width = use_width,
                fig.height = use_height,
                dpi = 600,
                prefix = 'gse-dotplot-simplified-unnamed-patient',
                suffix = comp )
    
    cat("\n\n")
    
    # Print legend
    res_legend <- get_legend(
      res_2 +
        guides(
          color = guide_legend(direction = "vertical"),
          size  = guide_legend(direction = "vertical"),
          fill  = guide_colourbar(direction = "vertical")
        ) +
        theme(
          legend.position = "bottom",
          legend.box = "horizontal"
        )
    )
    subchunkify(plot_grid(res_legend),
                fig.width = 1.8,
                fig.height = 1,
                dpi = 600,
                prefix = 'gse-dotplot-legend-simplified-unnamed-patient',
                suffix = comp )

    cat("\n\n")
  }
  
}
```








## Download results
```{r gse-table-download, echo=FALSE}
library(openxlsx)
comp <- names(gse_list_merged)[1]

# File name summary
rmd_file <- current_input()
if(is.null(rmd_file))
  rmd_file <- 'tmp'

file_xlsx <- file.path('./docs/file', 
                           rmd_file, 'gse_integration.xlsx'
                           )

dir.create(dirname(file_xlsx), recursive = TRUE, showWarnings = FALSE)
  
wb <- createWorkbook()

# Add a table with unique pathways
gset_unique <- foreach(i=gse_list_merged) %do% {
  i %>% pull(ID)
} %>% unlist %>% unique %>% data.frame() %>% set_names(c('ID')) %>% mutate(Group=NA) 
addWorksheet(wb, 'Unique_gene_sets')
writeData(wb, 'Unique_gene_sets', gset_unique)

# Add a table for each comparison
for(comp in names(gse_list_merged)) {
  x <- gse_list_merged[[comp]] %>% 
    arrange(desc(n_up), desc(avg_nes)) %>% 
    mutate(
      is.sign = ifelse(p.adjust < 0.05, 'yes', 'no'),
      log10P = -log10(p.adjust),
      Description = clean_msigdb_names(Description),
      Description = fct_reorder(Description, avg_nes)
    )
  addWorksheet(wb, comp)
  writeData(wb, comp, x)
}

saveWorkbook(wb, file_xlsx, TRUE)
 file_xlsx <- gsub("docs/", "", file_xlsx)
cat(paste0("- [**GSEA integration**](", file_xlsx, ")"), '\n\n')

```





