---
title: "PDAC-3 spatial transcriptomics"
subtitle: "Barcodes analysis - 3D analysis"
author: "Francesc Castro-Giner"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
params:
  date: '`r format(Sys.Date(), "%B %d, %Y")`'
  donor: PDAC-3
  output_dir: ./output/rnaseq-spatial/pdac3
  seurat_dir: ./data/rnaseq-spatial/pdac3
---



## Configuration

Load libraries, additional functions and data

Setup environment
```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(results='asis', echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.align = 'center', fig.width = 3.5, fig.asp = 0.618, dpi = 600, dev = c("png", "pdf"), fig.showtext = FALSE, engine.opts = list(bash = "-l"))

options(stringsAsFactors = FALSE)

use_seed <- 1100101
set.seed(use_seed)

dir.create(params$output_dir, recursive = TRUE, showWarnings = FALSE)
```

Load packages
```{r load-libraries}
library(tidyverse)
library(foreach)
# library(DT)
library(knitr)
# library(patchwork)
# library(kableExtra)
# # library(cowplot)
# # # library(colorblindr)
library(RColorBrewer)
# library(ggbeeswarm)
# # library(ggrepel)
library(magrittr)
# library(UpSetR)
# library(GGally)
# library(openxlsx)
library(ggpubr)
# library(rstatix)
library(arsenal)
# #
library(Seurat)
library(SummarizedExperiment)
library(semla)
# library(edgeR)
# library(DESeq2)
# library(scuttle)
# library(scran)
# library(pheatmap)
# library(PCAtools)
# library(ComplexHeatmap)
# library(circlize)

library(plotly)
library(dbscan)
```

Load ggplot theme
```{r ggplot-theme}
source("./configuration/rmarkdown/ggplot_theme.R")
```

Set color palettes
```{r, echo = FALSE}
source("./configuration/rmarkdown/color_palettes.R")
```

Clean files generated in previous runs
```{r clean-files}
rmd_file <- current_input()
if(!is.null(rmd_file)) {
  figures_dir <- file.path('./docs/figure',rmd_file)
  if(dir.exists(figures_dir)) {
    unlink(file.path(figures_dir, "*"))
  }
}
```

Load custom functions
```{r load-functions}
source('./code/R-functions/semla_functions.r')
```


## Data processing

### Load initial data

Load human seurat data
```{r dw-load-seurat}
seu <- readRDS(file.path(params$seurat_dir, 'seurat_hs/seurat_merged_semla_geneNMF.rds'))

# Change rownames
rownames(seu@assays[['Spatial']]) <-  gsub('GRCh38-', '',rownames(seu@assays[['Spatial']]))

i <- 'SCT'
rownames(seu@assays[[i]]@counts) <-  gsub('GRCh38-', '', rownames(seu@assays[[i]]@counts))
rownames(seu@assays[[i]]@data) <-  gsub('GRCh38-', '', rownames(seu@assays[[i]]@data))
rownames(seu@assays[[i]]@scale.data) <-  gsub('GRCh38-', '', rownames(seu@assays[[i]]@scale.data))
seu@assays[[i]]@var.features <- gsub('GRCh38-', '', seu@assays[[i]]@var.features)
```

List of individual seurat files with full features
```{r dw-list-seurat}
seurat_files <- list.files(file.path(params$seurat_dir, "seurat"), pattern = "S[0-9]+.*.rds", full.names = TRUE)
```

Load barcode frequencies
```{r dw-bc-ctc-spot-freq-combined-load}
bc_spot_freq <- readRDS(file.path(params$output_dir, 'seurat/bc_spot_freq.rds'))
bc_freq <- readRDS(file.path(params$output_dir, 'seurat/bc_freq.rds'))
```

Load final data from barcode analysis report
```{r dw-load-processed-data}
bc_mat <- readRDS(file.path(params$seurat_dir, "seurat_bc", "bc_matrix.rds"))
bc_freq <- readRDS(file.path(params$output_dir, 'seurat/bc_freq.rds'))
seu_full_list <- readRDS(file.path(params$seurat_dir, "seurat", 'seurat_merged_list.rds'))
seu_bc <- readRDS(file.path(params$output_dir, 'seurat_hs/seurat_bc_hotspots.rds'))

# Load hotspot data
hotspot_bc <- readRDS(file.path(params$output_dir, 'seurat_bc/semla_hotspots.rds'))
caf_hotspots <- readRDS(file.path(params$output_dir, "hotspots_tme", "spots_caf_hotspot_2.rds"))
caf_hotspot_r_dist <- readRDS(file.path(params$output_dir, 'caf/caf_hotspot_r_dist.rds'))
```

Add annotation to Seurat object
```{r dw-add-annotation-seurat}
# Add CAF rdist
seu[[]] %<>% 
  rownames_to_column('barcode') %>% 
  left_join(caf_hotspot_r_dist, by = 'barcode') %>%
  column_to_rownames('barcode')

# Add CAF spots: Filtering criteria must follow the same approach as in chunk::caf-distance-conf
use_bc <- caf_hotspots %>%
  filter(hotspot != 'singleton') %>%
  filter(hotspot_size >= 3) %>%
  pull(sample_spot)

seu$caf_hotspots <- ifelse(colnames(seu) %in% use_bc, 'yes', NA)

# Add hotspots defined based on barcodes. We use the same criteria as in chunk::dw-bc-hotspots-to-seurat
st_spots <- hotspot_bc$st_specific %>% 
  filter(hotspot != 'singleton') %>% 
  filter(hotspot_size >= 5) %>%
  pull(sample_spot)
ctc_spots <- hotspot_bc$ctc_top %>% 
  filter(hotspot != 'singleton') %>% 
  filter(hotspot_size >= 5) %>%
  pull(sample_spot)

seu[[]] %<>% 
  rownames_to_column('sample_spot') %>%
  mutate(
    bc_group = ifelse(sample_spot %in% st_spots, 'PT', ifelse(sample_spot %in% ctc_spots, 'CTC', 'other'))
  ) %>% 
  column_to_rownames('sample_spot')


# Change BC groups levels
seu_bc[[]] %<>% 
  mutate(
    bc_group = ifelse(bc_group == 'ST', 'PT ref', bc_group),
    bc_group_cluster = ifelse(bc_group_cluster == 'ST', 'PT ref', as.character(bc_group_cluster)),
    bc_group_cluster = factor(bc_group_cluster, levels = c('PT ref', 'CTC1', 'CTC2'))
  )

```

Create a Seurat object combining CAF hotspots and BC group annotation. We will use this Seurat object for the spatial distribution analysis.
```{r dw-seurat-bc-caf}
# Seurat object combining CAF and BC group annotation
seu_bc_caf <- seu

# Add bc_group_cluster information
bc_group_cluster_spots <- seu_bc[[]] %>% rownames_to_column('sample_spot') %>% 
  select(sample_spot, bc_group_cluster, clusters)

# Rename groups
seu_bc_caf[[]] %<>% 
  rownames_to_column('sample_spot') %>% 
  left_join(bc_group_cluster_spots) %>% 
  mutate(
    group_hotspots = paste(bc_group, caf_hotspots, sep = '_'),
    group_hotspots = gsub("_NA", "", group_hotspots),
    group_hotspots = gsub("NA_", "", group_hotspots),
    group_hotspots = gsub("yes", "CAF", group_hotspots),
    group_hotspots = gsub("other_CAF", "CAF", group_hotspots),
    
    group_hotspots_extended = paste(bc_group_cluster, caf_hotspots, sep = '_'),
    group_hotspots_extended = gsub("_NA", "", group_hotspots_extended),
    group_hotspots_extended = gsub("NA_", "", group_hotspots_extended),
    group_hotspots_extended = gsub("yes", "CAF", group_hotspots_extended),
    group_hotspots_extended = ifelse(group_hotspots_extended == 'NA', 'other', group_hotspots_extended)
  ) %>% 
  column_to_rownames('sample_spot')

# Remove spots with no annotation
seu_bc_caf <- seu_bc_caf[,seu_bc_caf$group_hotspots !='other']
```

Modify coordinates
```{r dw-add-xy-um}
use_seu <- seu

# Get pxl to microns conversion factor. We will use the median of the pixel spot diameter across the different slices to calculate the conversion factor, and we will use the physical spot diameter of 55 microns provided by 10x as reference for the conversion. 
physical_spot_um <- 55
use_coords_list <- semla:::.get_coords_list(use_seu)
spatnet_list <- lapply(use_coords_list, GetSpatialNetwork, maxDist = 50)
i <- spatnet_list[[1]]

um_per_pxl <- foreach(i = spatnet_list) %do% {
  pixel_spot_diameter <- i[[1]] %>% group_by(from) %>% summarise(min_distance = min(distance)) %>% pull(min_distance) %>% median
  physical_spot_um / pixel_spot_diameter
} %>% unlist %>% median


# Convert_coordinates
xy_coord <- use_seu@tools$Staffli@meta_data %>%  
  select(barcode, pxl_col_in_fullres_transformed, pxl_row_in_fullres_transformed) %>%
  set_names(c('barcode', 'x', 'y')) %>% 
  mutate(
    x = x * um_per_pxl,
    y = y * um_per_pxl
  )

xyz_coord <- use_seu[[]] %>% 
  mutate(
    z = z_um
  ) %>% 
  rownames_to_column('barcode') %>%
  left_join(xy_coord, by = 'barcode') %>% 
  dplyr::select(barcode, x, y ,z)

# Add coordinates to seurat objects
seu[[]] %<>% 
  rownames_to_column('barcode') %>%
  # select(-z) %>%
  left_join(xyz_coord, by = 'barcode') %>%
  column_to_rownames('barcode')
seu_bc[[]] %<>% 
  rownames_to_column('barcode') %>%
  # select(-z) %>%
  left_join(xyz_coord, by = 'barcode') %>%
  column_to_rownames('barcode')
seu_bc_caf[[]] %<>% 
  rownames_to_column('barcode') %>%
  # select(-z) %>%
  left_join(xyz_coord, by = 'barcode') %>%
  column_to_rownames('barcode')

```



## 3D visualization
```{r bc-group-3d}
use_data <- seu[[]] %>% 
  rownames_to_column('barcode') %>%
  left_join(seu_bc[[]] %>% rownames_to_column('barcode') %>% select(barcode, bc_group_cluster), by = 'barcode')

point_size <- 2
use_palette <- bc_group_cluster_palette
color_var <- 'bc_group_cluster'

res_plot <- plot_ly()

# Add NA points
res_plot %<>%
  add_markers(
    data = use_data %>% filter(is.na(.[[color_var]])),
    x = ~x, y = ~y, z = ~z,
    name = 'Other',
    marker = list(
      color = 'black',
      size = point_size,
      opacity = 0.02
    )
  )

# Add !NA points
for(i in levels(use_data[[color_var]])) {
  res_plot %<>% 
    add_markers(
      data = use_data %>% filter(.[[color_var]] == i),
      x = ~x, y = ~y, z = ~z,
      name = i,
      marker = list(
        color = use_palette[[i]],
        size = point_size,
        opacity = 0.5
      )
    ) 
}


res_plot %>% 
  layout(legend= list(itemsizing='constant'), # keep the legend makers the same size
         scene = list(
           xaxis = list(showgrid = FALSE, zeroline = FALSE,
                        showticklabels=FALSE, title = ''),
           yaxis = list(showgrid = FALSE, zeroline = FALSE,
                        showticklabels=FALSE, title = ''),
           zaxis = list(showgrid = FALSE, zeroline = FALSE,
                        showticklabels=FALSE, title = ''),
           camera = list(
             eye = list(x = 1.1, y = 1.1, z = 1.1)
           )
         )
  ) 
```




## Hb synthesis score

Values with an score <0.1 have an increased transparency.

```{r rbc-score-3d}

hb_ucell <- readRDS(file.path(params$seurat_dir, 'seurat_mm/ucell_hb_synthesis.rds'))

# Add to seurat objects
use_data <- seu[[]] %>% 
  rownames_to_column('sample_barcode') %>%
  left_join(hb_ucell, by = 'sample_barcode')

point_size <- 2
res_plot <- plot_ly()

# Add 0 points
res_plot %<>% 
  add_markers(
    data = use_data %>% filter(Hemoglobin_synthesis_UCell < 0.1),
    x = ~x, y = ~y, z = ~z,
    color = ~Hemoglobin_synthesis_UCell, 
    colors = rev(brewer.pal(11, "Spectral")),
    name = 'Hb-',
    marker = list(
      opacity = 0.1,
      size = point_size
    )
  )

# Add exprs > 0 points
res_plot %<>% 
  add_markers(
    data = use_data %>% filter(Hemoglobin_synthesis_UCell >= 0.1),
    x = ~x, y = ~y, z = ~z,
    color = ~Hemoglobin_synthesis_UCell, 
    colors = rev(brewer.pal(11, "Spectral")),
    opacity = ~Hemoglobin_synthesis_UCell,
    marker = list(
      # opacity = 0.5,
      size = 2
    )
  )

res_plot %>% 
  layout(
    title = 'Hemoglobin synthesis score',
    scene = list(
      xaxis = list(showgrid = FALSE, zeroline = FALSE,
                   showticklabels=FALSE, title = ''),
      yaxis = list(showgrid = FALSE, zeroline = FALSE,
                   showticklabels=FALSE, title = ''),
      zaxis = list(showgrid = FALSE, zeroline = FALSE,
                   showticklabels=FALSE, title = '')
    )
  )
```
