---
title: "PDAC-3 spatial transcriptomics"
subtitle: "Differential expression of spots by CAF distance using human reads"
author: "Francesc Castro-Giner"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
params:
  date: '`r format(Sys.Date(), "%B %d, %Y")`'
  donor: PDAC-3
  output_dir: ./output/rnaseq-spatial/pdac3
  seurat_dir: ./data/rnaseq-spatial/pdac3
  min_counts: 5
  min_present_prop: 0.25
  msigdb_dir: ./data/resources/MSigDB
  msigdb_vs: v2023.1.Hs
  ncores: 4
---



## Configuration

Load libraries, additional functions and data

Setup environment
```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(results='asis', echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.align = 'center', dpi = 600, dev = c("png", "pdf"), fig.showtext = FALSE, engine.opts = list(bash = "-l"))

options(stringsAsFactors = FALSE)

use_seed <- 1100101
set.seed(use_seed)

dir.create(params$output_dir, recursive = TRUE, showWarnings = FALSE)
```

Load packages
```{r load-libraries}
library(tidyverse)
library(foreach)
library(knitr)
library(patchwork)
library(kableExtra)
library(RColorBrewer)
library(ggbeeswarm)
library(ggrepel)
library(magrittr)
library(UpSetR)
library(GGally)
library(openxlsx)
library(ggpubr)
library(rstatix)
library(arsenal)
library(Seurat)
library(SummarizedExperiment)
library(semla)
library(edgeR)
library(scran)
library(ComplexHeatmap)
library(circlize)
```

Load ggplot theme
```{r ggplot-theme}
source("./configuration/rmarkdown/ggplot_theme.R")
```

Set color palettes
```{r, echo = FALSE}
source("./configuration/rmarkdown/color_palettes.R")
```

Load custom functions
```{r load-functions}
source('./code/R-functions/semla_functions.r')
source('./code/R-functions/dge_report.r')
source('./code/R-functions/gse_omnibus.r')
source('./code/R-functions/gse_report.r')
source('./code/R-functions/subchunkify.R')
```

Load MSigDB gene sets
```{r load-msigdb}
gmt_files_symbols <- list(
  msigdb.c2.cp = paste0(params$msigdb_dir, '/', params$msigdb_vs, '/c2.cp.', params$msigdb_vs, '.symbols.gmt')
)
```

Clean files generated in previous runs
```{r clean-files}
rmd_file <- current_input()
if(!is.null(rmd_file)) {
  figures_dir <- file.path('./docs/figure',rmd_file)
  if(dir.exists(figures_dir)) {
    unlink(file.path(figures_dir, "*"))
  }
}
```


## Data processing 

### Load initial data

Load human seurat data
```{r dw-load-seurat}
seu <- readRDS(file.path(params$seurat_dir, 'seurat_hs/seurat_merged_semla_geneNMF.rds'))
```

### CAF distance

Here we will define the distance of each spot to the nearest CAF spot, which we will use later to define the distance of each barcode hotspot to CAFs. *CHECK DESCRIPTION: We will use the same approach as in the previous analysis, but we will use the spatial coordinates of the spots instead of the UMAP coordinates. We will also use a different threshold for defining CAF spots, which is based on the expression of a CAF marker gene, such as FAP or PDGFRB. We will define CAF spots as those that have at least 1 count of the marker gene in the spatial data. We will then calculate the distance of each spot to the nearest CAF spot using the `semla::GetSpatialNetwork` function, which calculates a spatial network based on the spatial coordinates of the spots and a distance threshold. We will use a distance threshold that is based on the median distance between spots in the spatial data, which we will calculate using the `semla::GetSpatialNetwork` function with a large distance threshold and then taking the median of the distances between spots.*

Warning: Data produced by **pdac3-visium-bc-reanalysis.Rmd**

Load dataset and add CAF distance to seurat object
```{r caf-distance-df-load}
caf_hotspot_r_dist <- readRDS(file.path(params$output_dir, 'caf/caf_hotspot_r_dist.rds'))

```



### Process seurat object
```{r dw-process-seurat, eval = FALSE}
seu <- PrepSCTFindMarkers(seu, assay = "SCT")


# Change rownames
rownames(seu@assays[['Spatial']]) <-  gsub('GRCh38-', '',rownames(seu@assays[['Spatial']]))

i <- 'SCT'
rownames(seu@assays[[i]]@counts) <-  gsub('GRCh38-', '', rownames(seu@assays[[i]]@counts))
rownames(seu@assays[[i]]@data) <-  gsub('GRCh38-', '', rownames(seu@assays[[i]]@data))
rownames(seu@assays[[i]]@scale.data) <-  gsub('GRCh38-', '', rownames(seu@assays[[i]]@scale.data))
seu@assays[[i]]@var.features <- gsub('GRCh38-', '', seu@assays[[i]]@var.features)

# Add distance to CAF spots
seu[[]] %<>% 
  rownames_to_column('barcode') %>% 
  left_join(caf_hotspot_r_dist, by = 'barcode') %>%
  column_to_rownames('barcode')
```





### Compare spots by CAF distance
```{r dw-findmarkers-caf-regions, eval = FALSE}
chunk_suffix <- 'caf_distance_100_1000'
seu[[]] <- seu[[]] %>% 
  mutate(
    caf_dist_cat = 'intermediate',
    caf_dist_cat = ifelse(
      r_dist_caf_hotspot_3d_transformed < 100, 'l100um',
      caf_dist_cat
    ),
    caf_dist_cat = ifelse(
      r_dist_caf_hotspot_3d_transformed > 1000, 'u1000um',
      caf_dist_cat
    ),
    caf_dist_cat = factor(caf_dist_cat)
  )

# use_comparisons <- combn(rev(levels(seu$caf_dist_cat)), 2, simplify = FALSE)
use_comparisons <- list(
  c('l100um', 'u1000um')
)


# All pairwise comparisons
use_seu <- seu
res_markers <- foreach(i = use_comparisons) %do% {
  spots_sample <- colnames(use_seu)[use_seu$caf_dist_cat == i[1]]
  spots_ref <- colnames(use_seu)[use_seu$caf_dist_cat == i[2]]
  
  de_markers <- FindMarkers(use_seu, 
                            ident.1 = spots_sample,
                            ident.2 = spots_ref,
                            logfc.threshold = 0.2,
                            min.pct = 0.01,
                            test.use = "wilcox")
  list(
    result = de_markers %>% rownames_to_column('gene'),
    ident.1 = spots_sample,
    ident.2 = spots_ref,
    ident.1.name = i[1],
    ident.2.name = i[2]
  )
}
names(res_markers) <- foreach(i = use_comparisons) %do% {
  paste0(i[1], '-over-', i[2])
} %>% unlist

saveRDS(res_markers, file.path(params$output_dir, 'seurat_hs', paste0('results_findMarkers_', chunk_suffix,'.rds')))

```


Perform functional enrichment for specific comparisons
```{r dw-findmarkers-gse_omnibus-comparisons, eval = FALSE}
chunk_suffix <- 'caf_distance_100_1000'
res_markers <- readRDS(file.path(params$output_dir, 'seurat_hs', paste0('results_findMarkers_', chunk_suffix,'.rds')))
# Run functional enrichment
library(org.Hs.eg.db)

dge_i <-  res_markers[[1]]

gse_list <- foreach(dge_i = res_markers) %do%{
  dge <- dge_i$result
  gse_res <- gse_omnibus(
    feature_names = dge$gene,
    p = dge$p_val_adj,
    fc = dge$avg_log2FC,
    fc_thrs = 0,
    gmt_files = gmt_files_symbols,

    run_enricher = FALSE,
    run_enrichGO = FALSE,
    run_enrichReactome = FALSE,

    run_GSEA = TRUE,
    run_gseGO = FALSE,
    run_gseReactome = FALSE,
    
    go_ontologies = c('BP', 'MF'),
    go_simplify_p = 0.05,
    go_simplify_cutoff = 0.7,
    
    args_enricher = list(qvalueCutoff = 1, 
                         minGSSize = 10, 
                         maxGSSize = 250, 
                         pvalueCutoff = 1),
    
    args_gse = list(minGSSize = 10, 
                    maxGSSize = 250, 
                    pvalueCutoff = 1, 
                    BPPARAM = BiocParallel::MulticoreParam(params$ncores)
    )
  )
  
  return(gse_res)
  
}
names(gse_list) <- names(res_markers)

saveRDS(gse_list, file.path(params$output_dir, 'seurat_hs', paste0('results_findMarkers_gse_', chunk_suffix,'.rds')))
```

## FindMarkers results

The tables of results can be downloaded using the following links:

```{r  write-dge-xlsx-comp}
chunk_suffix <- 'caf_distance_100_1000'
dge_list <- readRDS(file.path(params$output_dir, 'seurat_hs', paste0('results_findMarkers_', chunk_suffix,'.rds')))

# File name summary
rmd_file <- current_input()
if(is.null(rmd_file))
  rmd_file <- 'tmp'
file_xlsx <- file.path('./docs/file',
                       rmd_file, 
                       paste0('dge_', chunk_suffix, '.xlsx')
                       )

dir.create(dirname(file_xlsx), recursive = TRUE, showWarnings = FALSE)

# Generate workbook
wb <- createWorkbook()
for(i in names(dge_list)) {
  addWorksheet(wb, i)
  res <- dge_list[[i]]$result
  writeData(wb, i, res)
}
saveWorkbook(wb, file_xlsx, TRUE)

dge_file_xlsx <- file_xlsx
```

- [**FindMarkers for CAF distance categorised**](`r gsub("docs/", "" , dge_file_xlsx)`)



## GSEA for specific comparisons
```{r gsea-conf-comparisons}
chunk_suffix <- 'caf_distance_100_1000'
gse_list <- readRDS(file.path(params$output_dir, 'seurat_hs', paste0('results_findMarkers_gse_', chunk_suffix,'.rds')))

# Get gene set grouping 
gsets_grouping_file <- './data/external/pdac3_caf_distance_gsea_gsets_groups.xlsx' 
gsets_sheets <- readxl::excel_sheets(path = gsets_grouping_file)
gsets_groups <- foreach(i = gsets_sheets) %do% {
  readxl::read_xlsx(gsets_grouping_file, sheet = i)
}
names(gsets_groups) <- gsets_sheets
```


### Download results

<!-- Write GSEA results to xlsx file -->
```{r  write-gsea-xlsx-comparisons}
gse_collections <- list(
  GSEA = names(gmt_files_symbols)
  # gseGO = c("BP", "MF"),
  # gseReactome = 'gseReactome'
)
    
# File name summary
rmd_file <- current_input()
if(is.null(rmd_file))
  rmd_file <- 'tmp'

file_xlsx_list <- list()
for(sr in names(gse_collections)) {
  file_xlsx_list[[sr]] <- list()
  for(gc in gse_collections[[sr]]) {
    # Generate file
    file_xlsx <- file.path('./docs/file', 
                           rmd_file, 
                           paste0('gsea_', chunk_suffix, '_', tolower(sr), '_', tolower(gc), '.xlsx')
                           )
    file_xlsx_list[[sr]][[gc]] <- file_xlsx
    dir.create(dirname(file_xlsx), recursive = TRUE, showWarnings = FALSE)
    
    # Generate workbook
    wb <- createWorkbook()
    for(i in names(gse_list)) {
      addWorksheet(wb, i)
      if(gc != sr){
        if(!is.null(gse_list[[i]][[sr]][[gc]])) {
          res <- gse_list[[i]][[sr]][[gc]]@result
          writeData(wb, i, res)
        }
      }
      if(gc == sr){
        if(!is.null(gse_list[[i]][[sr]])) {
          res <- gse_list[[i]][[sr]]@result
          writeData(wb, i, res)
        }
      }
    }
    saveWorkbook(wb, file_xlsx, TRUE)
  }
}
```

<!-- Write links to download GSEA results -->
```{r write-ctc-gsea-xlsx-links-comparisons, echo = FALSE}
for(sr in names(file_xlsx_list)) {
  for(gc in names(file_xlsx_list[[sr]])) {
    file_xlsx <- gsub("docs/", "", file_xlsx_list[[sr]][[gc]])
    if(sr == 'GSEA') {
      cat(paste0("- [**Gene set enrichment analysis: ", gc, "**](", file_xlsx, ")"), '\n\n')
    }
    if(sr == 'gseGO') {
      cat(paste0("- [**Gene set enrichment analysis: GO:", gc, "**](", file_xlsx, ")"), '\n\n')
    }   
    if(sr == 'gseKEGG') {
      cat(paste0("- [**Gene set enrichment analysis: KEGG**](", file_xlsx, ")"), '\n\n')
    }  
    if(sr == 'gseReactome') {
      cat(paste0("- [**Gene set enrichment analysis: Reactome**](", file_xlsx, ")"), '\n\n')
    }
  }
}
```


### Dotplots
<!-- # Subchunkify dotplots -->
<!-- ## chunk options need to be set to echo=FALSE, results='asis' -->
<!-- ## Cause results='asis',we must use raw HTML to print text and headings -->
```{r gse-dotplot-comparisons, echo=FALSE, results='asis'}
knitr::opts_knit$set(
  base.dir = "docs"
  )

gse_collections <- list(
  GSEA = names(gmt_files_symbols)
  # gseGO = c("BP", "MF"),
  # gseReactome = 'gseReactome'
)

use_prefix <- 'gse-dotplot'
use_linewidth <- 1/2.141959
root_section_number <- 4.2
section_n <- 1
subsection_n <- 1

i <- names(gse_list)[1]
sr <- names(gse_collections)[1]
gc <- gse_collections[[sr]][1]

for(i in names(gse_list)) {
  h_section_number <- paste(root_section_number, section_n, sep=".")
  cat("<h4> <span class=\"header-section-number\">",h_section_number,"</span>Comparison", i, "</h4>\n\n") #Caue results='asis',we must use raw HTML
  section_n <- section_n + 1
  for(sr in names(gse_collections)) {
    for(gc in gse_collections[[sr]]) {
      if(sr == 'GSEA') {
        x <- gse_list[[i]][[sr]][[gc]]@result %>% 
          filter(p.adjust <= 0.05) %>% 
          mutate(Description = clean_msigdb_names(Description))
      }
      if(sr == 'gseGO') {
        x <- gse_list[[i]][[sr]][[gc]]@result %>% 
          filter(p.adjust <= 0.05 & simplify)
      }
      if(sr %in% c('gseKEGG', 'gseReactome')) {
        x <- gse_list[[i]][[sr]]@result %>% 
          filter(p.adjust <= 0.05)
      }
      if(nrow(x) > 0) {
        h_subsection_number <- paste(h_section_number, subsection_n, sep=".")
        cat("<h5> <span class=\"header-section-number\">",h_subsection_number,"</span>Collection", gc, "</h5>\n\n") #Caue results='asis',we must use raw HTML
        subsection_n <- subsection_n + 1
        
        x %<>% 
          mutate(log10P = -log10(p.adjust)) %>% 
          arrange(p.adjust) %>% 
          head(30)
        
        res <- gse.dotplot(
          x,
          x = 'NES',
          color = 'NES',
          size = "log10P",
          # color2log10 = TRUE,
          showCategory = 30,
          label_format = 60
        ) +
          geom_vline(xintercept = 0, lty = 3) +
          scale_size_continuous(range=c(0.3, 2.5)) +
          scale_color_gradient2(trans = "reverse") +
          scale_x_continuous(expand = expansion(mult = c(0.05, 0.05))) +
          labs(
            size = expression(paste("lo", g[10],"(Adjusted P value)")),
            caption = 'Showing only top-30 gene sets\nNES = Normalized Enrichment Score'
            # size = 'Gene set\nsize'
          ) +
          theme(
            panel.border = element_rect(color = "black", fill = NA, size = use_linewidth/2),
            
            axis.line = element_blank(),
            axis.ticks = element_line(color = "black", linewidth = use_linewidth/2),
            axis.text.y = element_text(size = 4),
            axis.text.x = element_text(size = 4),
            
            legend.text = element_text(size = 4), 
            legend.title = element_text(size = 4),
            legend.key.width = unit(0.2, "cm")
          )
        
        subchunkify(res, 
                    fig.width = 3.5, 
                    fig.height = (3.5*0.3) + nrow(x)*0.05,
                    dpi = 600,
                    prefix = use_prefix,
                    suffix = paste0(tolower(i) %>% gsub(" ", "-", .), '_', sr, '_', gc) )
        cat("\n\n")
      }
    }
  }
}


knitr::opts_knit$set(
  base.dir = NULL
  )
cat("\n\n")
```


### Gene-set similarity matrix by gene set collection
```{r gse-jcmatrix-comparisons-conf, echo=FALSE}

gse_collections <- list(
  GSEA = names(gmt_files_symbols)
)

#
# Generate a list of JC similarity matrices by gene-set
#
library(enrichplot)
p_thrs <- 0.05
topn <- 50
use_comp <- names(gse_list)[1]
jc_matrix_list <- foreach(use_comp = names(gse_list)) %do% {
  data_gse <- foreach(sr = names(gse_collections)) %do% {
    data_by_collection <- foreach(gc = gse_collections[[sr]])  %do% {
      
      if(sr == 'GSEA') {
        use_enrich_results <- gse_list[[use_comp]][[sr]][[gc]]
        top_descriptions <- use_enrich_results@result %>% 
          filter(p.adjust <= p_thrs) %>% 
          head(topn) %>% 
          pull(Description)
        use_n_categories <-  use_enrich_results@result %>% 
          filter(p.adjust <= p_thrs) %>% 
          nrow
      }
      if(sr == 'gseGO') {
        use_enrich_results <- gse_list[[use_comp]][[sr]][[gc]]
        top_descriptions <- use_enrich_results@result %>% 
          filter(p.adjust <= p_thrs & simplify) %>% 
          head(topn) %>% 
          pull(Description)
        use_n_categories <-  use_enrich_results@result %>% 
          filter(p.adjust <= p_thrs) %>% 
          nrow
      }
      if(sr %in% c('gseKEGG', 'gseReactome')) {
        use_enrich_results <- gse_list[[use_comp]][[sr]]
        top_descriptions <- use_enrich_results@result %>% 
          filter(p.adjust <= p_thrs) %>% 
          head(topn) %>% 
          pull(Description)
        use_n_categories <-  use_enrich_results@result %>% 
          filter(p.adjust <= p_thrs) %>% 
          nrow
      }
      
      if(length(top_descriptions) > 1) {
        use_enrich_results <- pairwise_termsim(
          use_enrich_results, 
          showCategory = use_n_categories, 
          method = "JC")
        use_enrich_results@termsim[top_descriptions, top_descriptions]
      } else {
        NULL
      }
    }
    names(data_by_collection) <- gse_collections[[sr]]
    return(compact(data_by_collection))
  }
  names(data_gse) <- names(gse_collections)
  return(data_gse)
}
names(jc_matrix_list) <- names(gse_list)
```

Gene-set similarity matrix by comparison and gene set collections. Heatmap showing the pair-wise similarity matrix of top `r topn` gene sets with an adjusted P value < `r p_thrs` in each gene group. Heatmap colors represent the Jaccard similarity coefficient.


<!-- # Subchunkify dotplots -->
<!-- ## chunk options need to be set to echo=FALSE, results='asis' -->
<!-- ## Cause results='asis',we must use raw HTML to print text and headings -->
```{r gse-jcmatrix-comparisons, echo=FALSE, results='asis'}
knitr::opts_knit$set(
  base.dir = "docs"
  )

library(NbClust)

gse_collections <- list(
  GSEA = names(gmt_files_symbols)
  # gseGO = c("BP", "MF"),
  # gseReactome = 'gseReactome'
)

use_prefix <- 'gse-jcmatrix-comparisons'
use_linewidth <- 1/2.141959
root_section_number <- 8.3
section_n <- 1
subsection_n <- 1

col_fun <- colorRamp2(seq(0, 1, length.out = 4), brewer.pal(4, "GnBu"))


i <- names(jc_matrix_list)[1]
for(i in names(jc_matrix_list)) {
  h_section_number <- paste(root_section_number, section_n, sep=".")
  
  cat("<h4> <span class=\"header-section-number\">",h_section_number,"</span>Comparison", i, "</h4>\n\n") #Caue results='asis',we must use raw HTML
  
  section_n <- section_n + 1
  
  for(sr in names(jc_matrix_list[[i]])) {
    
    for(gc in names(jc_matrix_list[[i]][[sr]])) {
      use_mat <- jc_matrix_list[[i]][[sr]][[gc]]
      if(nrow(use_mat) >= 10) {
        h_subsection_number <- paste(h_section_number, subsection_n, sep=".")
        cat("<h5> <span class=\"header-section-number\">",h_subsection_number,"</span>Collection", gc, "</h5>\n\n") #Caue results='asis',we must use raw HTML
        subsection_n <- subsection_n + 1
        colnames(use_mat) <- colnames(use_mat) %>% clean_msigdb_names %>% tolower() %>% str_to_title %>% ep_str_wrap(70)
        rownames(use_mat) <- rownames(use_mat) %>% clean_msigdb_names %>% tolower() %>% str_to_title %>% ep_str_wrap(70)
        
        # Define number of clusters
        use_max.nc <- ifelse(nrow(use_mat) <= 15, 5, round(nrow(use_mat) * 0.3))
        nbclust_res <- NbClust(data = NULL, diss = as.dist(1-use_mat), distance = NULL, 
                               min.nc = 2, max.nc = use_max.nc, method = "ward.D",
                               index = 'silhouette')
        
        
        n_split <- nbclust_res$Best.nc['Number_clusters']
        
        ht <- Heatmap(
          use_mat,
          name = 'Jaccard index',
          column_split = n_split,
          row_split = n_split,
          column_title = NULL,
          row_title = NULL,
          row_names_gp = gpar(fontsize = 4),
          col = col_fun,
          show_column_dend = FALSE,
          show_column_names = FALSE,
          border = TRUE,
          # top_annotation = ha_top,
          heatmap_legend_param = list(title_gp = gpar(fontsize = 8),
                                      labels_gp = gpar(fontsize = 8)
          ),
          width = unit(7, "cm"),
          
          clustering_method_rows = "ward.D",
          clustering_method_columns = "ward.D"
          # clustering_distance_rows = function(x) as.dist(1 - x),
          # clustering_distance_columns = function(x) as.dist(1 - x)
        )
        
        subchunkify(draw(ht), 
                    fig.width = 7.2,
                    fig.height = (7.2*0.3) + nrow(use_mat)*0.15,
                    dpi = 600,
                    prefix = use_prefix,
                    suffix = paste0(tolower(i) %>% gsub(" ", "-", .), '_', sr, '_', gc) )
        
        cat("\n\n")
      }
    }
  }
}

knitr::opts_knit$set(
  base.dir = NULL
  )
cat("\n\n")
```


### Dotplots grouped (complete)
<!-- # Subchunkify dotplots -->
<!-- ## chunk options need to be set to echo=FALSE, results='asis' -->
<!-- ## Cause results='asis',we must use raw HTML to print text and headings -->
```{r gse-dotplot-grouped-comparisons, echo=FALSE, results='asis'}
knitr::opts_knit$set(
  base.dir = "docs"
  )

gse_collections <- list(
  GSEA = "msigdb.c2.cp"
)

use_prefix <- 'gse-dotplot-grouped'
use_linewidth <- 1/2.141959
root_section_number <- 8.4
section_n <- 1
subsection_n <- 1

use_comps <- intersect(names(gse_list), names(gsets_groups))
i <- use_comps[1]
sr <- names(gse_collections)[1]
gc <- gse_collections[[sr]][1]

for(i in use_comps) {
  h_section_number <- paste(root_section_number, section_n, sep=".")
  cat("<h4> <span class=\"header-section-number\">",h_section_number,"</span>Comparison", i, "</h4>\n\n") #Caue results='asis',we must use raw HTML
  section_n <- section_n + 1
  for(sr in names(gse_collections)) {
    for(gc in gse_collections[[sr]]) {
      if(sr == 'GSEA') {
        x <- gse_list[[i]][[sr]][[gc]]@result %>% 
          filter(p.adjust <= 0.05) %>% 
          mutate(Description = clean_msigdb_names(Description))
      }
      if(sr == 'gseGO') {
        x <- gse_list[[i]][[sr]][[gc]]@result %>% 
          filter(p.adjust <= 0.05 & simplify)
      }
      if(sr %in% c('gseKEGG', 'gseReactome')) {
        x <- gse_list[[i]][[sr]]@result %>% 
          filter(p.adjust <= 0.05)
      }
      if(nrow(x) > 0) {
        h_subsection_number <- paste(h_section_number, subsection_n, sep=".")
        cat("<h5> <span class=\"header-section-number\">",h_subsection_number,"</span>Collection", gc, "</h5>\n\n") #Caue results='asis',we must use raw HTML
        subsection_n <- subsection_n + 1
         
        x %<>% 
          mutate(
            is.sign = ifelse(p.adjust < 0.05, 'yes', 'no'),
            log10P = -log10(p.adjust)
          ) %>% 
          arrange(p.adjust) %>% 
          left_join(gsets_groups[[i]]) %>% 
          filter(!is.na(Group))
        
        # Duplicated descriptions
        dup_descriptions <- x$Description[x$Description %>% duplicated]
        if(length(dup_descriptions) > 0) {
          x %<>% 
            mutate(
              gset_db = gsub("_.*", "", ID),
              Description = ifelse(
                Description %in% dup_descriptions, 
                paste0(Description," (", gset_db,")"), 
                Description)
            )
        }
        
        group_levels <- x %>% 
          group_by(Group) %>%
          summarise(median_NES = median(NES)) %>% 
          arrange(desc(median_NES)) %>% 
          pull(Group)
        
        res <- x %>% 
          mutate(
            Group = factor(
              default_labeller(20)(Group), 
              levels = default_labeller(20)(group_levels)),
            Description = fct_reorder(Description, NES)
          ) %>%
          ggplot(aes(x = NES, y = Description, fill = NES, size = log10P, stroke = linewidth_default/2)) +
          geom_point(pch = 21, aes(color = is.sign)) +
          geom_vline(xintercept = 0, lty = 3) +
          scale_size_continuous(range=c(0.8, 2)) +
          scale_fill_gradient2(trans = "reverse") +
          scale_color_manual(values = c(no = 'white', yes = 'black')) +
          scale_y_discrete(labels = default_labeller(80))+
          scale_x_continuous(expand = expansion(mult = c(0.20, 0.20))) +
          facet_grid(rows = vars(Group), scales = 'free_y', space = 'free_y') +
          labs(
            y = '',
            size = expression(paste("lo", g[10],"(Adjusted P value)")),
            color = 'FDR < 0.05',
            caption = paste0('Showing only top-', nrow(x),' gene sets\nNES = Normalized Enrichment Score')
          ) +
          theme(
            panel.border = element_rect(color = "black", fill = NA, size = linewidth_default),
            strip.text.y = element_text(angle = 0, hjust = 0, size = 4),
            
            axis.line = element_blank(),
            axis.ticks = element_line(color = "black", linewidth = linewidth_default/2),
            axis.text.y = element_text(size = 4),
            axis.text.x = element_text(size = 4),
            
            legend.text = element_text(size = 4), 
            legend.title = element_text(size = 4),
            legend.key.width = unit(0.2, "cm"),
            
            panel.spacing=unit(0.1, "lines")
          )
          
       n_gsets <- x$ID %>% unique() %>% length()
       # use_width <-  7.2 + n_gsets * 0.02
       use_width <- 4.2
       # use_asp <-  0.318 + n_gsets * 0.017
       # use_asp <-  n_gsets * 0.05
       use_asp <-  n_gsets * 0.021
       use_height <- use_width * use_asp
        
       subchunkify(res,
                   fig.width = use_width,
                   fig.height = use_height,
                   dpi = 600,
                   prefix = use_prefix,
                   suffix = paste0(tolower(i) %>% gsub(" ", "-", .), '_', sr, '_', gc) )

        cat("\n\n")
      }
    }
  }
}


knitr::opts_knit$set(
  base.dir = NULL
  )
cat("\n\n")
```


### Dotplots grouped (unnamed)
<!-- # Subchunkify dotplots -->
<!-- ## chunk options need to be set to echo=FALSE, results='asis' -->
<!-- ## Cause results='asis',we must use raw HTML to print text and headings -->
```{r gse-dotplot-grouped-comparisons-unnamed, echo=FALSE, results='asis'}
knitr::opts_knit$set(
  base.dir = "docs"
  )

gse_collections <- list(
  GSEA = "msigdb.c2.cp"
)

use_prefix <- 'gse-dotplot-grouped-unnamed'
root_section_number <- 8.5
section_n <- 1
subsection_n <- 1

use_comps <- intersect(names(gse_list), names(gsets_groups))
i <- use_comps[1]
sr <- names(gse_collections)[1]
gc <- gse_collections[[sr]][1]

for(i in use_comps) {
  h_section_number <- paste(root_section_number, section_n, sep=".")
  cat("<h4> <span class=\"header-section-number\">",h_section_number,"</span>Comparison", i, "</h4>\n\n") #Caue results='asis',we must use raw HTML
  section_n <- section_n + 1
  for(sr in names(gse_collections)) {
    for(gc in gse_collections[[sr]]) {
      if(sr == 'GSEA') {
        x <- gse_list[[i]][[sr]][[gc]]@result %>% 
          filter(p.adjust <= 0.05) %>% 
          mutate(Description = clean_msigdb_names(Description))
      }
      if(sr == 'gseGO') {
        x <- gse_list[[i]][[sr]][[gc]]@result %>% 
          filter(p.adjust <= 0.05 & simplify)
      }
      if(sr %in% c('gseKEGG', 'gseReactome')) {
        x <- gse_list[[i]][[sr]]@result %>% 
          filter(p.adjust <= 0.05)
      }
      if(nrow(x) > 0) {
        h_subsection_number <- paste(h_section_number, subsection_n, sep=".")
        cat("<h5> <span class=\"header-section-number\">",h_subsection_number,"</span>Collection", gc, "</h5>\n\n") #Caue results='asis',we must use raw HTML
        subsection_n <- subsection_n + 1

        x %<>% 
          mutate(
            is.sign = ifelse(p.adjust < 0.05, 'yes', 'no'),
            log10P = -log10(p.adjust)
          ) %>% 
          arrange(p.adjust) %>% 
          left_join(gsets_groups[[i]]) %>% 
          filter(!is.na(Group)) %>% 
          group_by(Description) %>% 
          # For entries with duplicated Description, keep only one based on min P-value
          # slice_min(pvalue, n =1 , with_ties = FALSE) %>% 
          ungroup()
                
        # Duplicated descriptions
        dup_descriptions <- x$Description[x$Description %>% duplicated]
        if(length(dup_descriptions) > 0) {
          x %<>% 
            mutate(
              gset_db = gsub("_.*", "", ID),
              Description = ifelse(
                Description %in% dup_descriptions, 
                paste0(Description," (", gset_db,")"), 
                Description)
            )
        }
        
        group_levels <- x %>% 
          group_by(Group) %>%
          summarise(median_NES = median(NES)) %>% 
          arrange(desc(median_NES)) %>% 
          pull(Group)
        
        res <- x %>% 
          mutate(
            Group = factor(
              default_labeller(20)(Group), 
              levels = default_labeller(20)(group_levels)),
            Description = fct_reorder(Description, NES)
          ) %>%
          ggplot(aes(x = NES, y = Description, fill = NES, size = log10P)) +
          geom_point(pch = 21, aes(color = is.sign), stroke = linewidth_default/2) +
          geom_vline(xintercept = 0, lty = 3) +
          scale_size_continuous(range=c(0.5, 1.1)) +
          scale_fill_gradient2(trans = "reverse") +
          scale_color_manual(values = c(no = 'white', yes = 'black')) +
          scale_y_discrete(labels = default_labeller(80))+
          scale_x_continuous(expand = expansion(mult = c(0.20, 0.20))) +
          facet_grid(rows = vars(Group), scales = 'free_y', space = 'free_y') +
          labs(
            x = '',
            y = '',
            size = expression(paste("lo", g[10],"(FDR)")),
            color = 'FDR\n< 0.05'
          ) +
          theme(
            panel.border = element_rect(color = "black", fill = NA, size = linewidth_default),
            strip.text.y = element_blank(),
            strip.text.x = element_blank(),
            
            axis.line = element_blank(),
            axis.ticks = element_line(color = "black", linewidth = linewidth_default/2),
            axis.text.y = element_blank(),
            axis.text.x = element_blank(),
            
            legend.text = element_text(size = 3), 
            legend.title = element_text(size = 3),
            legend.key.width = unit(0.1, "cm"),
            legend.key.height = unit(0.1, "cm"),
            
            panel.spacing=unit(0.1, "lines"),
            plot.margin = unit(c(0.0217, 0.0217, 0.0217, 0.0217), "inches")
          )
        
        n_gsets <- x$ID %>% unique() %>% length()
        n_gsgroups <- 1
        
        
        
        # Calculate width = left margin + total width of panels + total width of spacing + right margin. Everything scaled by conversion factor to get final width in inches. Conversion factor calculater with a rule in the screen
        # nmodels <- x$model %>% unique() %>% length()
        nmodels <- 2
        
        # Calculate width: left margin + tickmarks + total width of panels + total width of spacing + right margin
        # use_width <- 0.1165 + 0.0173 + (0.18898*nmodels) + (0.035*(nmodels - 1))+ 0.0217
        use_width <- 0.1165 + 0.0173 + (0.18898*nmodels) + (0.035*(nmodels - 1))+ 0.0217
        
        # Calculate height: top margin + tickmarks + bottom margin and legend
        use_height <- 0.0217 + 0.0217 + n_gsets * 0.0492 + (0.035*(nmodels - 1)) + 0.4984
        subchunkify(res +
                      guides(
                        size  = guide_legend(direction = "vertical",
                                             keyheight = unit(0.0289, "inches"),
                                             keywidth = unit(0.0289, "inches")
                        ),
                        color = "none",
                        fill  = "none"
                        # color = guide_legend(direction = "vertical"),
                        # fill  = guide_colourbar(direction = "vertical")
                      ) +
                      theme(
                        legend.position = "bottom",
                        legend.box = "horizontal"
                      ),
                    fig.width = use_width,
                    fig.height = use_height,
                    dpi = 600,
                    prefix = use_prefix,
                    suffix = paste0(tolower(i) %>% gsub(" ", "-", .), '_', sr, '_', gc) )
        
        cat("\n\n")
        
        # Print legend
        res_legend <- get_legend(
          res +
            guides(
              color = guide_legend(direction = "vertical"),
              size  = guide_legend(direction = "vertical"),
              fill  = guide_colourbar(direction = "vertical")
            ) +
            theme(
              legend.position = "bottom",
              legend.box = "horizontal"
            )
        )
        subchunkify(plot_grid(res_legend),
                    fig.width = 1.8,
                    fig.height = 1,
                    dpi = 600,
                    prefix = paste0(use_prefix,'-legend-unnamed'),
                    suffix = paste0(tolower(i) %>% gsub(" ", "-", .), '_', sr, '_', gc) )
        
        cat("\n\n")
      }
    }
  }
}


knitr::opts_knit$set(
  base.dir = NULL
  )
cat("\n\n")
```

### Dotplots simplified (complete)

Here we show a single point for each group. 
<!-- # Subchunkify dotplots -->
<!-- ## chunk options need to be set to echo=FALSE, results='asis' -->
<!-- ## Cause results='asis',we must use raw HTML to print text and headings -->
```{r gse-dotplot-grouped-simplified, echo=FALSE, results='asis'}
knitr::opts_knit$set(
  base.dir = "docs"
  )

gse_collections <- list(
  GSEA = "msigdb.c2.cp"
)

use_prefix <- 'gse-dotplot-grouped-simplified'
root_section_number <- 8.6
section_n <- 1
subsection_n <- 1

use_comps <- intersect(names(gse_list), names(gsets_groups))
i <- use_comps[1]
sr <- names(gse_collections)[1]
gc <- gse_collections[[sr]][1]

for(i in use_comps) {
  h_section_number <- paste(root_section_number, section_n, sep=".")
  cat("<h4> <span class=\"header-section-number\">",h_section_number,"</span>Comparison", i, "</h4>\n\n") #Caue results='asis',we must use raw HTML
  section_n <- section_n + 1
  for(sr in names(gse_collections)) {
    for(gc in gse_collections[[sr]]) {
      if(sr == 'GSEA') {
        x <- gse_list[[i]][[sr]][[gc]]@result %>% 
          filter(p.adjust <= 0.05) %>% 
          mutate(Description = clean_msigdb_names(Description))
      }
      if(sr == 'gseGO') {
        x <- gse_list[[i]][[sr]][[gc]]@result %>% 
          filter(p.adjust <= 0.05 & simplify)
      }
      if(sr %in% c('gseKEGG', 'gseReactome')) {
        x <- gse_list[[i]][[sr]]@result %>% 
          filter(p.adjust <= 0.05)
      }
      if(nrow(x) > 0) {
        h_subsection_number <- paste(h_section_number, subsection_n, sep=".")
        cat("<h5> <span class=\"header-section-number\">",h_subsection_number,"</span>Collection", gc, "</h5>\n\n") #Caue results='asis',we must use raw HTML
        subsection_n <- subsection_n + 1
         
        
        x %<>% 
          left_join(gsets_groups[[i]]) %>% 
          filter(!is.na(Group))
        
        x_median <- x %>% 
          group_by(Group) %>% 
          summarise(
            group_n = n(),
            group_nes = median(NES)
          ) %>% 
          mutate(
            group_sign = ifelse(group_nes > 0, 'activated', 'supressed')
          ) %>% 
          arrange(group_nes)
        
        x_simplified <- x %>% 
          left_join(x_median) %>% 
          filter(.sign == group_sign) %>% 
          group_by(Group) %>%
          slice_min(p.adjust, n = 1, with_ties = FALSE) %>% 
          mutate(
            log10P = -log10(p.adjust)
          ) %>% 
          ungroup
        

        res <- x_simplified %>% 
          mutate(
            # Group = factor(
            #   Group, 
            #   levels = x_median$Group)
            Group = fct_reorder(Group, NES)
          ) %>%
          ggplot(aes(x = NES, y = Group, fill = NES, size = log10P)) +
          geom_point(pch = 21, color = 'black') +
          geom_vline(xintercept = 0, lty = 3) +
          scale_size_continuous(range=c(0.8, 2)) +
          scale_fill_gradient2(trans = "reverse") +
          scale_y_discrete(labels = default_labeller(80))+
          scale_x_continuous(expand = expansion(mult = c(0.20, 0.20))) +
          labs(
            y = '',
            size = expression(paste("lo", g[10],"(Adjusted P value)"))
            # caption = 'NES = Normalized Enrichment Score'
          ) +
          theme(
            panel.border = element_rect(color = "black", fill = NA, size = linewidth_default),
            strip.text.y = element_text(angle = 0, hjust = 0, size = 4),
            
            axis.line = element_blank(),
            axis.ticks = element_line(color = "black", linewidth = linewidth_default),
            axis.text.y = element_text(size = 4),
            axis.text.x = element_text(size = 4),
            
            legend.text = element_text(size = 4), 
            legend.title = element_text(size = 4),
            legend.key.width = unit(0.2, "cm"),
            
            panel.spacing=unit(0.1, "lines")
          )
        
       #    
       n_gsets <- x_simplified$Group %>% unique() %>% length()
       # use_width <-  7.2 + n_gsets * 0.02
       use_width <- 2.8
       # use_asp <-  0.318 + n_gsets * 0.017
       # use_asp <-  n_gsets * 0.05
       use_asp <-  n_gsets * 0.051
       use_height <- use_width * use_asp

       subchunkify(res,
                   fig.width = use_width,
                   fig.height = use_height,
                   dpi = 600,
                   prefix = use_prefix,
                   suffix = paste0(tolower(i) %>% gsub(" ", "-", .), '_', sr, '_', gc) )

        cat("\n\n")
      }
    }
  }
}


knitr::opts_knit$set(
  base.dir = NULL
  )
cat("\n\n")
```



### Dotplots simplified (unnamed)

Here we show a single point for each group. 
<!-- # Subchunkify dotplots -->
<!-- ## chunk options need to be set to echo=FALSE, results='asis' -->
<!-- ## Cause results='asis',we must use raw HTML to print text and headings -->
```{r gse-dotplot-grouped-simplified-unnamed, echo=FALSE, results='asis'}
knitr::opts_knit$set(
  base.dir = "docs"
  )

gse_collections <- list(
  GSEA = "msigdb.c2.cp"
)

use_prefix <- 'gse-dotplot-grouped-simplified-unnamed'
root_section_number <- 8.7
section_n <- 1
subsection_n <- 1

use_comps <- intersect(names(gse_list), names(gsets_groups))
i <- use_comps[1]
sr <- names(gse_collections)[1]
gc <- gse_collections[[sr]][1]

for(i in use_comps) {
  h_section_number <- paste(root_section_number, section_n, sep=".")
  cat("<h4> <span class=\"header-section-number\">",h_section_number,"</span>Comparison", i, "</h4>\n\n") #Caue results='asis',we must use raw HTML
  section_n <- section_n + 1
  for(sr in names(gse_collections)) {
    for(gc in gse_collections[[sr]]) {
      if(sr == 'GSEA') {
        x <- gse_list[[i]][[sr]][[gc]]@result %>% 
          filter(p.adjust <= 0.05) %>% 
          mutate(Description = clean_msigdb_names(Description))
      }
      if(sr == 'gseGO') {
        x <- gse_list[[i]][[sr]][[gc]]@result %>% 
          filter(p.adjust <= 0.05 & simplify)
      }
      if(sr %in% c('gseKEGG', 'gseReactome')) {
        x <- gse_list[[i]][[sr]]@result %>% 
          filter(p.adjust <= 0.05)
      }
      if(nrow(x) > 0) {
        h_subsection_number <- paste(h_section_number, subsection_n, sep=".")
        cat("<h5> <span class=\"header-section-number\">",h_subsection_number,"</span>Collection", gc, "</h5>\n\n") #Caue results='asis',we must use raw HTML
        subsection_n <- subsection_n + 1
         
        
        x %<>% 
          left_join(gsets_groups[[i]]) %>% 
          filter(!is.na(Group))
        
        x_median <- x %>% 
          group_by(Group) %>% 
          summarise(
            group_n = n(),
            group_nes = median(NES)
          ) %>% 
          mutate(
            group_sign = ifelse(group_nes > 0, 'activated', 'supressed')
          ) %>% 
          arrange(group_nes)
        
        x_simplified <- x %>% 
          left_join(x_median) %>% 
          filter(.sign == group_sign) %>% 
          group_by(Group) %>%
          slice_min(p.adjust, n = 1, with_ties = FALSE) %>% 
          mutate(
            log10P = -log10(p.adjust)
          ) %>% 
          ungroup
        
        
        
        res <- x_simplified %>% 
          mutate(
            # Group = factor(
            #   Group, 
            #   levels = x_median$Group)
            Group = fct_reorder(Group, NES)
          ) %>%
          ggplot(aes(x = NES, y = Group, fill = NES, size = log10P)) +
          geom_point(pch = 21, color = 'black') +
          geom_vline(xintercept = 0, lty = 3) +
          scale_size_continuous(range=c(0.8, 2)) +
          scale_fill_gradient2(trans = "reverse") +
          scale_x_continuous(expand = expansion(mult = c(0.20, 0.20))) +
          labs(
            x = '',
            y = '',
            size = expression(paste("lo", g[10],"(Adjusted P value)"))
            # caption = 'NES = Normalized Enrichment Score'
          ) +
          theme(
            panel.border = element_rect(color = "black", fill = NA, size = linewidth_default),
            strip.text.y = element_blank(),
            strip.text.x = element_blank(),
            
            axis.line = element_blank(),
            axis.ticks = element_line(color = "black", linewidth = linewidth_default/2),
            axis.text.y = element_blank(),
            axis.text.x = element_blank(),
            
            legend.text = element_text(size = 3), 
            legend.title = element_text(size = 3),
            legend.key.width = unit(0.1, "cm"),
            legend.key.height = unit(0.1, "cm"),
            
            plot.margin = unit(c(0.0217, 0.0217, 0.0217, 0.0217), "inches")
          )
        
        
        n_gsets <- x$Group %>% unique() %>% length()
        
        # Calculate width = left margin + total width of panels + total width of spacing + right margin. Everything scaled by conversion factor to get final width in inches. Conversion factor calculater with a rule in the screen
        # nmodels <- x$model %>% unique() %>% length()
        nmodels <- 2
        
        # Calculate width: left margin + tickmarks + total width of panels + total width of spacing + right margin
        # use_width <- 0.1165 + 0.0173 + (0.18898*nmodels) + (0.035*(nmodels - 1))+ 0.0217
        use_width <- 0.1165 + 0.0173 + (0.18898*nmodels) + (0.035*(nmodels - 1))+ 0.0217
        
        # Calculate height: top margin + tickmarks + bottom margin and legend
        use_height <- 0.0217 + 0.0217 + (n_gsets*1.75) * 0.0492 + (0.035*(nmodels - 1)) + 0.4984
        subchunkify(res +
                      guides(
                        size  = guide_legend(direction = "vertical",
                                             keyheight = unit(0.0289, "inches"),
                                             keywidth = unit(0.0289, "inches")
                        ),
                        color = "none",
                        fill  = "none"
                        # color = guide_legend(direction = "vertical"),
                        # fill  = guide_colourbar(direction = "vertical")
                      ) +
                      theme(
                        legend.position = "bottom",
                        legend.box = "horizontal"
                      ),
                    fig.width = use_width,
                    fig.height = use_height,
                    dpi = 600,
                    prefix = use_prefix,
                    suffix = paste0(tolower(i) %>% gsub(" ", "-", .), '_', sr, '_', gc) )
        
        cat("\n\n")
        
        # Print legend
        res_legend <- get_legend(
          res +
            guides(
              color = guide_legend(direction = "vertical"),
              size  = guide_legend(direction = "vertical"),
              fill  = guide_colourbar(direction = "vertical")
            ) +
            theme(
              legend.position = "bottom",
              legend.box = "horizontal"
            )
        )
        subchunkify(plot_grid(res_legend),
                    fig.width = 1.8,
                    fig.height = 1,
                    dpi = 600,
                    prefix = paste0(use_prefix,'-legend-unnamed'),
                    suffix = paste0(tolower(i) %>% gsub(" ", "-", .), '_', sr, '_', gc) )
        
        cat("\n\n")
      }
    }
  }
}


knitr::opts_knit$set(
  base.dir = NULL
  )
cat("\n\n")
```




## Spatial distribution
```{r bc-caf-spatial}
# Add distance to CAF spots
seu[[]] %<>% 
  rownames_to_column('barcode') %>% 
  left_join(caf_hotspot_r_dist, by = 'barcode') %>%
  column_to_rownames('barcode') %>% 
  mutate(
    caf_dist_cat = 'intermediate',
    caf_dist_cat = ifelse(
      r_dist_caf_hotspot_3d_transformed < 100, 'l100um',
      caf_dist_cat
    ),
    caf_dist_cat = ifelse(
      r_dist_caf_hotspot_3d_transformed > 1000, 'u1000um',
      caf_dist_cat
    ),
    caf_dist_cat_4 = ifelse(r_dist_caf_hotspot_3d_transformed <= 0,
                            'CAF',
                            caf_dist_cat),
    caf_dist_cat = factor(caf_dist_cat),
    caf_dist_cat_4 = factor(caf_dist_cat_4)
  )

use_palette <- c(CAF = '#E69F00',
                intermediate =  'grey80',
                 l100um =  '#B5DFFD', 
                 u1000um = '#A40905')

i <- names(seu@images)[3]
for(i in names(seu@images)) {
  cat('###' , i, '\n')
  plot_list <- list()
  if(!is.null(seu@images[[i]])) {
    plot_list[['caf_dist_cat']] <-
      SpatialDimPlot(
        seu,
        group.by = 'caf_dist_cat',
        image.alpha = 0.5,
        pt.size.factor = 3,
        images = i
      ) +
      scale_fill_manual(values = use_palette) +
      labs(
        fill = "",
        title = 'Distance from CAF hotspots'
      )
    plot_list[['caf_dist_cat_4']] <-
      SpatialDimPlot(
        seu,
        group.by = 'caf_dist_cat_4',
        image.alpha = 0.5,
        pt.size.factor = 3,
        images = i
      ) +
      scale_fill_manual(values = use_palette) +
      labs(
        fill = "",
        title = 'Distance from CAF hotspots'
      )
  } else {
    plot_list[['caf_dist_cat']] <- plot_spacer()
    plot_list[['caf_dist_cat_4']] <- plot_spacer()
  }
  print(wrap_plots(plot_list))
  cat("\n\n")
}
```

